---
title: "Metabolomics + dietary metadata"
format: html
editor: visual
---

This code serves for exploring pattern in metabolomics of vegans and diet dietary metadata.

```{r}
library(ggplot2)
library(cluster)    # For silhouette
library(factoextra) # For easy plotting (optional)
library(clusterCrit) # For internal clustering metrics (like CH index)
library(tidyverse)
library(mclust)
library(umap)
library(ggpubr)
```

```{r}
umap_plot <- function(df, metadata,column="Cluster",
                      neighbors=10,
                      min_dist=0.1){
  
  custom.config <- umap.defaults
  custom.config$random_state <- 123
  custom.config$n_neighbors <- neighbors
  custom.config$min_dist <- min_dist
  
  umap_plot <- umap(df,config=custom.config)
  data_umap <- data.frame(umap_plot$layout)
  data_umap <- merge(data_umap %>% rownames_to_column("SampleID"), 
                     metadata,by="SampleID")
  
    p <- ggplot(data=data_umap, aes(x=X1,y=X2,color=!!sym(column))) + 
      geom_point() + 
      theme_bw() +
      xlab("UMAP 1") + ylab("UMAP 2") 
  
  return(p)
}
```

# Import

```{r}
metadata_df <- read.table("../../Data/metadata_vegans.tsv",header = TRUE)
lcms_df <- read.table("../../Data/preprocessed/metabolomics/lc_ms_preprocessed.tsv")
nmr_df <- read.table("../../Data/preprocessed/metabolomics/nmr_preprocessed.tsv")

mofa_integrated_df <- read.table("../../Data/preprocessed/metabolomics/mofa_integrated.tsv")

spca_integrated_df <-  read.table("../../Data/preprocessed/metabolomics/spca_integrated.tsv")

dietary_df <- read.table("../../Data/preprocessed/dietary/fg_merged_mean_preprocessed.tsv",header=TRUE)

dietary_df <- dietary_df %>%
  dplyr::select(-amount,-Energy.kJ.)
```

**Get only important samples**

```{r}
# dietary DF
dietary_df <- dietary_df %>%
  dplyr::filter(rownames(dietary_df) %in% metadata_df$SampleID)

# Metadata DF
metadata_df <- metadata_df %>%
  dplyr::filter(metadata_df$SampleID %in% rownames(dietary_df))

# metabolomics
## LCMS
lcms_df <- lcms_df %>%
  dplyr::filter(rownames(lcms_df) %in% metadata_df$SampleID)

## NMR
nmr_df <- nmr_df %>%
  dplyr::filter(rownames(nmr_df) %in% metadata_df$SampleID)

## MOFA
mofa_integrated_df <- mofa_integrated_df %>%
  dplyr::filter(rownames(mofa_integrated_df) %in% metadata_df$SampleID)

## SPCA
spca_integrated_df <- spca_integrated_df %>%
  dplyr::filter(rownames(spca_integrated_df) %in% metadata_df$SampleID)
```

# Dietary dataset

## PCA

```{r}
data.pca <- prcomp(dietary_df)
plot(data.pca)
```

```{r,fig.height=10,fig.width=10}
library(factoextra)

p <- fviz_pca_var(data.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

p
```

```{r}
pdf("../Results/dietary_PCA_biplot.pdf",height = 20,width = 20)
p
dev.off()
```

```{r,fig.height=10,fig.width=10}

p <- ggplot(data.pca$x,aes(x = PC1,y=PC2)) + 
  geom_point()

p
```

```{r}
pdf("../Results/dietary_PCA_samples.pdf",height = 20,width = 20)
p
dev.off()
```

## Correlation analysis

```{r,fig.height=20, fig.width=20}
library(ggcorrplot)
corr <- round(cor(dietary_df), 1)
p <- ggcorrplot(corr, method = "circle")

p
```

## Clustering

### Dendrogram

**Whole dataset**

```{r,fig.width=20,fig.height=10}
dist_mat <- dist(dietary_df, method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)

clusters_dendrogram_A <- cutree(hclust_avg, k = 10)
```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(factor(clusters_dendrogram_A,levels = c(1,2,3,4,5,6,7,8,9,10))) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  scale_color_manual(values=c("red","blue","green","grey","black",
                              "darkred","darkblue","darkgreen","darkgrey","brown"))

p
```

```{r}
pdf("../Results/dietary_PCA_dendrogram.pdf",height = 10,width = 10)
p
dev.off()
```

**PCA-based dataset**

```{r,fig.width=20,fig.height=10}
dist_mat <- dist(data.pca$x[,1:10], method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)
clusters_dendrogram_B <- cutree(hclust_avg, k = 10)
```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(factor(clusters_dendrogram_B,levels = c(1,2,3,4,5,6,7,8,9,10))) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  scale_color_manual(values=c("red","blue","green","grey","black",
                              "darkred","darkblue","darkgreen","darkgrey","yellow"))

p
```

```{r}
pdf("../Results/dietary_PCA_dendrogramB.pdf",height = 10,width = 10)
p
dev.off()
```

### k-means

Employ k-menas clustering, try different "k" = number of clusters. Analyze this number by two indices: Silhouette Index & Calinski–Harabasz Index

**The Silhouette Index** evaluates how well each data point fits within its cluster compared to other clusters.

For each point i: a(i): average distance to all other points in the same cluster b(i): average distance to all points in the nearest neighboring cluster (the one it is not assigned to)

s(i) = \[b(i) - a(i)\] / \[max{a(i),b(i)}\]

s(i) ranges from -1 to 1: - Near 1: well-matched to its own cluster, poorly matched to others - Near 0: on the boundary between two clusters - Near -1: probably misclassified

The average silhouette score across all points gives an overall measure of clustering quality.

**The Calinski–Harabasz (CH) Index** —also called the Variance Ratio Criterion—measures cluster separation relative to cluster compactness.

CH Index= \[Within cluster variance/(k-1)\] / \[Between cluster variance/(n-k)\]

k = number of clusters; n = number of data points

Higher CH index = better defined and more separated clusters.

It increases when:

-   Cluster centers are far apart

-   Clusters are tight and compac

#### The whole dataset

```{r}
data <- dietary_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans_A <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/dietary_PCA_kmeans.pdf",height = 10,width = 10)
p
dev.off()
```

```{r}
color_variable <- "Country"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

#### PCA-based dataset

```{r}
data <- data.pca$x[,1:10]

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans_B <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)

kmeans_clusters <- metadata_clusters
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/dietary_PCA_kmeansB.pdf",height = 10,width = 10)
p
dev.off()
```

```{r}
color_variable <- "Country"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

### GMM

**Gaussian Mixture Models (GMMs)** are a probabilistic clustering method that assumes your data is generated from a mixture of several Gaussian distributions (i.e. bell curves), each representing a different cluster.

Unlike k-means, which assigns each point to exactly one cluster, GMM assigns probabilities of belonging to each cluster — this is called soft clustering.

**The Bayesian Information Criterion (BIC)** is a model selection metric that helps you choose the best model among a set of candidates by balancing:

Goodness of fit (how well the model explains the data)

Model complexity (number of parameters) -\> the higher the better

#### The whole dataset

```{r}
data <- dietary_df
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)

# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 2

```{r}
data <- dietary_df
gmm_model <- Mclust(data,G = 2,modelNames = "EEI")

# Print summary
summary(gmm_model)

clusters_gmm_A <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_A) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")

p
```

```{r}
pdf("../Results/dietary_PCA_gmm.pdf",height = 10,width = 10)
p
dev.off()
```

#### PCA-based dataset

```{r}
data <- data.pca$x[,1:10]
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)


# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 3. However as was seen, one cluster has only couple of samples included, so lets stay at 2.

```{r}
data <- data.pca$x[,1:10]
gmm_model <- Mclust(data,G = 2,modelNames = "VEI")

# Print summary
summary(gmm_model)

clusters_gmm_B <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_B) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)

gmm_clusters <- metadata_clusters
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")

p
```

```{r}
pdf("../Results/dietary_PCA_gmmB.pdf",height = 10,width = 10)
p
dev.off()
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(as.data.frame(data.pca$x),metadata_clusters,
               neighbors = 10,
               min_dist = 0.05)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(as.data.frame(data.pca$x),metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/dietary_UMAP_GMM.pdf",height = 4,width = 4)
p
dev.off()
```

### Dietary clusters export

```{r}
dietary_clusters <- merge(kmeans_clusters,gmm_clusters,by=c("SampleID","Country","Group"),suffixes = c("_kmeans","_gmm"))
```

```{r}
write.table(dietary_clusters,"../Results/dietary_clusters.tsv",
            row.names = FALSE,sep="\t")
```

# Metabolomics dataset

## PCA

### MOFA

```{r,fig.height=4,fig.width=10}
mofa_plot_df <- merge(mofa_integrated_df %>% rownames_to_column("SampleID") ,
                      dietary_clusters,by="SampleID")

p <- ggplot(mofa_plot_df) + 
  geom_point(aes(x=Factor1,y=Factor2,color=Cluster_kmeans,)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

g <- ggplot(mofa_plot_df) + 
  geom_point(aes(x=Factor1,y=Factor2,color=Cluster_gmm)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

a <- ggarrange(p,g)

a
```

```{r}
pdf("../Results/MOFA_PCA_samples.pdf",height = 4,width = 4)
a
dev.off()
```

### SPCA

```{r,fig.width=10,fig.height=4}
spca_plot_df <- merge(spca_integrated_df %>% rownames_to_column("SampleID"),
                      dietary_clusters,by="SampleID")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_kmeans)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

g <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_gmm)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

a <- ggarrange(p,g)

a
```

```{r}
pdf("../Results/spca_PCA_samples.pdf",height = 4,width = 8)
a
dev.off()
```

### LCMS

```{r}
data.pca <- prcomp(lcms_df)
```

```{r,fig.height=10,fig.width=10}
library(factoextra)

p <- fviz_pca_var(data.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

p
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>% rownames_to_column("SampleID") ,dietary_clusters,by="SampleID")

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r,fig.width=10}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_kmeans)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

g <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_gmm)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

a <- ggarrange(p,g)

a
```

```{r}
pdf("../Results/lcms_PCA_samples.pdf",height = 4,width = 10)
a
dev.off()
```

### NMR

```{r}
data.pca <- prcomp(nmr_df)
```

```{r,fig.height=10,fig.width=10}
library(factoextra)

p <- fviz_pca_var(data.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

p
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>% rownames_to_column("SampleID") ,dietary_clusters,by="SampleID")

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_kmeans)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

g <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster_gmm)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

a <- ggarrange(p,g)

a
```

```{r}
pdf("../Results/NMR_PCA_samples.pdf",height = 4,width = 10)
a
dev.off()
```

## UMAP

### MOFA

```{r,fig.height=4,fig.width=10}
p <- umap_plot(mofa_integrated_df,dietary_clusters,
               column = "Cluster_kmeans",
               neighbors = 50,
               min_dist = 0.5)


g <- umap_plot(mofa_integrated_df,dietary_clusters,
               column = "Cluster_gmm",
               neighbors = 50,
               min_dist = 0.5)

a <- ggarrange(p,g)
a
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(mofa_integrated_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### SPCA

```{r,fig.height=4,fig.width=10}
p <- umap_plot(spca_integrated_df,dietary_clusters,
               column = "Cluster_kmeans",
               neighbors = 10,
               min_dist = 0.1)


g <- umap_plot(spca_integrated_df,dietary_clusters,
               column = "Cluster_gmm",
               neighbors = 10,
               min_dist = 0.1)

a <- ggarrange(p,g)
a
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(spca_integrated_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### LCMS

```{r,fig.height=4,fig.width=10}
p <- umap_plot(lcms_df,dietary_clusters,
               column = "Cluster_kmeans",
               neighbors = 50,
               min_dist = 0.5)


g <- umap_plot(lcms_df,dietary_clusters,
               column = "Cluster_gmm",
               neighbors = 50,
               min_dist = 0.5)

a <- ggarrange(p,g)
a
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(lcms_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### NMR

```{r,fig.height=4,fig.width=10}
p <- umap_plot(nmr_df,dietary_clusters,
               column = "Cluster_kmeans",
               neighbors = 10,
               min_dist = 0.1)


g <- umap_plot(nmr_df,dietary_clusters,
               column = "Cluster_gmm",
               neighbors = 10,
               min_dist = 0.1)

a <- ggarrange(p,g)
a
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(nmr_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

## Clustering

### Dendrogram

**Whole dataset**

```{r,fig.width=20,fig.height=10}
dist_mat <- dist(mofa_integrated_df, method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)

clusters_dendrogram_A <- cutree(hclust_avg, k = 20)
```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(factor(clusters_dendrogram_A,
                                          levels = c(1:20))) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r}

p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = 50,
               min_dist = 0.5)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/metabolomics_UMAP_dendrogram.pdf",height = 4,width = 4)
p
dev.off()
```

### k-means

```{r}
data <- mofa_integrated_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = 50,
               min_dist = 0.3)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/metabolomics_UMAP_kmeans.pdf",height = 4,width = 4)
p
dev.off()
```

### GMM

```{r}
data <- mofa_integrated_df
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)

# Plot clustering result
plot(gmm_model, what = "classification")  # Shows which sample belongs to which Gaussian

# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 2

```{r}
data <- mofa_integrated_df
gmm_model <- Mclust(data,G = 4,modelNames = "VVI")

# Print summary
summary(gmm_model)

clusters_gmm_A <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_A) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = 10,
               min_dist = 0.05)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/metabolomics_UMAP_gmm.pdf",height = 20,width = 20)
p
dev.off()
```

# DIETARY GROUPS

```{r}
dietary_groups_df <- fread("../../Docs/fg_merged_MAINGROUPS_full_mean.tsv") 

dietary_groups_df <- dietary_groups_df %>%
  dplyr::filter(dietary_groups_df$client %in% metadata_df$SampleID)

dietary_groups_df$`MAIN group` <- gsub(" $","",dietary_groups_df$`MAIN group`)
dietary_groups_df$`MAIN group` <- gsub("\t$","",dietary_groups_df$`MAIN group`)

dietary_df_raw <- fread("../../Docs/fg_merged_mean.tsv") %>%
  dplyr::select(client,`Calories(kcal)`) %>%
  `colnames<-`(c("client","all_Calories"))

groups <- unique(dietary_groups_df$`MAIN group`)
pca_plots <- list()
corr_plots <- list()
```

```{r}
for (group in groups){
  
  # subdf
subdf <- dietary_groups_df %>%
  dplyr::filter(`MAIN group`==group) %>%
  dplyr::select(-`MAIN group`)

if (nrow(subdf)!=length(unique(subdf$client))) {
  subdf <- subdf %>%
  group_by(client) %>%
  summarise(across(
    where(is.numeric), 
    mean, na.rm = TRUE
  )) 
}

# info about all calories
subdf <- merge(subdf,dietary_df_raw,by="client",all.x=TRUE)

# preprocessing

## imputation
subdf[subdf==""] <- 0
subdf[is.na(subdf)] <- 0

## normalization
subdf_norm <- subdf %>%
  dplyr::select(-c(client,amount,`Energy(kJ)`,`Calories(kcal)`))
  
subdf_norm <- apply(subdf_norm,2,function(x) x/subdf$`Calories(kcal)`)

if (is.null(nrow(subdf_norm))) next

subdf_norm <- as.data.frame(subdf_norm) %>%
  `rownames<-`(subdf$client) %>%
  rownames_to_column("client")

## log 
subdf_log <- subdf_norm
subdf_log[,-1] <- log(subdf_norm[,-1])

# LETS deal with -Inf -> -10
subdf_log[subdf_log==-Inf] <- -10

## scaling
subdf_scaled <- subdf_log
subdf_scaled[,-1] <- scale(subdf_scaled[,-1], center = TRUE, scale = TRUE)

subdf_scaled <- subdf_scaled %>% column_to_rownames("client")

# remove nans - no variance
subdf_filtered <- subdf_scaled[, colMeans(is.na(subdf_scaled)) < 0.5]
subdf_filtered <- subdf_filtered[rowMeans(is.na(subdf_scaled)) < 0.5,]

if ((ncol(subdf_filtered)<5) | (nrow(subdf_filtered)<5)) next 
data.pca <- prcomp(subdf_filtered)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")

p <- ggplot(as.data.frame(data.pca$x)) + 
  geom_point(aes(x=PC1,y=PC2)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

pca_plots[[length(pca_plots)+1]] <- p + ggtitle(group)

df <- merge(mofa_integrated_df %>% rownames_to_column("SampleID"),
           as.data.frame(data.pca$x) %>% rownames_to_column("SampleID"),
           by="SampleID") %>%
  dplyr::filter(Factor1<10)

a <- ggplot(data=df) + 
  geom_point(aes(x=Factor1,PC1)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

corr_plots[[length(corr_plots)+1]] <- a + ggtitle(group)

}
```

```{r}
a <- ggarrange(plotlist = pca_plots)
b <- ggarrange(plotlist = corr_plots)
```

```{r}
pdf("../Results/pca_perFG.pdf",height = 20,width = 20)
a
dev.off()

```

```{r}
pdf("../Results/corr_perFG.pdf",height = 20,width = 20)
b
dev.off()

```

# SCCA

Sparse Canonical Correlation Analysis (sCCA) is a powerful technique to uncover associations between two multivariate datasets—in your case, between dietary data and metabolomics data. Unlike classical CCA, sCCA performs variable selection, making interpretation easier by identifying only the most relevant features.

What sCCA Does in This Context Input:

X = dietary dataset (e.g., nutrients, amino acids, phytochemicals)

Y = metabolomics dataset (e.g., LC-MS or NMR metabolites)

Output:

Canonical variates: linear combinations of variables in X and Y

Correlated latent structures: highlights which features from each set are most associated

Why sparse?

With many features and few samples, sparsity avoids overfitting and improves interpretability by keeping only a subset of variables in each canonical component.

```{r}
library(sRDA)
```

```{r}
for (group in groups){
  
  # subdf
subdf <- dietary_groups_df %>%
  dplyr::filter(`MAIN group`==group) %>%
  dplyr::select(-`MAIN group`)

if (nrow(subdf)!=length(unique(subdf$client))) {
  subdf <- subdf %>%
  group_by(client) %>%
  summarise(across(
    where(is.numeric), 
    mean, na.rm = TRUE
  )) 
}

# info about all calories
subdf <- merge(subdf,dietary_df_raw,by="client",all.x=TRUE)

# preprocessing

## imputation
subdf[subdf==""] <- 0
subdf[is.na(subdf)] <- 0

## normalization
subdf_norm <- subdf %>%
  dplyr::select(-c(client,amount,`Energy(kJ)`,`Calories(kcal)`))
  
subdf_norm <- apply(subdf_norm,2,function(x) x/subdf$`Calories(kcal)`)

if (is.null(nrow(subdf_norm))) next

subdf_norm <- as.data.frame(subdf_norm) %>%
  `rownames<-`(subdf$client) %>%
  rownames_to_column("client")

## log 
subdf_log <- subdf_norm
subdf_log[,-1] <- log(subdf_norm[,-1])

# LETS deal with -Inf -> -10
subdf_log[subdf_log==-Inf] <- -10

## scaling
subdf_scaled <- subdf_log
subdf_scaled[,-1] <- scale(subdf_scaled[,-1], center = TRUE, scale = TRUE)

subdf_scaled <- subdf_scaled %>% column_to_rownames("client")

# remove nans - no variance
subdf_filtered <- subdf_scaled[, colMeans(is.na(subdf_scaled)) < 0.5]
subdf_filtered <- subdf_filtered[rowMeans(is.na(subdf_scaled)) < 0.5,]

if ((ncol(subdf_filtered)<5) | (nrow(subdf_filtered)<5)) next 
data.pca <- prcomp(subdf_filtered)

data_fg <- as.data.frame(data.pca$x[,1:10])
data_metab <- mofa_integrated_df[,1:10] %>% 
  dplyr::filter(rownames(data_metab) %in% rownames(data_fg))


x <- sCCA(dietary_df, mofa_integrated_df, penalization = "enet")
cca_result <- cc(data_fg, data_metab)
U <- as.data.frame(cca_result$scores$xscores)

# Number of observations
n <- nrow(data_fg)
# Number of variables in each dataset
p <- ncol(data_fg)
q <- ncol(data_metab)
# Calculate p-values for the canonical correlations
p_values <- p.asym(cca_result$cor, n, p, q, tstat = "Wilks")
# View the p-values
p_values

}
```

```{r}
# Load library
library(PMA)

# Assume your data is in two matrices:
#   - diet_matrix: rows = individuals, columns = dietary variables
#   - metabo_matrix: rows = individuals, columns = metabolite intensities

# Step 1: Scale the data (important for CCA)
diet_scaled <- dietary_df
metabo_scaled <- mofa_integrated_df

# Step 2: Tune penalty parameters (sparsity levels)
perm.out <- CCA.permute(x = diet_scaled, z = metabo_scaled, typex = "standard", typez = "standard")

# Step 3: Run sCCA with chosen penalties
cca.out <- CCA(x = diet_scaled, z = metabo_scaled, typex = "standard", typez = "standard",
               penaltyx = perm.out$bestpenaltyx,
               penaltyz = perm.out$bestpenaltyz,
               K = 1)  # K = number of canonical components

# Step 4: View results
# Canonical correlation
cca.out$cor

# Variable loadings (non-zero values indicate selected variables)
selected_diet_vars <- which(cca.out$u != 0)
selected_metabo_vars <- which(cca.out$v != 0)

diet_loadings <- cca.out$u[selected_diet_vars]
metabo_loadings <- cca.out$v[selected_metabo_vars]

# Display selected variables and loadings
cat("Selected dietary variables:\n")
print(colnames(diet_scaled)[selected_diet_vars])
print(diet_loadings)

cat("\nSelected metabolites:\n")
print(colnames(metabo_scaled)[selected_metabo_vars])
print(metabo_loadings)

# Optional: Biplots
plot(cca.out$u, cca.out$v, main = "sCCA Loadings", xlab = "Diet", ylab = "Metabolites")
abline(h=0, v=0, lty=2)

```

```{r}
a <- data.frame(MOFA1=mofa_integrated_df$Factor1,
                  DIET1=data.pca$x[,"PC1"]) %>%
  dplyr::filter(MOFA1<10)

ggplot(data=a) + 
  geom_point(aes(x=MOFA1,DIET1))
```
