---
title: "Metabolomics + VGscreener data"
format: html
editor: visual
---

This code serves for exploring patterns in metabolomics of vegans and omnivores.

```{r}
suppressWarnings({
  library(ggplot2)
library(cluster)    # For silhouette
library(factoextra) # For easy plotting (optional)
library(clusterCrit) # For internal clustering metrics (like CH index)
library(tidyverse)
library(mclust)
library(umap)
library(ggpubr)
library(Hmisc)
library(vegan)
library(caret)
library(ranger)
library(doParallel)
library(dummy)
library(patchwork)
  library(pROC)
})

```

# Functions

## Visualization

```{r}
umap_plot <- function(df, metadata,color_col="Cluster",
                      shape_col=NULL,
                      neighbors=10,
                      min_dist=0.1){
  
  custom.config <- umap.defaults
  custom.config$random_state <- 123
  custom.config$n_neighbors <- neighbors
  custom.config$min_dist <- min_dist
  
  umap_plot <- umap(df,config=custom.config)
  data_umap <- data.frame(umap_plot$layout)
  data_umap <- merge(data_umap %>% rownames_to_column("study_id"), 
                     metadata,by="study_id")
  
  if (is.null(shape_col)){
    p <- ggplot(data=data_umap, aes(x=X1,y=X2,color=!!sym(color_col))) + 
      geom_point() + 
      theme_bw() +
      xlab("UMAP 1") + ylab("UMAP 2") 
  } else {
    p <- ggplot(data=data_umap, aes(x=X1,y=X2,color=!!sym(color_col),shape=!!sym(shape_col))) + 
      geom_point() + 
      theme_bw() +
      xlab("UMAP 1") + ylab("UMAP 2") 
  }
    
  
  return(p)
}

pca_plot <- function(df,metadata,color_col,perform_pca=TRUE){
  
  colors = switch(  
    color_col,
    "eq_included_group"= 
      c("vegan" = "#1b9e77", "omnivore" = "#7570b3"), 
    "Country"= c(
      BE = "#FDDA24",   # gold
    CZ = "#11457E",   # blue
    DE = "#000000",   # black
    ES = "#630356",   # purple
    CH = "#D52B1E"    # red
    ),
    "age" = c("#deebf7", "darkblue"),
    "sex" = c(
      M = "#1f77b4",     # blue
      F = "#e377c2",   # pink/purple
    OTHER = "#7f7f7f"     # gray
    ),
    "bmi" = c("#ffffb2", "#bd0026"),
    "residence_area"=c(
      "Rural" = "darkgreen",   # green — nature, countryside
      "Urban" = "navy"   # purple — modern, built environment
      ),
    "VS_total" = c("#D73027", "#1A9850"),
    "eq_vegan_dur_total" = c("#deebf7", "darkblue"),
    NULL
    )  
  
  if (perform_pca){
    data.pca <- prcomp(df,scale. = TRUE,center = TRUE)
    imp_vec <- data.pca$sdev^2
    x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
    y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
    pca_df  <- as.data.frame(data.pca$x)
  } else {
    x_lab <- "PC1"
    y_lab <- "PC2"
    pca_df <- df
  }
  
  pca_df <- merge(pca_df %>% rownames_to_column("study_id"),
                        metadata,
                        by="study_id")

  p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_col))) +
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank(),
        legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab) 
  
  if (is.character(pca_df[,color_col]) & !is.null(colors)) {
    p <- p + 
      scale_color_manual(values=colors) + 
      stat_ellipse(aes(x=PC1,y=PC2,color=!!sym(color_col)),
               show.legend = FALSE)
  } else if (!is.null(colors)){
    p <- p + scale_color_gradient(low = colors[1],high = colors[2]) 
  } 

  return(p)
}

roc_curve <- function(model_object, 
                      group,
                      CI=FALSE){
  # Generates a ROC curve as the performance metric of a model 
  # inputs:
  # model_object - model object
  # group - names of groups that were compared, e.g. c("rPSC","non-rPSC")
  # outputs:
  # p - roc_curve
  
  family <- ifelse(length(unique(model_object$predictions$outcome))==2,
                   "binomial",
                   "multinomial")
  
  if (family=="binomial"){
    roc_c <- make_bi_roc_curve(
      model_object,
      group,
      CI=CI) 
  } else if (family=="multinomial") {
    roc_c <- make_multi_roc_curve(model_object,CI=CI) + 
      ggtitle(paste0(
        "Country",
        ' (AUC = ', round(model_object$model_summary$auc,3), 
        ', P = ',(mean(model_object$valid_performances$auc_validation<0.5)*3),
        ')'
      ))  
  }
  
  roc_c <- roc_c + 
    geom_segment(aes(x = 0, y = 0, xend = 1, yend = 1), 
                 inherit.aes = FALSE, linetype = "dashed", color = "grey")
    return(roc_c)
}

make_bi_roc_curve <- function(model_object,group,CI=FALSE){
  if (CI){
    p <- ggplot()
    color <- "#FF6347"
    
    # load the object 
    auc_val <- model_object$model_summary$auc
                  
    aucs <- sapply(model_object$kfold_rocobjs,function(df) df$auc)
    q1_auc <- quantile(aucs, probs = 0.025)
    q3_auc <- quantile(aucs, probs = 0.975)
    
    differences <- abs(aucs - q1_auc)
    min_auc <- which.min(differences)
    
    differences <- abs(aucs - q3_auc)
    max_auc <- which.min(differences)
    
    differences <- abs(aucs - auc_val)
    true_auc <- which.min(differences)
    
    c <- model_object$kfold_rocobjs[[true_auc]]
    ggroc_data_c <- ggroc(c)$data

    # Create a common set of x-axis points
    x_common <- seq(0, 1, length.out = 100)
    my_df <- data.frame(x=x_common)
    
    for (j in 1:length(aucs)){
      # get ggrocdata
      ggroc_data <- ggroc(model_object$kfold_rocobjs[[j]])$data
      
      # adjust the lowest and highest numbers
      ggroc_data$`1-specificity` <- 
        ggroc_data$`1-specificity` + (1:nrow(ggroc_data))/(1e+20)
      ggroc_data$`1-specificity`[nrow(ggroc_data)] <- 1
      
      # approximation (for equal axis coordinates)
      y_aprox <- approx(
        ggroc_data$`1-specificity`, 
        ggroc_data$sensitivity, xout = x_common)$y
      
      my_df <- cbind(my_df,y_aprox)
    }
    
    my_df <- my_df %>% column_to_rownames("x")
    q1_y <- c()
    q2_y <- c()
    q3_y <- c()
    for (j in 1:length(x_common)){
      q1_y <- c(q1_y,quantile(as.numeric(my_df[j,]), 
                              probs = 0.025,na.rm=TRUE))
      q2_y <- c(q2_y,mean(as.numeric(my_df[j,])))
      q3_y <- c(q3_y,quantile(as.numeric(my_df[j,]), 
                              probs = 0.975,na.rm=TRUE))
    }
    df <- data.frame(x = x_common, 
                     y1 = q1_y, 
                     y2 = q3_y, 
                     y3 = q2_y,
                     Comparison=paste(group[1],"vs",group[2]))
    df[is.na(df)] <- 0
    
    my_color <- color
    
    roc_c <- ggplot() + 
      geom_ribbon(data=df,
        aes(x =x,ymin = pmin(y1, y2), ymax = pmax(y1, y2),
            fill=Comparison), 
        alpha = 0.5,
        show.legend = TRUE) +
      geom_line(data=df,aes(x=x, y = y3, color=Comparison),
                size=1.5) + 
      theme_minimal() + 
      theme(panel.border = 
              element_rect(color = "black", fill = NA, size = 0),
            panel.grid = element_blank(),
            axis.ticks.x = 
              element_line(size=0.3,color = "black"),
            axis.ticks.y = 
              element_line(size=0.3,color="black"),
            axis.ticks.length = unit(4,"pt")) + 
      ylab("Sensitivity") + xlab("1-specificity") + 
      scale_color_manual(values=color) + 
      scale_fill_manual(values=color) + 
    ggtitle(paste0(
      group[1],' vs ',group[2],
      ' (AUC = ', round(model_object$model_summary$auc,3), 
      ', P = ',(mean(model_object$valid_performances$auc_validation<0.5)*3),
      ')'
      ))  
  } else {
    ggroc_data <- ggroc(model_object$kfold_rocobjs)$data
    roc_c <- ggplot(data=ggroc_data) + 
    geom_line(aes(x=`1-specificity`, y=sensitivity, 
                  by=name, color="red",alpha=0.9)) +
    theme_minimal() + 
    theme(legend.position = "none") + 
    ggtitle(paste0(
      group[1],' vs ',group[2],
      ' (AUC = ', round(model_object$model_summary$auc,3), 
      ', P = ',(mean(model_object$valid_performances$auc_validation<0.5)*3),
      ')'
      ))  
  }
  return(roc_c)
}

make_multi_roc_curve <- function(model_object,CI=FALSE){
  groups <- model_object$mapping
  colors <-  c(
  BE = "#FDDA24",   # gold
  CZ = "#11457E",   # blue
  DE = "#000000",   # black
  ES = "#630356",   # purple
  CH = "#D52B1E"    # red
  )
  
  if (CI){
    roc_c <- ggplot()
    for (group in groups){
      rocs <- lapply(model_object$kfold_rocobjs,function(l){
        true_outcome <- groups[l$response]
        true_outcome <- dummy(as.data.frame(true_outcome))
        colnames(true_outcome) <- paste0(groups,"_true")
        predicted_outcome = l$predictor
        colnames(predicted_outcome) <- paste0(groups,"_pred")
        subdf <- cbind(true_outcome,predicted_outcome) %>%
          dplyr::select(starts_with(group))
        colnames(subdf) <- c("true_outcome","predicted_outcome")
        roc(true_outcome  ~ predicted_outcome , 
                      data = subdf,
                      direction = '<',
                      levels = c(0, 1))
        
      })
      
      aucs <- sapply(rocs,function(x)x$auc)
      auc_val <- mean(aucs) #????? TO DO
      q1_auc <- quantile(aucs, probs = 0.025)
      q3_auc <- quantile(aucs, probs = 0.975)
      
      differences <- abs(aucs - q1_auc)
      min_auc <- which.min(differences)
      
      differences <- abs(aucs - q3_auc)
      max_auc <- which.min(differences)
      
      differences <- abs(aucs - auc_val)
      true_auc <- which.min(differences)
      
      true_roc <- rocs[[true_auc]]
      ggroc_data_true <- ggroc(true_roc)$data
      ggroc_data_true$Group <- "True"
      
      # Create a common set of x-axis points
      x_common <- seq(0, 1, length.out = 100)
      my_df <- data.frame(x=x_common)
      
      for (j in 1:length(aucs)){
        # get ggrocdata
        ggroc_data <- ggroc(rocs[[j]])$data
        
        # adjust the lowest and highest numbers
        ggroc_data$`1-specificity` <- 
          ggroc_data$`1-specificity` + (1:nrow(ggroc_data))/(1e+20)
        ggroc_data$`1-specificity`[nrow(ggroc_data)] <- 1
        
        # approximation (for equal axis coordinates)
        y_aprox <- approx(
          ggroc_data$`1-specificity`, 
          ggroc_data$sensitivity, xout = x_common)$y
        
        my_df <- cbind(my_df,y_aprox)
      }
      
      my_df <- my_df %>% column_to_rownames("x")
      q1_y <- c()
      q2_y <- c()
      q3_y <- c()
      for (j in 1:length(x_common)){
        q1_y <- c(q1_y,quantile(as.numeric(my_df[j,]), 
                                probs = 0.025,na.rm=TRUE))
        q2_y <- c(q2_y,mean(as.numeric(my_df[j,])))
        q3_y <- c(q3_y,quantile(as.numeric(my_df[j,]), 
                                probs = 0.975,na.rm=TRUE))
      }
      df <- data.frame(x = x_common, 
                       y1 = q1_y, 
                       y2 = q3_y, 
                       y3 = q2_y,
                       Comparison=group)
      df[is.na(df)] <- 0
      
      my_color <- colors[group]
      
      roc_c <- roc_c + 
        geom_ribbon(data=df,
                    aes(x=x,ymin = pmin(y1, y2), ymax = pmax(y1, y2),
                        fill=Comparison), 
                    alpha = 0.5,
                    show.legend = TRUE) +
        geom_line(data=df,aes(x=x, y = y3, color=Comparison),
                  size=1.5) + 
        theme_minimal() + 
        theme(panel.border = 
                element_rect(color = "black", fill = NA, size = 0),
              panel.grid = element_blank(),
              axis.ticks.x = 
                element_line(size=0.3,color = "black"),
              axis.ticks.y = 
                element_line(size=0.3,color="black"),
              axis.ticks.length = unit(4,"pt")) + 
        ylab("Sensitivity") + xlab("1-specificity") 
    }
    roc_c <- roc_c + 
      scale_color_manual(values=colors) + 
      scale_fill_manual(values=colors)
  } else {
    p <- ggplot()
    for (i in 1:length(model_object$kfold_rocobjs)){
      true_outcome <- groups[model_object$kfold_rocobjs[[i]]$response]
      true_outcome <- dummy(as.data.frame(true_outcome))
      colnames(true_outcome) <- paste0(groups,"_true")
      predicted_outcome = model_object$kfold_rocobjs[[i]]$predictor
      colnames(predicted_outcome) <- paste0(groups,"_pred")
      subdf <- cbind(true_outcome,predicted_outcome)
      for (group in groups){
        group_subdf <- subdf %>%
          dplyr::select(starts_with(group))
        colnames(group_subdf) <- c("outcome","predicted_num")
        rocobj <- roc(outcome ~ predicted_num, 
                  data = group_subdf,
                  direction = '<',
                  levels = c(0, 1))
        
        ggroc_data <- ggroc(rocobj)$data
        ggroc_data$Country <- group
        p <- p + 
          geom_line(data=ggroc_data,
                    aes(x=`1-specificity`, y=sensitivity,
                    color=Country),alpha=0.9) +
          theme_minimal() + 
          theme(legend.position = "right")  
      
    }
  }
  roc_c <- p + 
    scale_color_manual(name = 'Country',
                       breaks = c('BE', 'CZ', 'DE',"ES","CH"),
                       values = c(
                         'BE' = '#FDDA24', 
                         'CZ' = '#11457E', 
                         'DE' = '#000000',
                         "ES" = '#630356',
                         "CH" = '#D52B1E')
  )
  }
  
  return(roc_c)
}

volcano_plot <- function(data,comparison="diet"){
  if (comparison=="diet"){
    color_coding_labels <- c("Insignificant",
                             "Enriched in vegans",
                             "Enriched in omnivores")
    color_coding_colors <- c("Insignificant"="lightgrey",
                                "Enriched in vegans"="#1b9e77",
                                "Enriched in omnivores"="#7570b3")
    volcano_df <- data %>%
      dplyr::filter(str_starts(variable,"eq_included_group"))
  } else if (comparison=="sex"){
    color_coding_labels <- c("Insignificant",
                             "Enriched in males",
                             "Enriched in females")
    color_coding_colors <- c("Insignificant"="lightgrey",
                                "Enriched in males"="#1f77b4",
                                "Enriched in females"="#e377c2")

    volcano_df <- data %>%
      dplyr::filter(str_starts(variable,"sex"))
  }
  volcano_df <- volcano_df %>%
  dplyr::select(metabolite,Estimate,p_adjusted) %>%
  dplyr::mutate(p_to_plot=-log10(p_adjusted)) %>%
  dplyr::mutate(Color_coding=ifelse(p_adjusted>0.05,color_coding_labels[1],
                                    ifelse(Estimate>0,color_coding_labels[2],
                                           color_coding_labels[3]))) %>%
  dplyr::mutate(Label=ifelse(p_adjusted<0.05,metabolite,""))

maximum <- max(c(abs(min(volcano_df$Estimate)), abs(max(volcano_df$Estimate))))

p <- ggplot(data=volcano_df,aes(x=Estimate,y=p_to_plot,color=Color_coding,
                                label=Label)) + 
  geom_point(shape=19, size=3) + 
  theme_bw() + 
  xlab("Effect size") + ylab("- log"[10]~" p-adjusted value") + 
  scale_color_manual(values = color_coding_colors) + 
  coord_cartesian(xlim = c(-maximum,maximum), clip="on") + 
  geom_text_repel(show.legend = FALSE)

return(p)
}
```

## RF

```{r}

model_preparation <- function(metabo_df,metadata,predicting_col="Group"){
  
  if(predicting_col=="sex"){
    metadata <- metadata %>%
      dplyr::filter(sex %in% c("F","M"))
  }
  
  if (!("study_id" %in% colnames(metabo_df))){
    metabo_df <- metabo_df %>% rownames_to_column("study_id")
  }
  
  metabo_df <- metabo_df %>%
    dplyr::filter(study_id %in% metadata$study_id)
  
  if (nrow(metadata)!=nrow(metabo_df)){
     message("Check the dimensions of provided datasets.")
    return(NULL)
  }

  
    # keep only study_id column and desired column to be predicted
    metadata <- metadata %>% 
      dplyr::mutate(Group=!!sym(predicting_col)) %>%
      dplyr::select(study_id,Group) 
    
    
    prepared_df <- merge(metabo_df,
                        metadata,
                        by="study_id")
    
    
    return(prepared_df)
  
}

sampler <- function(data, 
                    outcome,
                    seed = 123, 
                    clust_var=NULL,
                    sample_method = 'atypboot',
                    N = 10){

  set.seed(seed)
  
  data <- data %>% 
    mutate(obs_id = as.character(1:nrow(data)))
  
  if (is.null(clust_var)) clust_var="obs_id" 
  if (clust_var != 'id'){
    data <- data %>% 
      mutate(id = data[[clust_var]]) %>% 
      dplyr::select(-dplyr::all_of(clust_var))
  }
  
  if (colnames(data[outcome]) != 'outcome'){
    data <- data %>% 
      mutate(outcome = data[[outcome]]) %>% 
      dplyr::select(-dplyr::all_of(c(outcome)))
  }
  
  data <- data %>% 
    mutate(obs_id = as.character(1:nrow(data))) %>%
    dplyr::select(id,dplyr::everything())
  
 if (sample_method == 'atypboot') {
    unique_ids <- unique(data$id)

    train_data <- list()
    valid_data <- list()
    
    for (i in 1:N) {
      repeat {
        train <- data.frame(
          id = sample(unique(data$id),
                      length(unique(data$id)),
                      replace = TRUE))
        
        temp_train <- train %>% 
          left_join(data, 
                    by = 'id', 
                    relationship = "many-to-many")
        
        temp_valid <- data.frame(
          obs_id = data[!data$obs_id %in% temp_train$obs_id, ]$obs_id) %>%
          left_join(data, by = 'obs_id')
        
        #if (family=="binomial"){
          train_outcome <- as.numeric(factor(temp_train$outcome)) - 1 
          valid_outcome <- as.numeric(factor(temp_valid$outcome)) - 1 
        #}
        
        
        if (!mean(train_outcome) %in% c(0, 1) & 
            !mean(valid_outcome) %in% c(0, 1)) {
          train_data[[i]] <- temp_train
          valid_data[[i]] <- temp_valid
          {break} 
        }
      }
    }
    return(list(train_data, valid_data))
  }

}


make_rf_model <- function(data, 
                        outcome = "Group",
                        sample_method = 'atypboot',
                        clust_var=NULL,
                        N = 10, # number of bootstrap datasets
                        overfitting_check=FALSE,
                        seed = 123,
                        reuse=FALSE,
                        file=NULL) {

  # Fits a RF model to dataset 
  # inputs:
  # data - prepared dataframe using binomial_prep(), contains only two groups
  # outcome - vector of outcome (labels)
  # sample_method - atypboot - out-of-sample boostrap
  # clust_var, name of clustering variable (here, it can be Patient), default NULL
  # N, number of bootstrap samples, default 10 (minimum 100 needed for reportable results)
  # family - A string specifying the family for the GLMNET model, default 'binomial'
  # overfitting_check - boolean, if random labels reshuffling should be performed, default FALSE
  # seed - random seed for reproducibility, default 123
  # reuse - boolean, should model just be reloaded?, default FALSE
  # file - name of the file for loading the pre-trained model
  # Q - analysis question - Q1/Q2/Q3..
  # outputs: 
  # rf_model - list(model_summary,valid_performances,
  #                   valid_performance,
  #                   predictions, 
  #                   roc_curve,
  #                   kfold_roc_curves,
  #                   trained_model)
  
  # if we are just reusing precomputed models
  if (reuse){
    # model for overfitting check
      if (overfitting_check) {
        load(
          file.path(
            "../intermediate_files/models_overfitting_check/",
            file,
            "rf_model.RData")
          )
      }
      # normal model
      else load(
        file.path("../intermediate_files/models/",
                  file,
                  "rf_model.RData")
        )
  } else {
    # if we are computing the model from scratch
    set.seed(seed)
    
    # if we are doing overfitting check, lets sample the Groups to be random
    if (overfitting_check) {
     data$Group <- sample(data$Group)
    } else {
      data <- data[sample(1:nrow(data)),] %>%
        `rownames<-`(NULL)
    }
  
    data <- data.frame(data,check.names=TRUE)
    
    ## where to save relevant information
    auc_validation <- vector('double', N)
    accuracy_validation <- vector('double', N)
    predictions <- vector("list", N)
    conf_matrices <- vector("list", N)
    kfold_rocobjs <- vector("list", N)
    
    ## original data in a matrix form
    original_outcome <- base::as.matrix(data[[outcome]])
    
    if(!is.numeric(original_outcome)){
      original_outcome <- factor(original_outcome)
      mapping <- setNames(
        levels(original_outcome), 
        0:(length(levels(original_outcome))-1))
      original_outcome <- as.numeric(factor(original_outcome)) - 1 
    }
    
    if (length(unique(original_outcome))==2){
      family="binomial"
    } else if (length(unique(original_outcome))>2) {
      family="multinomial"
    }
    # prepare the predictors and outcome
    original_predictors <- data %>% 
      dplyr::select(-dplyr::all_of(c("study_id",outcome,clust_var))) %>% 
      as.matrix()
    
    original_outcome <- as.factor(original_outcome)
    
    tune_grid <- expand.grid(
      mtry = seq(1, ncol(original_predictors),2), # Number of predictors 
      min.node.size = c(2, 5), # Minimum node size
      splitrule = "gini"
    )
    
    fitControl <- trainControl(method = "repeatedcv",
                               ## 5-fold CV...
                               number = 5,
                               ## repeated 5 times
                               repeats = 1)
    
    # OPTIMALIZATION Train the random forest model using ranger
    cl <- makePSOCKcluster(5)
    registerDoParallel(cl)

    rf_model <- caret::train(
      x = original_predictors, 
      y = original_outcome, 
      method = "ranger",  # Use kknn method for more flexibility
      tuneGrid = tune_grid,  # Manhattan or Euclidean distance
      metric = "Accuracy", # Specify the evaluation metric
      trControl = fitControl,
      importance = "impurity"
    )
    
    stopCluster(cl)
    
    # store the optimized hyperparameters
    optim_mtry = rf_model$bestTune$mtry
    optim_splitrule = rf_model$bestTune$splitrule
    optim_min.node.size = rf_model$bestTune$min.node.size
    
    # Combine predictors and outcome into a single data frame
    train_data <- data.frame(original_outcome, 
                             original_predictors, 
                             check.names = TRUE)
    
    # RF formula
    fit <- ranger(
      formula = original_outcome ~ ., # Explicitly specify the outcome 
      data = train_data,        # Use the combined data frame
      mtry = optim_mtry,  # Use the best value of mtry 
      splitrule = optim_splitrule, # splitrule
      min.node.size = optim_min.node.size, # minimum node size
      probability = TRUE, # store the probability,
      importance = "impurity"
    )
    
    # predictions
    if (family=="binomial"){
      predicted_num = stats::predict(
        fit, 
        data = as.data.frame(original_predictors))$predictions[,2]
      
      predicted_orig <- ifelse(predicted_num > 0.5, 1, 0)
      
      prediction <- data.frame(
        predicted = predicted_orig,
        predicted_num=predicted_num,
        outcome = original_outcome)
      
      rocobj <- roc(outcome ~ predicted_num, 
                  data = prediction,
                  direction = '<',
                  levels = c(0, 1))
      
    } else if (family=="multinomial"){
      predicted_num = stats::predict(
        fit,
        data = as.data.frame(original_predictors))$predictions
      
      predicted_orig <- apply(predicted_num,1,function(x) which.max(x)) - 1
      
      prediction <- list(
      predicted = predicted_orig,
      predicted_num=predicted_num,
      outcome = original_outcome)
      
      rocobj <- multiclass.roc(
        original_outcome,
        predicted_num,
        levels = colnames(predicted_num))
    }
    
    fitted <- data.frame(
      mtry = optim_mtry,  
      splitrule = optim_splitrule,
      min.node.size = optim_min.node.size,
      auc = rocobj$auc,
      accuracy = mean(prediction$predicted==prediction$outcome))
    
    # BOOTSTRAP
    sampled_data <- sampler(data, 
                            outcome = outcome,
                            clust_var = clust_var,
                            sample_method = sample_method,
                            N = N, 
                            seed = seed)
    
    if(!is.null(clust_var)) clust_var <- "id"
    
    for (i in 1:N){
      
      ## sampled data in a matrix form
      sampled_outcome <- as.matrix(sampled_data[[1]][[i]]$outcome)
      
      if (!is.numeric(sampled_outcome)){
        sampled_outcome <- factor(sampled_outcome)
        mapping <- setNames(levels(sampled_outcome), 0:(length(levels(sampled_outcome))-1))
        sampled_outcome <- as.numeric(factor(sampled_outcome)) - 1 
      }
      
      sampled_outcome <- as.factor(sampled_outcome)
      
      sampled_predictors <- sampled_data[[1]][[i]] %>% 
        dplyr::select(-dplyr::all_of(
          c("study_id", "outcome","id", "obs_id", clust_var))) %>%
        as.matrix()
      
      ## re-optimize parameters
      # OPTIMALIZATION
      
      cl <- makePSOCKcluster(5)
      registerDoParallel(cl)
      
      rf_model_sampled <- caret::train(
        x = sampled_predictors, 
        y = sampled_outcome, 
        method = "ranger",  # rf
        tuneGrid = tune_grid,  # gini
        metric = "Accuracy", # Specify the evaluation metric
        trControl = fitControl,
        importance = "impurity"
      )
      
      stopCluster(cl)
      
      optim_mtry_sampled = rf_model_sampled$bestTune$mtry
      optim_splitrule_sampled = rf_model_sampled$bestTune$splitrule
      optim_min.node.size_sampled = rf_model_sampled$bestTune$min.node.size
      
      # Combine predictors and outcome into a single data frame
      train_data_sampled <- data.frame(sampled_outcome, 
                                       sampled_predictors, 
                                       check.names = TRUE)
      # TRAINING
      sampled_fit <- ranger(
        formula = sampled_outcome ~ .,    # outcome variable 
        data = train_data_sampled,        # Use the combined data frame
        mtry = optim_mtry_sampled,  # Use the best value of k from tuning
        splitrule = optim_splitrule_sampled,
        min.node.size = optim_min.node.size_sampled,
        probability = TRUE
      )
      
      # VALIDATION
      valid_outcome <- as.factor(as.matrix(
        sampled_data[[2]][[i]]$outcome))
      
      valid_outcome <- as.numeric(factor(valid_outcome)) - 1 
      valid_outcome <- as.factor(valid_outcome)
       
      valid_predictors <- sampled_data[[2]][[i]] %>% 
        dplyr::select(-study_id, -outcome, -id, -obs_id) %>% 
        as.matrix()
      
      if (family=="binomial"){
      valid_predicted_num = stats::predict(
        sampled_fit, 
        data = as.data.frame(valid_predictors)
        )$predictions[,2]
      
      valid_predicted_orig <- ifelse(valid_predicted_num > 0.5, 1, 0)
      
      prediction_onValidation <- data.frame(
        predicted = valid_predicted_orig,
        predicted_num=valid_predicted_num,
        outcome = valid_outcome)
      
      kfold_rocobjs[[i]] <- roc(outcome ~ predicted_num, 
                  data = prediction_onValidation,
                  direction = '<',
                  levels = c(0, 1))
      
    } else if (family=="multinomial"){
      valid_predicted_num = stats::predict(
        sampled_fit,
        data = as.data.frame(valid_predictors))$predictions
      
      valid_predicted_orig <- apply(valid_predicted_num,1,function(x) which.max(x)) - 1
      
      prediction_onValidation <- list(
      predicted = valid_predicted_orig,
      predicted_num=valid_predicted_num,
      outcome = valid_outcome,
      iteration = i)
      
      ## record performances
      kfold_rocobjs[[i]] <- multiclass.roc(
        valid_outcome,
        valid_predicted_num,
        levels = colnames(valid_predicted_num))
      
    }
    auc_validation[i] <- kfold_rocobjs[[i]]$auc
    accuracy_validation[i] <- mean(
      prediction_onValidation$predicted == prediction_onValidation$outcome)
    
    prediction_onValidation[["id"]] <- sampled_data[[2]][[i]]$id
    predictions[[i]] <- prediction_onValidation

    }
    
    ## connect predictions
    predictions <- bind_rows(predictions)
    
    ## aggregate obtained information
    valid_performances <- data.frame(
      auc_validation,
      accuracy_validation)
    
    model_summary <- fitted %>% 
      mutate(
        auc = mean(valid_performances$auc_validation),
        auc_CIL = quantile(valid_performances$auc_validation, 
                           probs = 0.025),
        auc_CIU = quantile(valid_performances$auc_validation, 
                           probs = 0.975),
        accuracy = mean(valid_performances$accuracy_validation),
        accuracy_CIL = quantile(valid_performances$accuracy_validation, 
                                probs = 0.025),
        accuracy_CIU = quantile(valid_performances$accuracy_validation, 
                                probs = 0.975)
        )

    ## define outputs
    rf_model <- list(model_summary = model_summary, 
                    valid_performances = valid_performances, 
                    predictions = prediction, 
                    rocobj=rocobj,
                    kfold_rocobjs=kfold_rocobjs,
                    trained_model=fit,
                    mapping=mapping)
    
    # save results
    if (overfitting_check){
     if (!dir.exists(file.path("../intermediate_files/models_overfitting_check/",file))){
      dir.create(file.path("../intermediate_files/models_overfitting_check/",file))
     } 
      
      save(rf_model,
           file=file.path("../intermediate_files/models_overfitting_check/",
                          file,
                          "rf_model.RData")
           )
    } else {
      if (!dir.exists(file.path("../intermediate_files/models/",file))){
        dir.create(file.path("../intermediate_files/models/",file))
      }
      save(rf_model,
           file=file.path("../intermediate_files/models/",file,"rf_model.RData"))
    }
  }
  
  return(rf_model)
}
```

## Tests

results_list <- lapply(colnames(spca_integrated_df), 
                       function(x) fit_glm(x,spca_for_uni,
                                           tested_variable = "eq_included_group"))
results_df <- do.call(rbind, results_list)

```{r}
fit_glm <- function(metab,data,
                    tested_variable = NULL,
                    intcheck=FALSE) {
  if (tested_variable=="eq_included_group"){
    if (intcheck) {
    formula <- as.formula(paste(
      metab, "~ age + sex + bmi + eq_included_group"))
    }
    else {
      formula <- as.formula(paste(
      metab, "~ age + sex + bmi + Country * eq_included_group"))
    } 
  } else if (tested_variable=="Country"){
    formula <- as.formula(paste(
      metab, "~  age + sex + bmi + Country"))
  } else if (tested_variable=="sex"){
    if (intcheck) {
    formula <- as.formula(paste(
      metab, "~ age + bmi + eq_included_group + sex"))
    } else {
    formula <- as.formula(paste(
      metab, "~ age + bmi + eq_included_group * sex"))
  }
  }
  
  
  # Try to fit model, handle errors if any
  model <- glm(formula, data = data)

  # Extract coefficients summary, p-values, etc.
  coef_summary <- as.data.frame(summary(model)$coefficients) %>% 
    rownames_to_column("variable")
  
  # Return a data.frame with metabolite name and coefficients
  out <- data.frame(metabolite = metab, coef_summary)
  return(out)
  
}

pairwise_glm <- function(spca_for_uni,tested_variable=NULL){
  co <- data.frame("1"=c("BE","BE","BE","BE","CZ","CZ","CZ","DE","DE","ES"),
                   "2"=c("CZ","DE","ES","CH","DE","ES","CH","ES","CH","CH")) %>% t()
  
  final_results_df <- data.frame()
  for (i in 1:ncol(co)){
    comp <- co[,i]
    subdata <- spca_for_uni %>% 
      dplyr::filter(Country %in% comp)
    
    results_list <- lapply(colnames(spca_integrated_df), 
                           function(x) fit_glm(x,subdata,
                                               tested_variable = tested_variable))
    
    results_df <- do.call(rbind, results_list)
    results_df$Comparison <- paste0(comp[1],"_vs_",comp[2])
    results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
    final_results_df <- rbind(final_results_df,results_df)
  }
  
  df_listed <- final_results_df %>%
    # Filter only Country rows (ignore intercept, age, sex, etc.)
    filter(grepl("^Country", variable)) %>%
    # Only keep significant p-values (adjusted)
    filter(p_adjusted < 0.05) %>%
    # Group by Comparison
    group_by(Comparison) %>%
    # Collect the PC values where significant
    summarise(Significant_PCs = paste(unique(metabolite), collapse = ", ")) %>%
    arrange(Comparison)

  df_summary <- final_results_df %>%
  # Filter rows for country variable only
  filter(grepl("^Country", variable)) %>%
  # Create a flag for significance
  mutate(significant = ifelse(p_adjusted < 0.05, "YES", "NO")) %>%
  # Select only relevant columns
  select(metabolite, Comparison, significant) %>%
  # Pivot wider to get PCs as columns
  pivot_wider(
    names_from = metabolite,
    values_from = significant,
    values_fill = "NO"
  ) %>%
  arrange(Comparison)
  
  return(list(df_listed,df_summary))
}
```

# Import

```{r}

metadata_df <- read.table("../../data/preprocessed/dietary/veganscreener_preprocessed.tsv",header = TRUE)

lcms_df <- read.table("../../data/preprocessed/metabolomics/lc_ms_preprocessed.tsv",check.names = FALSE)

nmr_df <- read.table("../../data/preprocessed/metabolomics/nmr_preprocessed.tsv",check.names = FALSE)

spca_integrated_df <-  read.table("../../data/preprocessed/metabolomics/spca_integrated_all.tsv",
                                  check.names = FALSE)

```

**Get only important samples**

```{r}
# metabolomics
## LCMS
lcms_df <- lcms_df %>%
  dplyr::filter(rownames(lcms_df) %in% metadata_df$study_id)

## NMR
nmr_df <- nmr_df %>%
  dplyr::filter(rownames(nmr_df) %in% metadata_df$study_id)

## SPCA
spca_integrated_df <- spca_integrated_df %>%
  dplyr::filter(rownames(spca_integrated_df) %in% metadata_df$study_id)
```

# Metabolomics dataset

## PCA

### SPCA

```{r,fig.width=10,fig.height=4}
spca_plot_df <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../results/spca_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

```{r,fig.width=10}
plot_list <- list()
for (col in colnames(spca_plot_df)[22:101]){
  p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(col))) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

plot_list[[length(plot_list)+1]] <- p
}

a <- ggarrange(plotlist = plot_list)

```

```{r}
pdf("../Results/spca_PCAs.pdf",height = 20,width = 40)
a
dev.off()
```

### LCMS

```{r}
data.pca <- prcomp(lcms_df)
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>%
                  rownames_to_column("study_id"),
                metadata_df,by="study_id")

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r,fig.width=10}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

p
```

```{r}
pdf("../Results/lcms_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

```{r,fig.width=10}
plot_list <- list()
for (col in colnames(pca_df)[52:131]){
  p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(col))) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

plot_list[[length(plot_list)+1]] <- p
}

a <- ggarrange(plotlist = plot_list)

```

```{r}
pdf("../Results/lcms_PCAs.pdf",height = 20,width = 40)
a
dev.off()
```

### NMR

```{r}
data.pca <- prcomp(nmr_df)
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>%
                  rownames_to_column("study_id") ,
                metadata_df,by="study_id")


imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```


```{r}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

p
```

```{r}
pdf("../Results/NMR_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

## UMAP

### SPCA

```{r,fig.height=4,fig.width=10}
p <- umap_plot(spca_integrated_df,metadata_df,
               color_col =  "eq_included_group",
               neighbors = 10,
               min_dist = 0.5)


```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(spca_integrated_df,metadata_df,
               neighbors = n_i,
               min_dist = m_i,
               color_col = "eq_included_group") + 
          theme(legend.position = "none")
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### LCMS

```{r,fig.height=4,fig.width=10}
p <- umap_plot(lcms_df,metadata_df,
               color_col =  "eq_included_group",
               neighbors = 100,
               min_dist = 0.5)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(lcms_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### NMR

```{r,fig.height=4,fig.width=10}
p <- umap_plot(nmr_df,metadata_df,
               color_col = "eq_included_group",
               neighbors = 50,
               min_dist = 0.1)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(nmr_df,metadata_df,
               neighbors = n_i,
               min_dist = m_i,
               color_col = "eq_included_group") + 
          theme(legend.position = "none")
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

For now, we have decided to exlude the outliers from the analysis

```{r}
lcms_df <- lcms_df %>%
  dplyr::filter(!(rownames(lcms_df) %in% outliers$study_id))

spca_integrated_df <- spca_integrated_df %>%
  dplyr::filter(!(rownames(spca_integrated_df) %in% outliers$study_id))
```

## Clustering

### k-means

#### SPCA

```{r}
data <- spca_integrated_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("study_id") %>%
                       `colnames<-`(c("study_id","Cluster")),
                     by="study_id",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(spca_integrated_df,metadata_clusters,
               shape_col = "eq_included_group",
               neighbors = 50,
               min_dist = 0.3)

p
```

**PCA**

```{r}
spca_plot_df <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_clusters,by="study_id")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster,shape=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

#### LCMS

```{r}
data <- lcms_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("study_id") %>%
                       `colnames<-`(c("study_id","Cluster")),
                     by="study_id",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(lcms_df,metadata_clusters,
               shape_col = "eq_included_group",
               neighbors = 50,
               min_dist = 0.3)

p
```

**PCA**

```{r}

lcms_pca_df  <- as.data.frame(prcomp(lcms_df,scale. = TRUE,center = TRUE)$x)

lcms_plot_df <- merge(lcms_pca_df %>% rownames_to_column("study_id"),
                      metadata_clusters,by="study_id")

p <- ggplot(lcms_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster,shape=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

#### NMR

```{r}
data <- nmr_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("study_id") %>%
                       `colnames<-`(c("study_id","Cluster")),
                     by="study_id",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(nmr_df,metadata_clusters,
               shape_col = "eq_included_group",
               neighbors = 50,
               min_dist = 0.3)

p
```

**PCA**

```{r}

nmr_pca_df  <- as.data.frame(prcomp(nmr_df,scale. = TRUE,center = TRUE)$x)

nmr_plot_df <- merge(nmr_pca_df %>% rownames_to_column("study_id"),
                      metadata_clusters,by="study_id")


p <- ggplot(nmr_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=Cluster,shape=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```


# Global difference

```{r}
metadata_df <- read.table(
  "../../data/preprocessed/dietary/veganscreener_preprocessed.tsv",
  header = TRUE,
  check.names = FALSE) 

spca_integrated_df <-  read.table(
  "../../data/preprocessed/metabolomics/spca_integrated_all.tsv",
  check.names = FALSE )

metadata_df <- metadata_df %>%
  dplyr::filter(study_id %in% rownames(spca_integrated_df))

spca_integrated_df <- spca_integrated_df %>%
  dplyr::filter(rownames(spca_integrated_df) %in% metadata_df$study_id)
```

## PERMANOVA

We are testing how much of the variation in metabolomics is explained by:
 - sex
 - age
 - bmi
 - Country
 - diet group
 

Variables were added sequentially (Type I SS), so each p-value and R² reflects a partial effect, after adjusting for those above.
This model accounts for variation due to sex, age, and BMI. Then accounts for the main effects of country and dietary group. Then test whether the effect of dietary group varies by country (interaction).

Diet group explains **8.3%** of variation after adjusting for sex, age,bmi and country.

Country explains an **10%** of variation. The interaction effect is also significant (p=0.002), which suggest the effect of diet is different in each country. We will epxlore this in further analysis.

```{r}

# Create Euclidean distance matrix
dist_mat <- vegdist(spca_integrated_df, 
                    method = "euclidean")

# Test whether vegan vs omnivore differ in metabolomics
adonis_result <- adonis2(
  dist_mat ~ sex + age + bmi + Country*eq_included_group, 
  data = metadata_df , 
  permutations = 999,
  by="terms")

print(adonis_result)

```

## DISPERSION EFFECT

The dispersion effect is significant. However, as the two centroids are clearly separate, we will report this result as significant.

```{r}
# Test for homogeneity of dispersions
group <- metadata_df$eq_included_group
disp <- betadisper(dist_mat, group)
anova(disp)
permutest(disp)     # Permutation test
plot(disp)   
boxplot(disp)          # Distances per group

```

## PCA

```{r,fig.width=10}
plot_list <- list()
for (col in c("eq_included_group","Country","sex","age","bmi")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE) 

  plot_list[[length(plot_list)+1]] <- p 
}

```

```{r,fig.width=10}
a <- plot_list[[1]] + plot_list[[2]] + plot_list[[3]] + plot_list[[4]] + plot_list[[5]]

a
```

```{r,eval=FALSE}
pdf("../Results/pca_main_variables_spca.pdf",height = 5,width = 10)
a
dev.off()
```

## RF

**Diet**

```{r}
model_name <- "diet_model" 

prepared_df <- model_preparation(spca_integrated_df,
                                 metadata_df,
                                 predicting_col = "eq_included_group")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 10,
                          file = model_name,
                          reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("omnivores","vegans"),CI = TRUE)

roc_c
```

**Country**

```{r}
model_name <- "country_model" 

prepared_df <- model_preparation(spca_integrated_df,
                                 metadata_df,
                                 predicting_col = "Country")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 10,
                          file = model_name,
                          reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model,CI = TRUE)

roc_c
```

**Sex**

```{r}
model_name <- "sex_model" 

prepared_df <- model_preparation(spca_integrated_df,
                                 metadata_df,
                                 predicting_col = "sex")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 10,
                          file = model_name,
                          reuse=FALSE)

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("Female","Male"),CI = TRUE)

roc_c
```

# Univariate analysis

We will see which PC we can consider significant, then we will explore which metabolites are linked to this PC. From the visualization, there can be significant result from PC2.

## Diet-specific

### glm

```{r,fig.height=15}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE)

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = col, fill = "eq_included_group")) +
    geom_density(alpha = 0.5) +
    ggtitle(col) + 
    theme_minimal()
  
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list)

a
```

```{r,eval=FALSE}
pdf("../Results/preprocessing/density_spca_diet.pdf",height = 10,width = 20)
a
dev.off()
```

Call glm to explore which metabolites had significant different between dietary groups.

```{r}
results_list <- lapply(colnames(spca_integrated_df), 
                       function(x) fit_glm(x,spca_for_uni,
                                           tested_variable = "eq_included_group"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results - **PC1**, **PC3**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="eq_included_groupvegan")

```

Check the interaction effect - **none**

```{r}
df <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(str_ends(variable, ":eq_included_groupvegan"))

df

pc_to_be_removed <- df$metabolite
```

Find the individual directions within each country, mark the particular PC as "to_remove" if the direction is not the same across countries.

**No need because it is not significant**

```{r,eval=FAlSE}
# variable for saving the results
final_df <- data.frame()

# for each country
for (country in unique(spca_for_uni$Country)) {
  # create subdf with this country only
  subdf <- spca_for_uni %>%
    dplyr::filter(Country==country)
  
  # check if we have omnivores as well as vegans in one country
  if (length(unique(subdf$eq_included_group))>1){
    # call the fit_glm for interaction check - will check group
    # adjusted for sex, bmi, age
    country_results_list <- lapply(
      pc_to_be_removed, 
      function(x) fit_glm(x,subdf,tested_variable = "eq_included_group",
                          intcheck = TRUE))
    
    # consolidate the results into dataframe
    country_results_df <- do.call(rbind, country_results_list)
    country_results_df$Country <- country
    
    # save the country-specific results to one final df
    final_df <- rbind(final_df,country_results_df)
  }
}

# see the estimates for diet groups arrange by individual PCs 
final_df <- final_df %>%
  dplyr::filter(str_starts(variable,"eq_included_group")) %>%
  dplyr::arrange(metabolite)

# see the results
final_df

# save the PCs that do not have the same directions across countries
pc_to_remove <- final_df %>%
  group_by(metabolite) %>%
  summarise(
    has_positive = any(Estimate > 0),
    has_negative = any(Estimate < 0)
  ) %>%
  filter(has_positive & has_negative) %>%
  pull(metabolite)
```

Keep only PCs that have same direction when significant interaction.

```{r}
pc_to_remove <- NULL
significant_pcs <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="eq_included_groupvegan") %>%
  dplyr::filter(!(metabolite %in% pc_to_remove))

significant_pcs
```

**Volcano plot**

```{r}
vp <- volcano_plot(results_df)
vp
```

### RF importance

There is a clear pattern and match with univariate analysis results. When taking account also the density plot, we see that **PC1**, **PC3** are definitely important in the separation of omnivores and vegans.

```{r}
a <- load("../intermediate_files/models/diet_model/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) %>%
  dplyr::mutate(PC=factor(PC,levels = paste0("PC",1:20)))

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Loadings

We will analyze the metabolites that contribute to **PC1** and **PC3** the most.

```{r}
load("../results/preprocessing/spca_result_all.RData")
```

**Increased in vegans**

```{r}
loadings <- spca_result$rotation[, 1]  # loadings for PC1

# Sort by absolute value to find strongest contributors
top_vars <- sort(abs(loadings), decreasing = TRUE)

# View top 10 variables contributing to PC1
#top_vars[1:10]

# View the variables with the "significant" loadings, the threshold is determined
# by the formula 1/sqrt(number of variables)
increased_in_vegans <- top_vars[top_vars>(1/(sqrt(length(top_vars))))]
increased_in_vegans
```

```{r}
loadings <- spca_result$rotation[, 3]  # loadings for PC3

# Sort by absolute value to find strongest contributors
top_vars <- sort(abs(loadings), decreasing = TRUE)

# View the variables with the "significant" loadings, the threshold is determined
# be the formula 1/sqrt(number of variables)
decreased_in_vegans <- top_vars[top_vars>(1/(sqrt(length(top_vars))))]
decreased_in_vegans
```

```{r}
list_to_save <- list(
  "increased_in_vegans"=
    as.data.frame(increased_in_vegans) %>% rownames_to_column("metabolite"),
  "increased_in_omnivores"= 
    as.data.frame(decreased_in_vegans) %>% rownames_to_column("metabolite")
  )

write.xlsx(list_to_save,
  "../results/diet_associated_metabolites.xlsx",rownames=TRUE)
```

### Test of metabolites

```{r}
lc_ms_df <- read.table("../../data/preprocessed/metabolomics/lc_ms_preprocessed.tsv",check.names = FALSE)

nmr_df <- read.table("../../Data/preprocessed/metabolomics/nmr_preprocessed.tsv",check.names = FALSE)

metadata_df <- read.table("../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",header=TRUE)

lcms_df <- lcms_df %>%
  dplyr::filter(rownames(lcms_df) %in% metadata_df$study_id)

## NMR
nmr_df <- nmr_df %>%
  dplyr::filter(rownames(nmr_df) %in% metadata_df$study_id)

common_samples <- intersect(rownames(lc_ms_df), rownames(nmr_df))

lc_ms_df <- lc_ms_df[common_samples,]

nmr_df <- nmr_df[common_samples,]

```

```{r}
integrated_df <- cbind(lc_ms_df,nmr_df) %>%
  dplyr::select(names(increased_in_vegans),names(decreased_in_vegans))

df_for_uni <- merge(integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE)

```

```{r}
results_list <- lapply(colnames(integrated_df), 
                       function(x) fit_glm(x,df_for_uni,
                                           tested_variable = "eq_included_group"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results - **PC1**, **PC3**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="eq_included_groupvegan")

```
## Country-specific

### glm

There is a visual separation in PC1, PC3, PC4 in the density plot.

```{r,fig.height=15}
# prepare for testing
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE)

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = col, fill = "Country")) +
    geom_density(alpha = 0.5) +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values = c(
      BE = "#FDDA24",   # gold
      CZ = "#11457E",   # blue
      DE = "#000000",   # black
      ES = "#630356",   # purple
      CH = "#D52B1E"    # red
    ))
  
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list)

a
```

```{r,eval=FALSE}
pdf("../Results/preprocessing/density_spca_country.pdf",height = 10,width = 20)
a
dev.off()
```

Call glm to explore which metabolites had significant different between countries. There is a signal in **PC1**, **PC4**,**PC5**,**PC7**,**PC14**

```{r}
glm_res <- pairwise_glm(spca_for_uni,tested_variable = "Country")

# list of significances
glm_res[[1]]

# summarized significances
glm_res[[2]]

# in how many comparisons was specific PC significant?
colSums(glm_res[[2]]=="YES")
```

### RF importance

There is a clear pattern and match with univariate analysis results. When taking account also the density plot, we see that **PC1**,**PC3**, and **PC4** are important in the separation of countries.

```{r}
a <- load("../intermediate_files/models/country_model/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) %>%
  dplyr::mutate(PC=factor(PC,levels = paste0("PC",1:20)))

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Loadings ?

As there are no clearly important PC, we will not extract the metabolites from the loadings.

## Sex-specific

### glm

There can be a signal in **PC1** and **PC5**

```{r,fig.height=15}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE) %>%
  dplyr::filter(sex %in% c("F","M"))

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = col, fill = "sex")) +
    geom_density(alpha = 0.5) +
    ggtitle(col) + 
    theme_minimal()
  
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list)

a
```

```{r,eval=FALSE}
pdf("../Results/preprocessing/density_spca_sex.pdf",height = 10,width = 20)
a
dev.off()
```

Call glm to explore which metabolites had significant different between males and females.

```{r}
results_list <- lapply(
  colnames(spca_integrated_df),
  function(x) fit_glm(x,spca_for_uni,tested_variable = "sex"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results.

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="sexM")

```

Check the interaction effect.

```{r}
df <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(str_ends(variable, ":sexM"))

df

pc_to_be_removed <- df$metabolite
```

Find the individual directions within each country, mark the particular PC as "to_remove" if the direction is not the same across countries.

```{r,eval=FALSE}
final_df <- data.frame()
for (country in unique(spca_for_uni$Country)) {
  subdf <- spca_for_uni %>%
    dplyr::filter(Country==country)
  
  if (length(unique(subdf$eq_included_group))>1){
    country_results_list <- lapply(
      pc_to_be_removed, 
      function(x) fit_glm(x,subdf,tested_variable = "sex", intcheck = TRUE))
    country_results_df <- do.call(rbind, country_results_list)
    country_results_df$Country <- country
    final_df <- rbind(final_df,country_results_df)
  }
}

final_df <- final_df %>%
  dplyr::filter(str_starts(variable,"sex")) %>%
  dplyr::arrange(metabolite)

final_df

pc_to_remove <- final_df %>%
  group_by(metabolite) %>%
  summarise(
    has_positive = any(Estimate > 0),
    has_negative = any(Estimate < 0)
  ) %>%
  filter(has_positive & has_negative) %>%
  pull(metabolite)
```

Keep only PCs that have same direction when significant interaction.

```{r}
significant_pcs <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="sexM") %>%
  dplyr::filter(!(metabolite %in% pc_to_remove))

significant_pcs
```

```{r}
vp <- volcano_plot(results_df,comparison = "sex")
vp
```

### RF importance

There is a clear pattern and match with univariate analysis results. We can say that **PC8** is definitely important for SEX separation, as there is clear pattern in density plot, glm analysis and high RF variable importance.

```{r}
a <- load("../intermediate_files/models/sex_model/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) %>%
  dplyr::mutate(PC=factor(PC,levels = paste0("PC",1:20)))

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Loadings ??

We will analyze the metabolites that contribute to **PC1**, **PC4**, **PC5**, and **PC6** the most.

```{r}
load("../Results/preprocessing/spca_result.RData")
```

```{r}
loadings_PC8 <- spca_result$rotation[, 8]  # loadings for PC8

# Sort by absolute value to find strongest contributors
top_vars <- sort(abs(loadings_PC8), decreasing = TRUE)

# View top 10 variables contributing to PC8
top_vars[1:10]

# View the variables with the "significant" loadings, the threshold is determined
# be the formula 1/sqrt(number of variables)
top_vars[top_vars>(1/(sqrt(length(top_vars))))]
```

# Vegan analysis

Now, we will focus only on vegan participants. We will conduct:

-   metabolomics integration again with vegan participants only

-   exploratory analysis if we see the pattern in metabolites and VSscore

-   correlation analysis between metabolites and VSscore

-   ML approach to see if we can distinguish between top and bottom tertile of VSscore

-   Can we identify markers that are associated with high or low VSscore?

-   Can we propose a metabolite-based vegan index?

-   stratification of the proposed analysis by:

    -   time since vegan (new vs old vegans)

    -   age group

    -   sex

    -   country

**Import**

```{r}
metadata_df <- read.table("../../data/preprocessed/dietary/veganscreener_preprocessed.tsv",header = TRUE)

spca_integrated_df <-  read.table("../../data/preprocessed/metabolomics/spca_integrated.tsv")

metadata_vegans_df <- metadata_df %>%
  dplyr::filter(eq_included_group=="vegan")

spca_vegans_df <- spca_integrated_df %>%
  rownames_to_column("study_id") %>%
  dplyr::filter(study_id %in% metadata_vegans_df$study_id) %>%
  column_to_rownames("study_id")

metadata_vegans_df <- metadata_vegans_df %>% 
  dplyr::filter(study_id %in% rownames(spca_vegans_df))
```

## Exploratory analysis

### PCA

**with outliers**

```{r,fig.width=10,fig.height=4}
spca_plot_df <- merge(spca_vegans_df %>% rownames_to_column("study_id"),
                      metadata_vegans_df,
                      by="study_id")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/spca_PCA_vegans.pdf",height = 4,width = 6)
p
dev.off()
```

```{r}
outliers <- spca_vegans_df %>% 
  rownames_to_column("study_id") %>%
  dplyr::filter(PC1< -40) %>% dplyr::select(study_id)
```

**Without outliers**

```{r,fig.width=10,fig.height=4}
spca_plot_df <- merge(spca_vegans_df %>% rownames_to_column("study_id"),
                      metadata_vegans_df,
                      by="study_id") %>%
  dplyr::filter(!(study_id %in% outliers$study_id))


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/spca_PCA_vegans_nooutliers.pdf",height = 4,width = 6)
p
dev.off()
```

### UMAP

**With outliers**

```{r,fig.height=4,fig.width=10}
p <- umap_plot(spca_vegans_df,metadata_vegans_df,
               color_col =  "eq_included_group",
               neighbors = 10,
               min_dist = 0.5)

p
```

**Without outliers**

```{r,fig.height=4,fig.width=10}
p <- umap_plot(spca_vegans_df %>%
                 rownames_to_column("study_id") %>%
                 dplyr::filter(!(study_id %in% outliers$study_id)) %>%
                 column_to_rownames("study_id"),
               metadata_vegans_df %>%
                 dplyr::filter(!(study_id %in% outliers$study_id)),
               color_col =  "eq_included_group",
               neighbors = 10,
               min_dist = 0.5)

p
```

We will proceed without outliers

```{r}
metadata_vegans_df <-  metadata_vegans_df %>%
                 dplyr::filter(!(study_id %in% outliers$study_id))

spca_vegans_df <- spca_vegans_df %>%
                 rownames_to_column("study_id") %>%
                 dplyr::filter(!(study_id %in% outliers$study_id)) %>%
                 column_to_rownames("study_id")

```

### PERMANOVA

#### By terms

We are testing how much of the variation in metabolomics is explained by:

-   VSscore (main variable)

-   Country

-   sex

-   age

-   bmi

Variables were added sequentially (Type I SS), so each p-value and R² reflects a partial effect, after adjusting for those above.

VSscore explains **5.8%** of variation **before any adjustment**. Strong effect. Country explains an **additional \~12.8%** of variation, after VSscore is accounted for. Sex adds \~3.6% more variance explanation. Age has small but significant effect. BMI has also a small effect, but statistically significant. \~75.6% of the variation remains unexplained.

```{r}

# Create Euclidean distance matrix
dist_mat <- vegdist(spca_vegans_df, 
                    method = "euclidean")

# Test whether vegan vs omnivore differ in metabolomics
adonis_result <- adonis2(dist_mat ~ sex + age + bmi + Country + VS_total, 
                         data = metadata_vegans_df, 
                         permutations = 999,
                         by="terms")

print(adonis_result)

```

#### By margin

```{r,eval=FALSE}

# Create Euclidean distance matrix
vars <- colnames(metadata_vegans_df)[c(2,3,4,5,9,14:18,53:81)]

permanova_metadata_df <- metadata_vegans_df %>%
  dplyr::select(c("study_id",vars)) %>%
  drop_na()

permanova_spca_df <- spca_vegans_df[as.character(permanova_metadata_df$study_id),]

dist_mat <- vegdist(permanova_spca_df, 
                    method = "euclidean")


fml <- as.formula(paste("dist_mat ~", paste(vars, collapse = " + ")))

# Test whether vegan vs omnivore differ in metabolomics
adonis_result <- adonis2(fml,
                         data = permanova_metadata_df, 
                         permutations = 999,
                         by="margin")

adonis_result$p_adjusted <- p.adjust(adonis_result$Pr..F.,method = "BH")

write.table(adonis_result,"../Results/adonis_vegan_by_margin.tsv",
            sep="\t")

print(adonis_result)

```

```{r}
adonis_result <- read.table("../Results/adonis_vegan_by_margin.tsv") %>%
  rownames_to_column("Variable") %>%
  dplyr::mutate(Significance = ifelse(p_adjusted < 0.001, "***", ifelse(p_adjusted < 0.01, "**", ifelse(p_adjusted < 0.05, "*", "")))) %>%
  dplyr::filter(!(Variable %in% c("Total", "Residual")))

ggplot(adonis_result, aes(x = reorder(Variable, R2), y = R2, fill = Significance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "PERMANOVA R² by Variable",
       x = "Variable",
       y = "R² (Variance Explained)") +
  scale_fill_manual(values = c("***" = "red", "**" = "orange", "*" = "gold", "." = "gray70", "ns" = "gray90")) +
  theme_minimal()

```

### Color-coded PCA

```{r,fig.width=10}
plot_list <- list()
for (col in c("VS_total","Country","sex","age","bmi",
              "eq_vegan_dur_total","vegan_main_motive",
              "education","family_income","residence_area",
              "sleep_duration")){
  p <- pca_plot(spca_vegans_df,
              metadata_vegans_df,
              color_col = col,
              perform_pca = FALSE) 

  plot_list[[length(plot_list)+1]] <- p 
}

```

```{r,fig.width=15,fig.height=20}
a <- wrap_plots(plot_list,ncol = 2)

a
```

## Can we model the VSscore with metabolites? Regression random forest

## Can we model individual metabolites using VSscore? glm, correlation

## Can we distinguish between top and bottom tertile of VSscore with metabolites? RF

# Dietary dataset

## PCA

```{r}
data.pca <- prcomp(dietary_df)
plot(data.pca)
```

```{r,fig.height=10,fig.width=10}
library(factoextra)

p <- fviz_pca_var(data.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

p
```

```{r}
pdf("../Results/dietary_PCA_biplot.pdf",height = 20,width = 20)
p
dev.off()
```

```{r,fig.height=10,fig.width=10}

p <- ggplot(data.pca$x,aes(x = PC1,y=PC2)) + 
  geom_point()

p
```

```{r}
pdf("../Results/dietary_PCA_samples.pdf",height = 20,width = 20)
p
dev.off()
```

## Correlation analysis

```{r,fig.height=20, fig.width=20}
library(ggcorrplot)
corr <- round(cor(dietary_df), 1)
p <- ggcorrplot(corr, method = "circle")

p
```

## Clustering

### Dendrogram

**Whole dataset**

```{r,fig.width=20,fig.height=10}
dist_mat <- dist(dietary_df, method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)

clusters_dendrogram_A <- cutree(hclust_avg, k = 10)
```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(factor(clusters_dendrogram_A,levels = c(1,2,3,4,5,6,7,8,9,10))) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  scale_color_manual(values=c("red","blue","green","grey","black",
                              "darkred","darkblue","darkgreen","darkgrey","brown"))

p
```

```{r}
pdf("../Results/dietary_PCA_dendrogram.pdf",height = 10,width = 10)
p
dev.off()
```

**PCA-based dataset**

```{r,fig.width=20,fig.height=10}
dist_mat <- dist(data.pca$x[,1:10], method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)
clusters_dendrogram_B <- cutree(hclust_avg, k = 10)
```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(factor(clusters_dendrogram_B,levels = c(1,2,3,4,5,6,7,8,9,10))) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  scale_color_manual(values=c("red","blue","green","grey","black",
                              "darkred","darkblue","darkgreen","darkgrey","yellow"))

p
```

```{r}
pdf("../Results/dietary_PCA_dendrogramB.pdf",height = 10,width = 10)
p
dev.off()
```

### k-means

Employ k-menas clustering, try different "k" = number of clusters. Analyze this number by two indices: Silhouette Index & Calinski–Harabasz Index

**The Silhouette Index** evaluates how well each data point fits within its cluster compared to other clusters.

For each point i: a(i): average distance to all other points in the same cluster b(i): average distance to all points in the nearest neighboring cluster (the one it is not assigned to)

s(i) = \[b(i) - a(i)\] / \[max{a(i),b(i)}\]

s(i) ranges from -1 to 1: - Near 1: well-matched to its own cluster, poorly matched to others - Near 0: on the boundary between two clusters - Near -1: probably misclassified

The average silhouette score across all points gives an overall measure of clustering quality.

**The Calinski–Harabasz (CH) Index** —also called the Variance Ratio Criterion—measures cluster separation relative to cluster compactness.

CH Index= \[Within cluster variance/(k-1)\] / \[Between cluster variance/(n-k)\]

k = number of clusters; n = number of data points

Higher CH index = better defined and more separated clusters.

It increases when:

-   Cluster centers are far apart

-   Clusters are tight and compac

#### The whole dataset

```{r}
data <- dietary_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans_A <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/dietary_PCA_kmeans.pdf",height = 10,width = 10)
p
dev.off()
```

```{r}
color_variable <- "Country"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

#### PCA-based dataset

```{r}
data <- data.pca$x[,1:10]

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans_B <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)

kmeans_clusters <- metadata_clusters
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/dietary_PCA_kmeansB.pdf",height = 10,width = 10)
p
dev.off()
```

```{r}
color_variable <- "Country"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

### GMM

**Gaussian Mixture Models (GMMs)** are a probabilistic clustering method that assumes your data is generated from a mixture of several Gaussian distributions (i.e. bell curves), each representing a different cluster.

Unlike k-means, which assigns each point to exactly one cluster, GMM assigns probabilities of belonging to each cluster — this is called soft clustering.

**The Bayesian Information Criterion (BIC)** is a model selection metric that helps you choose the best model among a set of candidates by balancing:

Goodness of fit (how well the model explains the data)

Model complexity (number of parameters) -\> the higher the better

#### The whole dataset

```{r}
data <- dietary_df
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)

# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 2

```{r}
data <- dietary_df
gmm_model <- Mclust(data,G = 2,modelNames = "EEI")

# Print summary
summary(gmm_model)

clusters_gmm_A <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_A) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x ) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")

p
```

```{r}
pdf("../Results/dietary_PCA_gmm.pdf",height = 10,width = 10)
p
dev.off()
```

#### PCA-based dataset

```{r}
data <- data.pca$x[,1:10]
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)


# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 3. However as was seen, one cluster has only couple of samples included, so lets stay at 2.

```{r}
data <- data.pca$x[,1:10]
gmm_model <- Mclust(data,G = 2,modelNames = "VEI")

# Print summary
summary(gmm_model)

clusters_gmm_B <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_B) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)

gmm_clusters <- metadata_clusters
```

Lets do color coding of these clusters in the PCA plot.

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>% rownames_to_column("SampleID"),
                metadata_clusters,
                by="SampleID",all=TRUE)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
color_variable <- "Cluster"

p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=!!sym(color_variable))) + 
  xlab(x_lab) + ylab(y_lab)+ 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")

p
```

```{r}
pdf("../Results/dietary_PCA_gmmB.pdf",height = 10,width = 10)
p
dev.off()
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(as.data.frame(data.pca$x),metadata_clusters,
               neighbors = 10,
               min_dist = 0.05)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(as.data.frame(data.pca$x),metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/dietary_UMAP_GMM.pdf",height = 4,width = 4)
p
dev.off()
```

### Dietary clusters export

```{r}
dietary_clusters <- merge(kmeans_clusters,gmm_clusters,by=c("SampleID","Country","Group"),suffixes = c("_kmeans","_gmm"))
```

```{r,eval=FALSE}
write.table(dietary_clusters,"../Results/dietary_clusters.tsv",
            row.names = FALSE,sep="\t")
```

# Metabolomics dataset

## PCA

### SPCA

```{r,fig.width=10,fig.height=4}
spca_plot_df <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

```{r}
pdf("../Results/spca_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

### LCMS

```{r}
data.pca <- prcomp(lcms_df)
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>%
                  rownames_to_column("study_id"),
                metadata_df,by="study_id")

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r,fig.width=10}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

p
```

```{r}
pdf("../Results/lcms_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

We can see clear outliers in the dataset

```{r}
outliers <- df %>% dplyr::filter(PC1< -50) %>% dplyr::select(study_id)
```

Now lets explore if these samples are outliers using just metadata from veganscreener in some way.

```{r}
metadata_df_outliers <- metadata_df %>%
  dplyr::mutate(metabo_outlier=ifelse(metadata_df$study_id %in% outliers$study_id,"YES","NO"))
```

```{r}
data.pca <- prcomp(metadata_df,center = TRUE,scale = TRUE)

library(FactoMineR)
library(factoextra)

# Assuming your data frame is called 'df' and contains only categorical variables
res.mca <- MCA(metadata_df %>% dplyr::select(-study_id), graph = FALSE)

mca_df <- as.data.frame(res.mca$ind$coord)
mca_df$study_id <- metadata_df_outliers$study_id
mca_df$outlier <- metadata_df_outliers$metabo_outlier
```

```{r}
p <- ggplot(mca_df) + 
  geom_point(aes(x=`Dim 1`,y=`Dim 2`,color=outlier)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 
```

TO DO!!!!!!!! preco su outliers?

### NMR

```{r}
data.pca <- prcomp(nmr_df)
```

```{r}
pca_df <- merge(as.data.frame(data.pca$x) %>%
                  rownames_to_column("study_id") ,
                metadata_df,by="study_id")

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")
```

```{r}
p <- ggplot(pca_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right")  + 
  xlab(x_lab) + 
  ylab(y_lab)

p
```

```{r}
pdf("../Results/NMR_PCA_OM_vs_VG.pdf",height = 4,width = 6)
p
dev.off()
```

## UMAP

### SPCA

```{r,fig.height=4,fig.width=10}
p <- umap_plot(spca_integrated_df,metadata_df,
               color_col = "eq_included_group",
               neighbors = 10,
               min_dist = 0.5)


```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(spca_integrated_df,metadata_df,
               neighbors = n_i,
               min_dist = m_i,
               color_col = "eq_included_group") + 
          theme(legend.position = "none")
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### LCMS

```{r,fig.height=4,fig.width=10}
p <- umap_plot(lcms_df,metadata_df,
               color_col = "eq_included_group",
               neighbors = 100,
               min_dist = 0.5)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(lcms_df,kmeans_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

### NMR

```{r,fig.height=4,fig.width=10}
p <- umap_plot(nmr_df,metadata_df,
               color_col = "eq_included_group",
               neighbors = 50,
               min_dist = 0.1)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(nmr_df,metadata_df,
               neighbors = n_i,
               min_dist = m_i,
               color_col = "eq_included_group") + 
          theme(legend.position = "none")
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

For now, we have decided to exlude the outliers from the analysis

```{r}
lcms_df <- lcms_df %>%
  dplyr::filter(!(rownames(lcms_df) %in% outliers$study_id))

spca_integrated_df <- spca_integrated_df %>%
  dplyr::filter(!(rownames(spca_integrated_df) %in% outliers$study_id))
```

## Clustering

### k-means

```{r}
data <- spca_integrated_df

# Range of k values to test
k_range <- 2:10

# Store average scores 
silhouette_scores <- numeric(length(k_range))
ch_scores <- numeric(length(k_range))

# Loop over k values
for (i in seq_along(k_range)) {
  k <- k_range[i]
  
  # K-means clustering
  km_res <- kmeans(data, centers = k, nstart = 10)
  
  # Silhouette analysis
  sil <- silhouette(km_res$cluster, dist(data))
  silhouette_scores[i] <- mean(sil[, 3])
  
   # Calinski–Harabasz score
  ch <- intCriteria(as.matrix(data), as.integer(km_res$cluster), c("Calinski_Harabasz"))
  ch_scores[i] <- ch$calinski_harabasz
}


```

```{r}
# Combine results
results_df <- data.frame(
  k = k_range,
  Silhouette = silhouette_scores,
  Calinski_Harabasz = ch_scores
)

print(results_df)

# Plot both scores
par(mfrow = c(1, 2))
plot(results_df$k, results_df$Silhouette, type = "b", pch = 19,
     xlab = "k", ylab = "Average Silhouette Score", main = "Silhouette Index")
plot(results_df$k, results_df$Calinski_Harabasz, type = "b", pch = 19,
     xlab = "k", ylab = "Calinski–Harabasz Index", main = "CH Index")
```

We will proceed with k=2

```{r}
km_res <- kmeans(data, centers = 2, nstart = 10)
clusters_kmeans <- km_res$cluster

metadata_clusters <- merge(metadata_df,
                     as.data.frame(km_res$cluster) %>%
                       rownames_to_column("study_id") %>%
                       `colnames<-`(c("study_id","Cluster")),
                     by="study_id",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(spca_integrated_df,metadata_clusters,
               neighbors = 50,
               min_dist = 0.3)

p
```

PCA

```{r}
spca_plot_df <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_clusters,by="study_id")


p <- ggplot(spca_plot_df) + 
  geom_point(aes(x=PC1,y=PC2,color=eq_included_group,shape=eq_included_group)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

p
```

### GMM

```{r}
data <- mofa_integrated_df
gmm_model <- Mclust(data)

# Print summary
summary(gmm_model)

# Plot clustering result
plot(gmm_model, what = "classification")  # Shows which sample belongs to which Gaussian

# Plot BIC for different models and number of components
plot(gmm_model, what = "BIC")

# View optimal number of clusters chosen by BIC
best_k <- gmm_model$G
cat("Best number of clusters according to BIC:", best_k, "\n")
```

Lets choose k = 2

```{r}
data <- mofa_integrated_df
gmm_model <- Mclust(data,G = 4,modelNames = "VVI")

# Print summary
summary(gmm_model)

clusters_gmm_A <- gmm_model$classification

```

```{r}
metadata_clusters <- merge(metadata_df,
                     as.data.frame(clusters_gmm_A) %>%
                       rownames_to_column("SampleID") %>%
                       `colnames<-`(c("SampleID","Cluster")),
                     by="SampleID",
                     all.x = TRUE)
```

Lets do color coding of these clusters in the UMAP plot.

```{r,fig.height=3,fig.width=4}
p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = 10,
               min_dist = 0.05)

p
```

```{r, eval=FALSE}
n = c(5, 10, 15, 30, 50, 100)
m = c(0.01, 0.05, 0.1, 0.3, 0.5)
p_list <- list()

for (n_i in n){
    for (m_i in m){
        p <- umap_plot(mofa_integrated_df,metadata_clusters,
               neighbors = n_i,
               min_dist = m_i)
        p_list[[length(p_list)+1]] <- p
    }
}

a <- ggarrange(plotlist = p_list)
```

```{r}
pdf("../Results/metabolomics_UMAP_gmm.pdf",height = 20,width = 20)
p
dev.off()
```

# DIETARY GROUPS

```{r}
dietary_groups_df <- fread("../../Docs/fg_merged_MAINGROUPS_full_mean.tsv") 

dietary_groups_df <- dietary_groups_df %>%
  dplyr::filter(dietary_groups_df$client %in% metadata_df$SampleID)

dietary_groups_df$`MAIN group` <- gsub(" $","",dietary_groups_df$`MAIN group`)
dietary_groups_df$`MAIN group` <- gsub("\t$","",dietary_groups_df$`MAIN group`)

dietary_df_raw <- fread("../../Docs/fg_merged_mean.tsv") %>%
  dplyr::select(client,`Calories(kcal)`) %>%
  `colnames<-`(c("client","all_Calories"))

groups <- unique(dietary_groups_df$`MAIN group`)
pca_plots <- list()
corr_plots <- list()
```

```{r}
for (group in groups){
  
  # subdf
subdf <- dietary_groups_df %>%
  dplyr::filter(`MAIN group`==group) %>%
  dplyr::select(-`MAIN group`)

if (nrow(subdf)!=length(unique(subdf$client))) {
  subdf <- subdf %>%
  group_by(client) %>%
  summarise(across(
    where(is.numeric), 
    mean, na.rm = TRUE
  )) 
}

# info about all calories
subdf <- merge(subdf,dietary_df_raw,by="client",all.x=TRUE)

# preprocessing

## imputation
subdf[subdf==""] <- 0
subdf[is.na(subdf)] <- 0

## normalization
subdf_norm <- subdf %>%
  dplyr::select(-c(client,amount,`Energy(kJ)`,`Calories(kcal)`))
  
subdf_norm <- apply(subdf_norm,2,function(x) x/subdf$`Calories(kcal)`)

if (is.null(nrow(subdf_norm))) next

subdf_norm <- as.data.frame(subdf_norm) %>%
  `rownames<-`(subdf$client) %>%
  rownames_to_column("client")

## log 
subdf_log <- subdf_norm
subdf_log[,-1] <- log(subdf_norm[,-1])

# LETS deal with -Inf -> -10
subdf_log[subdf_log==-Inf] <- -10

## scaling
subdf_scaled <- subdf_log
subdf_scaled[,-1] <- scale(subdf_scaled[,-1], center = TRUE, scale = TRUE)

subdf_scaled <- subdf_scaled %>% column_to_rownames("client")

# remove nans - no variance
subdf_filtered <- subdf_scaled[, colMeans(is.na(subdf_scaled)) < 0.5]
subdf_filtered <- subdf_filtered[rowMeans(is.na(subdf_scaled)) < 0.5,]

if ((ncol(subdf_filtered)<5) | (nrow(subdf_filtered)<5)) next 
data.pca <- prcomp(subdf_filtered)

imp_vec <- data.pca$sdev^2
x_lab = paste("PC1 ", "(",round(imp_vec[1],2),"%", ")", sep="")
y_lab = paste("PC2 ", "(",round(imp_vec[2],2),"%", ")", sep="")

p <- ggplot(as.data.frame(data.pca$x)) + 
  geom_point(aes(x=PC1,y=PC2)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

pca_plots[[length(pca_plots)+1]] <- p + ggtitle(group)

df <- merge(mofa_integrated_df %>% rownames_to_column("SampleID"),
           as.data.frame(data.pca$x) %>% rownames_to_column("SampleID"),
           by="SampleID") %>%
  dplyr::filter(Factor1<10)

a <- ggplot(data=df) + 
  geom_point(aes(x=Factor1,PC1)) + 
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
        axis.ticks.x = element_line(size=0.3,color = "black"),
        axis.ticks.y = element_line(size=0.3,color="black"),
        axis.ticks.length = unit(4,"pt"),
        axis.text = element_text(face = "bold",colour = "black"),
        axis.title = element_text(face = "bold",colour = "black"),
        panel.grid = element_blank()) + 
    theme(legend.position = "right") 

corr_plots[[length(corr_plots)+1]] <- a + ggtitle(group)

}
```

```{r}
a <- ggarrange(plotlist = pca_plots)
b <- ggarrange(plotlist = corr_plots)
```

```{r}
pdf("../Results/pca_perFG.pdf",height = 20,width = 20)
a
dev.off()

```

```{r}
pdf("../Results/corr_perFG.pdf",height = 20,width = 20)
b
dev.off()

```

# SCCA

Sparse Canonical Correlation Analysis (sCCA) is a powerful technique to uncover associations between two multivariate datasets—in your case, between dietary data and metabolomics data. Unlike classical CCA, sCCA performs variable selection, making interpretation easier by identifying only the most relevant features.

What sCCA Does in This Context Input:

X = dietary dataset (e.g., nutrients, amino acids, phytochemicals)

Y = metabolomics dataset (e.g., LC-MS or NMR metabolites)

Output:

Canonical variates: linear combinations of variables in X and Y

Correlated latent structures: highlights which features from each set are most associated

Why sparse?

With many features and few samples, sparsity avoids overfitting and improves interpretability by keeping only a subset of variables in each canonical component.

```{r}
library(sRDA)
```

```{r}
for (group in groups){
  
  # subdf
subdf <- dietary_groups_df %>%
  dplyr::filter(`MAIN group`==group) %>%
  dplyr::select(-`MAIN group`)

if (nrow(subdf)!=length(unique(subdf$client))) {
  subdf <- subdf %>%
  group_by(client) %>%
  summarise(across(
    where(is.numeric), 
    mean, na.rm = TRUE
  )) 
}

# info about all calories
subdf <- merge(subdf,dietary_df_raw,by="client",all.x=TRUE)

# preprocessing

## imputation
subdf[subdf==""] <- 0
subdf[is.na(subdf)] <- 0

## normalization
subdf_norm <- subdf %>%
  dplyr::select(-c(client,amount,`Energy(kJ)`,`Calories(kcal)`))
  
subdf_norm <- apply(subdf_norm,2,function(x) x/subdf$`Calories(kcal)`)

if (is.null(nrow(subdf_norm))) next

subdf_norm <- as.data.frame(subdf_norm) %>%
  `rownames<-`(subdf$client) %>%
  rownames_to_column("client")

## log 
subdf_log <- subdf_norm
subdf_log[,-1] <- log(subdf_norm[,-1])

# LETS deal with -Inf -> -10
subdf_log[subdf_log==-Inf] <- -10

## scaling
subdf_scaled <- subdf_log
subdf_scaled[,-1] <- scale(subdf_scaled[,-1], center = TRUE, scale = TRUE)

subdf_scaled <- subdf_scaled %>% column_to_rownames("client")

# remove nans - no variance
subdf_filtered <- subdf_scaled[, colMeans(is.na(subdf_scaled)) < 0.5]
subdf_filtered <- subdf_filtered[rowMeans(is.na(subdf_scaled)) < 0.5,]

if ((ncol(subdf_filtered)<5) | (nrow(subdf_filtered)<5)) next 
data.pca <- prcomp(subdf_filtered)

data_fg <- as.data.frame(data.pca$x[,1:10])
data_metab <- mofa_integrated_df[,1:10] %>% 
  dplyr::filter(rownames(data_metab) %in% rownames(data_fg))


x <- sCCA(dietary_df, mofa_integrated_df, penalization = "enet")
cca_result <- cc(data_fg, data_metab)
U <- as.data.frame(cca_result$scores$xscores)

# Number of observations
n <- nrow(data_fg)
# Number of variables in each dataset
p <- ncol(data_fg)
q <- ncol(data_metab)
# Calculate p-values for the canonical correlations
p_values <- p.asym(cca_result$cor, n, p, q, tstat = "Wilks")
# View the p-values
p_values

}
```

```{r}
# Load library
library(PMA)

# Assume your data is in two matrices:
#   - diet_matrix: rows = individuals, columns = dietary variables
#   - metabo_matrix: rows = individuals, columns = metabolite intensities

# Step 1: Scale the data (important for CCA)
diet_scaled <- dietary_df
metabo_scaled <- mofa_integrated_df

# Step 2: Tune penalty parameters (sparsity levels)
perm.out <- CCA.permute(x = diet_scaled, z = metabo_scaled, typex = "standard", typez = "standard")

# Step 3: Run sCCA with chosen penalties
cca.out <- CCA(x = diet_scaled, z = metabo_scaled, typex = "standard", typez = "standard",
               penaltyx = perm.out$bestpenaltyx,
               penaltyz = perm.out$bestpenaltyz,
               K = 1)  # K = number of canonical components

# Step 4: View results
# Canonical correlation
cca.out$cor

# Variable loadings (non-zero values indicate selected variables)
selected_diet_vars <- which(cca.out$u != 0)
selected_metabo_vars <- which(cca.out$v != 0)

diet_loadings <- cca.out$u[selected_diet_vars]
metabo_loadings <- cca.out$v[selected_metabo_vars]

# Display selected variables and loadings
cat("Selected dietary variables:\n")
print(colnames(diet_scaled)[selected_diet_vars])
print(diet_loadings)

cat("\nSelected metabolites:\n")
print(colnames(metabo_scaled)[selected_metabo_vars])
print(metabo_loadings)

# Optional: Biplots
plot(cca.out$u, cca.out$v, main = "sCCA Loadings", xlab = "Diet", ylab = "Metabolites")
abline(h=0, v=0, lty=2)

```

```{r}
a <- data.frame(MOFA1=mofa_integrated_df$Factor1,
                  DIET1=data.pca$x[,"PC1"]) %>%
  dplyr::filter(MOFA1<10)

ggplot(data=a) + 
  geom_point(aes(x=MOFA1,DIET1))
```
