---
title: "VEGANScreener + metabolomics"
subtitle: "Q2: Associations between VS_total and the metabolome in vegans."
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code is used to explore patterns in the metabolome profiles of vegans, particularly by linking them to the VEGANScreener score (VS_total).**


The script is divided into two parts, according to the integration method of metabolome datasets (see preprocessing scripts for details). At first, the **sPCA dataset** is analyzed. This dataset is used for multivariate analysis, PCA visualization, PERMANOVA and ML approach. The **concatenated dataset** of both LCMS and NMR is used for univariate analysis. 

The main parts of the analysis is composed of:

1.  Global visualization - 3 components color-coding **VS_total**

2.  PERMANOVA - how much of the variation can be explained by various external variables:
  dist(Metabolome) \~ sex + bmi + age + veganDuration + country***VS_total**

3.  Univariate analysis (feature-wise model) - linear modelling of metabolite using various external variables:
  Metabolite \~  sex + bmi + age + veganDuration + Country***VS_total**
  
4.  Random fores (RF) classification - the model is trained and validated through bootstrapping with 500 iterations. The RF model is used to distinguish between participants with **1st and 3rd tercile of VS_total**. The two types of models are trained:
a) ML model on original sPCA dataset
b) ML model on sPCA dataset after regressing out the covariates.

```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  file =  "../../data/preprocessed/metadata_preprocessed_v3.tsv",
  header = TRUE)  

lcms_df <- read.table(
  file= "../../data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",
  check.names = FALSE)

nmr_df <- read.table(
  file = "../../data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",
  check.names = FALSE)

spca_integrated_df <-  read.table(
  file = "../../data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",
  check.names = FALSE)

```

**Scaling** - when using columns from metadata as covariates, they need to be scaled, so mean is 0 and SD is 0.

```{r}
metadata_scaled_df <- metadata_df
metadata_scaled_df[,-c(1,3,5,6,8)] <- apply(metadata_df[,-c(1,3,5,6,8)],2,scale)
metadata_scaled_df <- metadata_scaled_df %>%
  mutate(sex=as.factor(sex),
         country=as.factor(country),
         group=as.factor(group),
         diet_supplements=as.factor(diet_supplements))
```

# Descriptive statistics

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=15,fig.height=10}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total")
i <- 1
for (col in c("age","bmi","vegan_duration","VS_total")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country")
for (col in c("sex","country")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20)) +
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count() %>%
              dplyr::mutate(n=n+10), aes(x=!!sym(col),y=n,label=n)) + 
    xlab(name[i]) 
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```

# SPCA

The SPCA dataset was created by concatenating LCMS and NMR dataset and applying Mixomics::spca() function. See preprocessing script for details. 

mixOmics::spca() performs **Sparse Principal Component Analysis**, which is similar to  regular PCA but adds a sparsity constraint so that only a subset of variables contributes to each component.
This sparsity is controlled by penalization, making interpretation easier because irrelevant or noisy variables get loadings of zero.

## Visualization


```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/umap.pdf",height = 4,width = 6)
p
dev.off()
```

### PCA

```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE)

p <- ggtheme(p)

p

p_pca <- list()
p_pca[[1]] <- p
```

```{r}
f2a <- p
save(f2a,file="../results/vegan/figures/f2a.RData")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/PCA.pdf",height = 4,width = 6)
p
dev.off()
```

```{r,fig.width=15,fig.height=4}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p1 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC1","PC2"))

p1 <- ggtheme(p1)

p2 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC2","PC3"))

p2 <- ggtheme(p2)

p3 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC1","PC3"))

p3 <- ggtheme(p3)

p1 + p2 + p3 + 
plot_layout(guides = "collect")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/PCA_all.pdf",height = 4,width = 15)
p1 + p2 + p3 + 
plot_layout(guides = "collect")
dev.off()
```



### Boxplots

```{r,fig.height=15,fig.width=20}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_scaled_df,by="study_id",all = TRUE) 

spca_for_uni <- spca_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  )

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = "tertile_label",y = col,fill="tertile_label")) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) +
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p)
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_boxplots.pdf",height = 15,width = 20)
a
dev.off()
```


### Heatmap

```{r}
metadata_df <- metadata_df %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  ) %>%
  dplyr::select(-tertile_cut)
```

```{r}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

p_df <- spca_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,tertile_label) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "tertile_label"=VS_colors
)
names(ann_colors[["tertile_label"]]) <- names(VS_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```

**Tertile averages**

```{r}
metadata_df %>%
  dplyr::select(VS_total, tertile_label) %>%
  group_by(tertile_label) %>%
  summarise(mean_VS = mean(VS_total, na.rm = TRUE))
```

## PERMANOVA

The test show how much of the variation in metabolomics is explained by: sex - age - bmi -vegan duration - Country - **VS_total**

Variables were added sequentially (Type I SS), so each p-value and R² reflects a partial effect, after adjusting for those above. This model accounts for variation due to sex, age, vegan diet duration, BMI and country of origin. Then accounts for the main effects of VS_total. Then tests whether the effect of VS_total varies by country of origin (interaction).

----------------------

VS_total explains **2.588%** of variation after adjusting for sex, age, bmi,country and duration of the diet. This result is statistically significant (p<0.001).

However, the country of origin is still explaining the most of the variation ~ **10%**. The interaction effect is not significant, which suggest the differences in VS_total is not inconsistent throughout the countries.

```{r}
# Create Euclidean distance matrix
dist_mat <- vegdist(spca_integrated_df, 
                    method = "euclidean")

# Test whether vegan vs omnivore differ in metabolomics
adonis_result <- adonis2(
  dist_mat ~ sex + bmi + age + vegan_duration + country*VS_total,
  data = metadata_scaled_df , 
  permutations = 999,
  by="terms")

print(adonis_result)
```

```{r,fig.width=6,fig.height=2}
adonis_result <- adonis_result[-c(7,9),] %>% rownames_to_column("Var")
adonis_result$Var <- factor(adonis_result$Var,levels = rev(c("sex","bmi","age", "vegan_duration","country","VS_total", "Residual")))
p <- ggplot(adonis_result, aes(fill=Var, x=R2, y="Metabolome")) + 
    geom_bar(position="stack", stat="identity") + 
  ylab("") + xlab("Partial variance explained (R²)") + 
  scale_fill_manual(values=rev(c("#ffffcc","#b3cde3","#ccebc5",
                                 "#decbe4","#fed9a6","darkred",
                                 "lightgrey")),name = "Covariate")  

p <- ggtheme(p)

p

p_pca[[2]] <- p
p_pca[[3]] <- adonis_result
```

## RF model

### Original model

```{r}
sub_metadata_df <- metadata_df %>%
dplyr::filter(
  tertile_label %in% c("1st_VStertile","3rd_VStertile"))

sub_spca_integrated_df <- spca_integrated_df[as.character(sub_metadata_df$study_id),]
```

```{r,fig.height=5,fig.width=5}
model_name <- "VS_tertile_vegan" 

prepared_df <- model_preparation(sub_spca_integrated_df,
                                 sub_metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                            outcome="Group",
                            N = 500,
                           file = model_name,
                            reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("low VS_total","high VS_total"),CI = TRUE) 

roc_c <- ggtheme(roc_c) + 
  theme(legend.position="none",title = element_text(size=8))

roc_c
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/orig_rf_vstotal.pdf",width = 5,height = 5)
roc_c
dev.off()
```

**Accuracy**

```{r}
# Find the "best" threshold using Youden's J Index
# This maximizes (Sensitivity + Specificity - 1)
roc_obj <- rf_model$rocobj
best_threshold_data <- coords(roc_obj, "best", 
                              ret = "threshold", 
                              best.method = "youden")
optimal_threshold <- best_threshold_data$threshold
optimal_threshold
```

```{r}
# saving the results
save_rf_details("../results/vegan/RF_models_performance.xlsx",model_name,rf_model)
```

### Adjusted model

```{r,fig.height=5,fig.width=5}
# Covariates to adjust for
covars <- c("sex","bmi", "age", "country", "vegan_duration")
integrated_df <- sub_spca_integrated_df
all(rownames(integrated_df) == sub_metadata_df$study_id)
adjusted_data <- adjust_for_covars(integrated_df, covars, sub_metadata_df)

model_name <- "adjusted_vstotal_model_vegan" 

prepared_df <- model_preparation(adjusted_data,
                                 sub_metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 500,
                          file = model_name,
                          reuse=TRUE)

rf_model$model_summary

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("low VS_total","high VS_total"),CI = TRUE) 

roc_c <- ggtheme(roc_c) + 
  theme(legend.position="none",title = element_text(size=8))

roc_c


```

```{r}
f2b <- roc_c
save(f2b,file="../results/vegan/figures/f2b.RData")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/adjusted_rf_vstotal.pdf",width = 5,height = 5)
roc_c
dev.off()
```

```{r}
# Find the "best" threshold using Youden's J Index
# This maximizes (Sensitivity + Specificity - 1)
roc_obj <- rf_model$rocobj
best_threshold_data <- coords(roc_obj, "best", 
                              ret = "threshold", 
                              best.method = "youden")
optimal_threshold <- best_threshold_data$threshold
optimal_threshold
```

```{r}
# saving the results
save_rf_details("../results/vegan/RF_models_performance.xlsx",model_name,rf_model)
```

## RF importance

### Original model

**PC5** and **PC10** are the most important for RF model.

```{r,fig.width=10}
load("../intermediate_files/models/VS_tertile_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Adjusted model

Similarly as above, the most important is **PC5**, with significant drop of importance after this PC. 

```{r,fig.width=10}
load("../intermediate_files/models/adjusted_vstotal_model_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Visualization

To explore the important PCs, their visualization is shown below. We will show **PC5**, as it was consistently picked by two models and **PC10** as this PC remained in the list of the most important PCs after adjustment. 


```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,
              axis = c("PC5","PC10"))

p <- ggtheme(p)

p

p_pca[[4]] <- p
```

```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,
              axis = c("PC5","PC12"))

p <- ggtheme(p)

p

```

```{r}
p <- ggplot(spca_for_uni, 
             aes_string(x = "tertile_label",y = "PC5",fill="tertile_label")) +
    geom_boxplot() +
    ggtitle("PC5") + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
p <- ggtheme(p) 

p

```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/important_pca_boxplots.pdf",height = 4,width = 8)
p
dev.off()
```


## SPCA loadings {#SPCA_loadings}

Here, we will explore the loadings of the most important PC (**PC5**)

```{r,collapse=TRUE}
load("../results/preprocessing/spca_result_vegans_INT_v2.RData")
df <- data.frame (
  Metabolite=names(spca_result$loadings$X[,5]),
  Loading=spca_result$loadings$X[,5])

write.table(df,sep="\t","../results/vegan/Q2_VStotal/loadings_PC5.tsv")
```

This plot shows the loading of individual metabolites to **PC5**, ordered by its absolute value.

```{r,fig.height=10,fig.width=5,collapse=TRUE}
ggplot(df %>% 
         dplyr::mutate(
           Metabolite=factor(Metabolite,
                             levels = Metabolite[order(abs(Loading),decreasing = FALSE)])
           )) + 
  geom_col(aes(x=Metabolite,y=Loading)) + 
  theme(axis.text.x = element_text(angle=90,hjust = 1)) + 
  coord_flip()

spca_loadings <- df$Metabolite[abs(df$Loading)>=0.1]
```

# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```

## Visualization

### Boxplots

```{r,fig.height=40,fig.width=20}
concat_for_uni <- merge(concat_integrated_df %>% rownames_to_column("study_id"),
                      metadata_scaled_df,by="study_id",all = TRUE) 

concat_for_uni <- concat_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  )

plot_list <- list()
for (col in colnames(concat_integrated_df)){
  p <- ggplot(concat_for_uni, 
              aes(x = tertile_label,
                  y = !!sym(col),
                  fill=tertile_label)) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) +
    geom_signif(comparisons = 
                  combn(as.character(unique(concat_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p) + 
    theme(axis.text.x = element_text(angle=45,hjust = 1))
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_boxplots.pdf",height = 35,width = 20)
a
dev.off()
```

### Heatmap

```{r}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

p_df <- concat_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,tertile_label) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "tertile_label"=VS_colors
)
names(ann_colors[["tertile_label"]]) <- names(VS_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              show_rownames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```


## Univariate analysis

We will test the differences between diets and VS total using generalize linear model while adjusting for covariates.

### Metabolite-wise model

#### Continuous analysis

Call glm to explore which metabolites are significantly related to VS score while adjusting for age, sex, Country and dietary group:

Metabolite ~ sex + bmi + age + eq_vegan_dur_total + Country*VS_total

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_glm(x,concat_for_uni,
                                           tested_variable = "VS_total"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results - **9 metabolites**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total")

```

```{r}
results_all <- (results_df %>%
  dplyr::filter(variable=="VS_total"))

# save 
write.table(results_all,"../results/vegan/Q2_VStotal//lm_continous_all.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")
```

Check the interaction effect - **none**

```{r}
df <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(str_ends(variable, ":VS_total"))

df

pc_to_be_removed <- df$metabolite
```

```{r,collapse=TRUE}
concat_significant_df <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))

concat_univariate_results <- results_df

concat_significant <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))$metabolite

# save and show
write.table(concat_significant_df,"../results/vegan/Q2_VStotal//lm_continous.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")

knitr::kable(concat_significant)
```

**Volcano plot**

```{r}
vp <- volcano_plot(results_df,comparison = "VS_total")
vp
```

**Correlation**

```{r}
sign <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))$metabolite

```

```{r,fig.height=5,fig.width=5}
df_for_plot <- concat_integrated_df %>% dplyr::select(sign)
corr_object <- cor(concat_for_uni$VS_total,df_for_plot)
corrplot::corrplot(corr_object,method = "number")
```

**Vegan duration**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="vegan_duration")

```

### Pairwise analysis

Metab ~ sex + bmi + age + eq_vegan_dur_total + Country + tertile_label
 
**LM**

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "tertile_label"))
results_df <- do.call(rbind, results_list)

vs_effects <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

concat_significant_cat <-  vs_effects %>%
  filter(p_adj < 0.05)

# save and show
write.table(concat_significant_cat,"../results/vegan/Q2_VStotal//lm_cat.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")

knitr::kable(concat_significant_cat)
```

```{r}
results_all <- vs_effects

# save 
write.table(results_all,"../results/vegan/Q2_VStotal/lm_emmeans_all.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")
```

**Dispersion**

```{r}
dispersion <- vs_effects %>%
  group_by(metabolite) %>%
  summarise(
    range      = max(emmean) - min(emmean),      # simplest: spread between highest and lowest 
    sd         = sd(emmean),                      # standard deviation of the 5 country means
    cv         = sd(emmean) / abs(mean(emmean)),  # coefficient of variation (relative dispersion)
    min_mean   = min(emmean),
    max_mean   = max(emmean),
    min_VS = tertile_label[which.min(emmean)],
    max_VS = tertile_label[which.max(emmean)]
  ) %>%
  arrange(desc(range))

dispersion

write.table(dispersion,"../results/vegan/Q2_VStotal/dispersion.tsv",sep="\t",row.names = FALSE,quote=FALSE)
```

```{r}
results_list_pairwise <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm_pairwise(x,concat_for_uni,
                                           tested_variable = "tertile_label"))
results_df_pairwise <- do.call(rbind, results_list_pairwise)

vs_effects_pairwise <- results_df_pairwise %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

vs_significant_pairwise <-  vs_effects_pairwise %>%
  filter(p_adj < 0.05)

```

#### Heatmap

```{r}
df2 <- vs_effects_pairwise %>%
  mutate(
    contrast = gsub("_VStertile", "", contrast),  
    signif = p_adj < 0.05
  )

df2$Significance <- ifelse(df2$p_adj<0.001,"***",
                                       ifelse(df2$p_adj<0.01,"**",
                                              ifelse(df2$p_adj<0.05,"*","")))

```

```{r,fig.height=8,fig.width=5}
p_pairwise_heatmap <- ggplot(df2, aes(x = contrast, y = Metabolite, fill = estimate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance,vjust=0.6), color = "black", size = 2) +
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0) + 
  theme_minimal() +
  theme(axis.text.y = element_text(size=4),
        axis.text.x = element_text(angle = 45,face="bold",hjust = 1),
        panel.grid = element_blank()) + 
  xlab("") + ylab("")

p_pairwise_heatmap
```

Network construction:

#### Network construction

**Edge list**

```{r}
edges <- vs_effects %>%
  filter(p_adj < 0.05) %>%
  #filter(abs(emmean)>0.3) %>%
  dplyr::select(from = tertile_label, to = metabolite, 
         emmean = emmean) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(emmean))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
tertiles <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(tertiles, metabs),
  type = c(rep(TRUE, length(tertiles)), rep(FALSE, length(metabs))) # TRUE = tertile, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r,fig.height=6,fig.width=6}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     VS_colors[V(g)$name],
                     "gray70")

set.seed(123)
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue", "lightpink"), labels = c("<0", ">0")) +
  geom_node_point(aes(shape=type,color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +  # node colors
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  guides(color = "none")

```

#### Degree calculation


```{r}
# Calculate degree
deg <- degree(g, mode = "all")

# Add to node dataframe
node_degree_df <- data.frame(
  name = names(deg),
  degree = deg,
  type = V(g)$type
)

# Separate interpretation
vs_nodes <- node_degree_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_degree_df %>% dplyr::filter(type == FALSE)

vs_nodes %>% arrange(desc(degree))
metabolite_nodes %>% arrange(desc(degree))
```

**Weighted degree**
- Should reflect cumulative association magnitude, not just number of links.

```{r}
strength_vals <- strength(g, weights = E(g)$weight)

node_strength_df <- data.frame(
  name = names(strength_vals),
  strength = strength_vals,
  type = V(g)$type
)

# Separate interpretation
vs_nodes <- node_strength_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_strength_df %>% dplyr::filter(type == FALSE)

vs_nodes %>% arrange(desc(strength))
metabolite_nodes %>% arrange(desc(strength))
```

#### Forest plot

```{r, fig.height=15,fig.width=5}
# Define custom colours
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

vs_effects <- merge(vs_effects,dispersion[,c("metabolite","range"),all.x=TRUE,by="metabolite"])

# Optional: order metabolites by mean effect size
forest_df <- vs_effects %>%
  group_by(metabolite) %>%
  mutate(range_eff = mean(range)) %>%
  ungroup()

forest_df$metabolite <- factor(
  forest_df$metabolite,
  levels = forest_df %>% 
    arrange(range_eff) %>% 
    pull(metabolite) %>% 
    unique()
)

# Forest plot WITHOUT ANY TEXT LABELS
forest_df <- forest_df %>% mutate(sig = p_adj < 0.05)
p_forest <- ggplot(forest_df, 
                   aes(x = emmean, 
                       y = metabolite, 
                       color = tertile_label)) +
  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2.5,aes(shape = sig)) +
  scale_color_manual(values = VS_colors) +
  scale_shape_manual(values = c(`FALSE` = 20, `TRUE` = 19),   # 1 = open, 19 = solid
                     labels = c("ns", "FDR < 0.05"),
                     name   = "Significance") + 
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6),
        panel.grid.minor = element_blank()) +

  labs(
    x = "Estimated mean (with 95% CI)",
    color = "VS_total"
  )

p_forest <- fix_labels(p_forest)

p_forest

```

```{r}
f <- p_forest
save(f,file="../results/vegan/figures/FS_VSForest.RData")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/final_significance_forest.pdf",height = 15,width = 6)
p_forest
dev.off()
```

#### Top-group assignment

**All together**

```{r}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850",
               "other"="gray70")

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     VS_colors[V(g)$name],
                     "gray70")

top_group <- edges %>%
  dplyr::group_by(to) %>%
  slice_max(order_by = abs(weight), n = 1, with_ties = FALSE) %>%
  dplyr::select(from,to, weight) %>%
  ungroup()

top_group

top_group_viz <- rbind(top_group, 
                       data.frame(
                         from=unique(edges$from),
                         to=unique(edges$from),
                         weight=NA))

membership_df <- data.frame(
  vertex = top_group_viz$to,
  cluster = top_group_viz$from) 

membership_df <- membership_df %>% arrange(match(vertex,names(V(g))))

```

```{r}
vs_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:3,]

(membership_df %>%
  group_by(cluster) %>%
  summarise(Number_of_metabolites = length(unique(vertex))-1))[1:3,]

vs_clusters
```

```{r}
V(g)$fill <- VS_colors[membership_df$cluster]
V(g)$color <- VS_colors[membership_df$cluster]
```

```{r,fig.height=6,fig.width=6}
set.seed(123)
p_all <- ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue","lightpink"),labels = c("<0", ">0")) +
  geom_node_point(aes(size=type,color = name,fill=name),shape=21,stroke=1.2) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name))   + # node colors
  scale_fill_manual(values = setNames(V(g)$fill, V(g)$name)) +  # node fill
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() + 
  theme(legend.position = "none")


p_all
```


# Final significant metabolites

```{r,fig.height=10,fig.width=15}
list_core <- list()

list_core[["concat_univariate"]] <- unique(concat_significant)

```

::: {.output-background style="background-color: lightgreen; padding: 10px;"}

```{r,echo=TRUE}
final_significance <- list_core[["concat_univariate"]]
final_significance
```

:::

# Metabolite-based veganscreener score

## Continuous univariate test

To create a vegan index based on concatenated dataset, we will use linear regression in order to get coefficients on significant variables that were revealed across the analysis:

```{r}
sign <- final_significance
sign
```


```{r}
# define the variables
plot_list <- list()
rhos <- rep(NA,5)

# split the datasets for cross-validation
concat_for_index <- concat_for_uni[sample(rownames(concat_for_uni)),]
cv_splits <- cv_folds(concat_for_index,folds = 5)

# define formula - significant metabolites
vars <- sign
# Wrap all variable names in backticks
vars_safe <- paste0("`", vars, "`")

# Create the formula
fml <- as.formula(paste("VS_total ~", paste(vars_safe, collapse = " + ")))

final_df_veganindex <- data.frame()

# cross validation loop
for (i in 1:length(cv_splits)){
  # fit the model
  model <- lm(fml, data = concat_for_uni[cv_splits[[i]]$train,])
  # predict the result on validation dataset
  VG_index <- stats::predict(model,concat_for_uni[cv_splits[[i]]$val,])
  
  # store the predicted index
  sub_df <- concat_for_uni[cv_splits[[i]]$val,]
  sub_df$VGindex <- VG_index
  final_df_veganindex <- rbind(final_df_veganindex,sub_df)
  
  # calculate correlation 
  cor_test <- cor.test(sub_df$VGindex,
               sub_df$VS_total,
               method = "spearman")
  cor_val <- cor_test$estimate
  if (cor_test$p.value >0.05) cor_val = 0

  # plot the result
  p <- ggplot(sub_df, 
              aes_string(x = "VGindex", y = "VS_total"),
              color="blue") +
       geom_point() + 
    geom_smooth(method = "lm") + 
    theme_minimal() + 
    ggtitle(paste("Rho =",round(cor_val,2)))

  plot_list[[i]] <- p
  rhos[i] <- round(cor_val,2)
  
}

```

```{r,warning=FALSE,fig.height=5,fig.width=6}
wrap_plots(plot_list)
```

```{r}
mean(rhos)
```

```{r}
# calculate correlation 
  cor_test <- cor.test(final_df_veganindex$VGindex,
               final_df_veganindex$VS_total,
               method = "spearman")
  cor_val <- cor_test$estimate
  if (cor_test$p.value >0.05) cor_val = 0

  # plot the result
  p <- ggplot(final_df_veganindex, 
              aes_string(x = "VGindex", y = "VS_total"),
              color="blue") +
       geom_point() + 
    geom_smooth(method = "lm") + 
    theme_minimal() + 
      theme(panel.border = 
              element_rect(color = "black", fill = NA, size = 0),
            panel.grid = element_blank(),
            axis.ticks.x = 
              element_line(size=0.3,color = "black"),
            axis.ticks.y = 
              element_line(size=0.3,color="black"),
            axis.ticks.length = unit(4,"pt"),
            axis.text=element_text(face="bold",size=12),
            axis.title = element_text(face="bold",size=15)) + 
    ggtitle(paste("Rho =",round(cor_val,2))) + 
    xlab("Metabolite-based Vegan score") + 
    ylab("Vegan score (z-scored)") 
  
p 
```

```{r,fig.height=4,fig.width=5}
p
```

```{r}
f <- p + ggtitle("")
save(f,file="../results/vegan/figures/F5b_VMI.RData")
```

# Plot for publication

## Heatmap

### Continuous

```{r,fig.height=10,fig.width=15}
ordered_participants <- metadata_df %>%
  dplyr::arrange(VS_total)

concat_significant_df <- concat_significant_df %>%
  dplyr::filter(metabolite %in% final_significance) %>%
  dplyr::arrange(Estimate)

heatmap_concat_integrated_df <- concat_integrated_df

heatmap_concat_integrated_df <- heatmap_concat_integrated_df[as.character(ordered_participants$study_id),
                             as.character(concat_significant_df$metabolite)]

# Create a grouping variable for each block of 10 rows
group_id <- rep(1:ceiling(nrow(heatmap_concat_integrated_df) / 10), each = 10, 
                length.out = nrow(heatmap_concat_integrated_df))

# Calculate the mean for each group
df_grouped <- aggregate(heatmap_concat_integrated_df, 
                        by = list(Group = group_id), FUN = median)

# Remove the Group column name prefix if you want just the values
rownames(df_grouped) <- paste0("Group_", df_grouped$Group)
df_grouped$Group <- NULL

```


```{r}
df_long <- df_grouped %>% t() %>% as.data.frame() %>%
  rownames_to_column("Metabolite") %>%
  pivot_longer(cols = -Metabolite, names_to = "Participant", values_to = "value")

df_long$Metabolite <- factor(df_long$Metabolite, levels=concat_significant_df$metabolite)
df_long$Participant <- factor(as.character(df_long$Participant), levels=paste0("Group_",1:37))


p_heat <- ggplot(df_long, aes(x = Participant, y = Metabolite, fill = value)) +
  geom_tile(color = "white",width=1) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_text(face = "bold",size = 9),
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") + 
  labs(fill = "Median metabolite z-score") +
  xlab("Vegans (ordered and grouped by Vegan score)") + 
  ylab("") + 
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0)

# get legend
leg_heat_p <- ggpubr::get_legend(p_heat)

# remove legend
p_heat <- p_heat + theme(legend.position = "right")
```


```{r}
annotation_df <- concat_significant_df
annotation_df$metabolite <- factor(annotation_df$metabolite,
                                   levels=annotation_df$metabolite)


ann_p <- ggplot(
  data=annotation_df,
  aes(x = "Estimate", y = metabolite, fill = Estimate)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "#313695", mid = "white", high = "#A50026", midpoint = 0) +
  theme_minimal() + 
  theme(axis.text.y = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.y= element_text(face = "bold"),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") + 
  ylab("Significant metabolites") + 
  xlab("") + 
  labs(fill="LM estimate")

ann_p <- fix_labels(ann_p)

# get legend
leg_ann_p <- ggpubr::get_legend(ann_p)

# Remove legend from the original plot
ann_p <-ann_p + theme(legend.position = "right")

```

```{r}
# correlations
df_for_plot <- concat_integrated_df %>% dplyr::select(final_significance)
df_for_plot <- df_for_plot[,concat_significant_df$metabolite]
corr_object <- cor(concat_for_uni$VS_total,df_for_plot,method="spearman")
corr_df <- data.frame(Metabolite=colnames(corr_object),
                      Rho=corr_object[1,])
corr_df$Metabolite <- factor(corr_df$Metabolite,
                             levels=corr_df$Metabolite)

p_corr <- ggplot(corr_df) + 
  geom_point(aes(x="Rho",y= Metabolite,size = abs(Rho), fill = Rho, color=Rho), 
             shape = 21) +
  scale_fill_gradient2(name = "Rho", 
                       low = "blue", mid = "white", high = "red", midpoint = 0) +
    scale_color_gradient2(name = "Rho", 
                          low = "blue", mid = "white", 
                          high = "red", midpoint = 0) + 
  guides(size = "none") + 
  theme_minimal() + 
  theme(axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y= element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "right") + 
  xlab("") 


leg_corr_p <- ggpubr::get_legend(p_corr)

# Remove legend from the original plot
p_corr <-
  p_corr +
  theme(
    legend.position = "right")
```

```{r}
legends <- ggarrange(leg_heat_p, leg_ann_p, leg_corr_p,nrow = 3, align = "hv")

p <- wrap_plots(ann_p,p_corr,p_heat,legends,ncol=4,widths = c(0.1,0.15, 1, 1))
```


```{r}
cbar <- guide_colourbar(barheight = unit(1.8, "cm"),
                        barwidth  = unit(0.35, "cm"),
                        title.position = "top")

p_heat <- p_heat + guides(fill   = cbar)   # z-score (fill)
ann_p  <- ann_p  + guides(fill   = cbar)   # LM estimate (fill)
p_corr <- p_corr + guides(fill = cbar)   # Rho (colour, not fill)

p <- wrap_plots(ann_p, p_corr, p_heat, ncol = 3, widths = c(0.1, 0.15, 1)) +
  plot_layout(guides = "collect") &
  theme(legend.position  = "right",
        legend.box       = "vertical",
        legend.title     = element_text(size = 9, face = "bold"),
        legend.text      = element_text(size = 8))
```


```{r,fig.height=5,fig.width=20}
p
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/final_significance.pdf",height = 5,width = 7)
p
dev.off()
```


```{r}
f <- p
save(f,file="../results/vegan/figures/FS_VSheatmap.RData")
```

### Pairwise

```{r,fig.height=8,fig.width=5}
p_pairwise_heatmap
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal//final_significance_pairwise.pdf",height = 8,width = 5)
p_pairwise_heatmap
dev.off()
```

## Forest plot

```{r, fig.height=15,fig.width=5}
p_forest
```

```{r}
save(p_forest,file = "../results/vegan/figures/FS2.RData")
```

## Network

```{r,fig.width=6,fig.height=6}
p_all
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal//final_significance_network.pdf",height = 6,width = 6)
p_all
dev.off()
```
