---
title: "VEGANScreener + metabolomics"
subtitle: "Q1h: Helper script for Countries ~ PDI groups associations"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  file =  "../../data/preprocessed/metadata_preprocessed_v3.tsv",header = TRUE)

lcms_df <- read.table(
  file= "../../data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",
  check.names = FALSE)

nmr_df <- read.table(
  file = "../../data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",
  check.names = FALSE)

spca_integrated_df <-  read.table(
  file = "../../data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",
  check.names = FALSE)

pdi_df <- as.data.frame(fread(
  file = "../../data/preprocessed/dietary/pdi/pdi_avg_residuals_scaled_v3.tsv")) %>%
  dplyr::select(-c(age,gender,bmi,eq_included_group,study_center,PDI_sweets_and_desserts_alt_avg))

```

**We will remove the animal-related groups.**

```{r}
pdi_df <- pdi_df %>% dplyr::select(-c(PDI_animal_fat_avg,PDI_eggs_avg,PDI_fish_or_seafood_avg,PDI_meat_avg,PDI_dairy_avg))
```

**Scaling** - when using columns from metadata as covariates, they need to be scaled, so mean is 0 and SD is 0.

```{r}
metadata_scaled_df <- metadata_df
metadata_scaled_df[,-c(1,3,5,6,8)] <- apply(metadata_df[,-c(1,3,5,6,8)],2,scale)
metadata_scaled_df <- metadata_scaled_df %>%
  mutate(sex=as.factor(sex),
         country=as.factor(country),
         group=as.factor(group),
         diet_supplements=as.factor(diet_supplements))
```

# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```


```{r}
concat_for_uni <- merge(concat_integrated_df %>% rownames_to_column("study_id"),
                      metadata_scaled_df,by="study_id",all = TRUE) 
concat_for_uni <- merge(concat_for_uni,pdi_df,by="study_id",all = TRUE)
```

## Visualization

```{r,fig.height=40,fig.width=20}
pdi_for_uni <- merge(pdi_df,
                      metadata_scaled_df,by="study_id",all = TRUE) 

plot_list <- list()
for (col in colnames(pdi_for_uni)[-1]){
  p <- ggplot(pdi_for_uni, 
              aes(x = country,y = !!sym(col),fill=country)) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values=c(
      BE = "#FDDA24",   # gold
      CZ = "#11457E",   # blue
      DE = "#000000",   # black
      ES = "#630356",   # purple
      CH = "#D52B1E"    # red
    )) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(concat_for_uni$country)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p)
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q1_country/concat_boxplots_pdi.pdf",height = 20,width = 20)
a
dev.off()
```

## Univariate analysis

The lm model is used to test the differences in countries, while adjusting for covariates:

*PDI group* ~ sex + bmi + age + VS_total + vegan_duration + Country
    

### LM

**LM**

```{r}
results_list <- lapply(colnames(pdi_df)[-c(1,14:20)], 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "country"))
results_df <- do.call(rbind, results_list)

country_effects <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH")) %>%
  dplyr::mutate(PDI_group=metabolite)

concat_significant <-  country_effects %>%
  filter(p_adj < 0.05)
```

```{r}
results_list_pairwise <- lapply(colnames(pdi_df)[-c(1,14:20)], 
                       function(x) fit_netglm_pairwise(x,concat_for_uni,
                                           tested_variable = "country"))
results_df_pairwise <- do.call(rbind, results_list_pairwise)

country_effects_pairwise <- results_df_pairwise %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH")) %>%
  dplyr::mutate(PDI_group=Metabolite)

concat_significant_pairwise <-  country_effects_pairwise %>%
  filter(p_adj < 0.05)
```

### Heatmap

```{r}
df2 <- country_effects_pairwise %>%
  mutate(
    contrast = gsub(" ", "", contrast),   # remove spaces, e.g. "BE-DE"
    signif = p_adj < 0.05
  )

df2$Significance <- ifelse(df2$p_adj<0.001,"***",
                                       ifelse(df2$p_adj<0.01,"**",
                                              ifelse(df2$p_adj<0.05,"*","")))

```

```{r,fig.height=4,fig.width=5}
p_pairwise_heatmap <- ggplot(df2, aes(x = contrast, y = PDI_group, fill = estimate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance,vjust=0.6), color = "black", size = 2) +
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0) + 
  theme_minimal() +
  theme(axis.text.y = element_text(size=4),
        axis.text.x = element_text(angle = 45,face="bold",hjust = 1),
        panel.grid = element_blank()) + 
  xlab("") + ylab("")

p_pairwise_heatmap
```

### Network

#### All edges (>0.3)

**Edge list**

```{r}
edges <- country_effects %>%
  filter(abs(emmean) > 0.3) %>%
  dplyr::select(from = country, to = PDI_group, 
         emmean = emmean) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(emmean))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
library(igraph)
library(ggraph)

countries <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(countries, metabs),
  type = c(rep(TRUE, length(countries)), rep(FALSE, length(metabs))) # TRUE = country, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r,fig.height=5,fig.width=8}
country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E"   # red
)

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     country_colors[V(g)$name],
                     "gray70")

library(stringr)
V(g)$label <- V(g)$name |>
  str_remove("^PDI_") |>
  str_remove("_avg$") |>
  str_replace_all("_", " ")

overrides <- c(
  "legume vegetarian prot a" = "legumes & vegetarian protein alternatives",
  "sugar sweetened beverage" = "sugar-sweetened beverages"
)

V(g)$label <- ifelse(V(g)$label %in% names(overrides),
                     overrides[V(g)$label],
                     V(g)$label)

set.seed(123)
 p <- ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_width_continuous(name = "EMM magnitude") +
  scale_edge_color_manual(name = "EMM direction",
                          values = c("lightblue", "lightpink"),
                          labels = c("<0", ">0")) +
  geom_node_point(aes(shape = type, color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +
  scale_shape_manual(name = "Variable",
                     values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("FALSE" = "PDI group", "TRUE" = "Country")) +
  geom_node_text(aes(label = label), repel = TRUE, size = 3,
                 max.overlaps = Inf, box.padding = 0.6, point.padding = 0.4,
                 min.segment.length = 1, segment.size = 0.2,
                 segment.color = "grey60", seed = 123) +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(20, 60, 20, 60)) +
  guides(color = "none")

p
```

#### Significant edges (p_adj<0.1)

**Edge list**

```{r}
edges <- country_effects %>%
  #filter(abs(emmean) > 0.1) %>%
  filter(p_adj<0.1) %>%
  dplyr::select(from = country, to = PDI_group, 
         emmean = emmean) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(emmean))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
library(igraph)
library(ggraph)

countries <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(countries, metabs),
  type = c(rep(TRUE, length(countries)), rep(FALSE, length(metabs))) # TRUE = country, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r,fig.height=5,fig.width=5}
country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E"   # red
)

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     country_colors[V(g)$name],
                     "gray70")

library(stringr)
V(g)$label <- V(g)$name |>
  str_remove("^PDI_") |>
  str_remove("_avg$") |>
  str_replace_all("_", " ")

overrides <- c(
  "legume vegetarian prot a" = "legumes & vegetarian protein alternatives",
  "sugar sweetened beverage" = "sugar-sweetened beverages"
)

V(g)$label <- ifelse(V(g)$label %in% names(overrides),
                     overrides[V(g)$label],
                     V(g)$label)
```

```{r,fig.height=4,fig.width=7}
set.seed(123)

p <- ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_width_continuous(name = "EMM magnitude") +
  scale_edge_color_manual(name = "EMM direction",
                          values = c("lightblue", "lightpink"),
                          labels = c("<0", ">0")) +
  geom_node_point(aes(shape = type, color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +
  scale_shape_manual(name = "Variable",
                     values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("FALSE" = "PDI group", "TRUE" = "Country")) +
  geom_node_text(aes(label = label), repel = TRUE, size = 3,
                 max.overlaps = Inf, box.padding = 0.6, point.padding = 0.4,
                 min.segment.length = 1, segment.size = 0.2,
                 segment.color = "grey60", seed = 123) +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(20, 60, 20, 60)) +
  guides(color = "none")
p
```

```{r}
f <- p
save(f,file="../results/vegan/figures/F3d_CountryPDI.RData")
```

