---
title: "VEGANScreener + metabolomics"
subtitle: "Q5: Associations between PDI groups and the metabolome."
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    df-print: paged
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code is used to explore patterns in the metabolome profiles of vegans, particularly by exploring associations with their dietary records (PDI groups).**


The main parts of the analysis is composed of:

1. Canonical correlation analysis (CCA) - find linear combinations of **PDI groups** and metabolites that are maximally correlated. - SPCA

2. Mantel test - provides a global similarity measure - SPCA

3. Linear modeling - Metabolite ~ sex + bmi + age + country***PDI_group**

4. Network analysis - bipartite graph. Edges of the graph correspond to the significant associations (Estimates from LM). 


```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  file =  "../../data/preprocessed/metadata_preprocessed_v3.tsv",
  header = TRUE)  %>%
  dplyr::filter(group=="VG")

lcms_df <- read.table(
  file= "../../data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",
  check.names = FALSE)

nmr_df <- read.table(
  file = "../../data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",
  check.names = FALSE)

spca_integrated_df <-  read.table(
  file = "../../data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",
  check.names = FALSE)

pdi_df <- as.data.frame(fread(
  file = "../../data/preprocessed/dietary/pdi/pdi_avg_residuals_scaled_v3.tsv")) %>%
  dplyr::select(-c(age,gender,bmi,eq_included_group,study_center,PDI_sweets_and_desserts_alt_avg))

```


```{r}
metadata_scaled_df <- metadata_df
metadata_scaled_df[,-c(1,3,5,6,8)] <- apply(metadata_df[,-c(1,3,5,6,8)],2,scale)
metadata_scaled_df <- metadata_scaled_df %>%
  mutate(sex=as.factor(sex),
         country=as.factor(country),
         group=as.factor(group),
         diet_supplements=as.factor(diet_supplements))
```


# Descriptive statistics

## Metadata

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=15,fig.height=8}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total")
i <- 1
for (col in c("age","bmi","vegan_duration","VS_total")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country")
for (col in c("sex","country")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20)) +
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count(), aes(x=!!sym(col),y=n+10,label=n)) + 
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
    xlab(name[i]) 
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```

## PDI

Basic plots for visualizing the distribution:

```{r,fig.width=25,fig.height=20}
plot_list <- list()
name <- colnames(pdi_df)[-1]
i <- 1
for (col in name){
  p <- ggplot(pdi_df) + 
    geom_histogram(aes(x=as.numeric(!!sym(col))),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(pdi_df[,col],na.rm = TRUE),2),"+-", round(sd(pdi_df[,col],na.rm = TRUE),2)))
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}

wrap_plots(plot_list)

```

**We will remove the animal-related groups.**

```{r}
pdi_df <- pdi_df %>% dplyr::select(-c(PDI_animal_fat_avg,PDI_eggs_avg,PDI_fish_or_seafood_avg,PDI_meat_avg,PDI_dairy_avg))
```

# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```

## Mantel test

The Mantel test measures the correlation between two distance (or dissimilarity) matrices.

In our case:

Dataset 1: PDI profiles (e.g., pdi groups intake per client)
Dataset 2: Metabolomic profiles (e.g., metabolite concentrations per client)

The Mantel test asks: Are clients who eat similarly also similar in their metabolome?

----------------------

```{r}
#nutri <- pdi_df[,c(-19,-20)] %>% column_to_rownames("study_id")
nutri <- pdi_df[,-c(14:21)] %>% column_to_rownames("study_id")

# delete the zero variance columns
cols_to_del <- which(apply(nutri, 2, var, na.rm = TRUE) == 0)
if (length(cols_to_del)>0){
  nutri <- nutri[,-cols_to_del]

}
```

### Euclidean distance

**Simple Mantel test**

The Mantel statistic r: 0.1303 (p=0.001).

```{r}
# Compute distance matrices (e.g., Euclidean)
dist_nutri <- vegdist(nutri,method = "euclidean",na.rm = TRUE)
dist_metabo <-  vegdist(concat_integrated_df,method="euclidean",na.rm = TRUE)

anyNA(dist_nutri)
anyNA(dist_metabo)

# Mantel test
mantel_result <- mantel(dist_nutri, dist_metabo, method = "spearman", 
                        permutations = 999)
print(mantel_result)
```

```{r}
dist_nutri_df <- dist_nutri %>%
  as.matrix() %>% 
  as_tibble(rownames = "A") %>%
  pivot_longer(cols = -A, names_to = "B", values_to = "nutri") %>%
  mutate_if(is.character, str_trim)

dist_metabo_df <- dist_metabo %>%
  as.matrix() %>% 
  as_tibble(rownames = "A") %>%
  pivot_longer(cols = -A, names_to = "B", values_to = "metabo") %>%
  mutate_if(is.character, str_trim)
```

```{r}
combined <- inner_join(dist_nutri_df,dist_metabo_df, by=c("A","B")) %>%
  dplyr::filter(A!=B) %>%
  dplyr::filter(A < B)
  
ggplot(combined,aes(nutri,metabo)) + 
  geom_point() + 
  geom_smooth(se=FALSE)

cor.test(combined$nutri,combined$metabo,method = "spearman")
```

**Partial Mantel test**

The partial mantel is also a correlation test between two matrices, but controlling for the effect of a third distance matrix. This control is done by calculating the correlation between the residuals of each of the two main distance matrices after a linear regression on the third distance matrix.


After adjusting for the country effect, the mantel statistic r: 0.1224  (p=0.001) 
      
```{r}
country_vec <- metadata_df$country  # same order as in your data

dist_country <- as.dist(
  outer(country_vec, country_vec, FUN = Vectorize(function(x, y) as.numeric(x != y)))
)

mantel.partial(dist_nutri, dist_metabo, dist_country, permutations = 999,method = "spearman")

```

## PERMANOVA

```{r}
df_to_test <- merge(pdi_df,metadata_scaled_df,by="study_id",all=TRUE)
```

```{r}
vars_to_test <- colnames(nutri)

adonis_result_all <- data.frame(Var=NA,
                                Df=NA,
                                SumOfSqs=NA,
                                R2=NA,
                                `F`=NA,
                                `Pr(>F)`=NA,
                                check.names = FALSE)
for (variable in vars_to_test){
  metadata_sub_df <- df_to_test %>% dplyr::select(all_of(
    c("study_id","sex", "bmi","age","VS_total","vegan_duration", variable))) %>% drop_na()
  sub_spca <- spca_integrated_df[as.character(metadata_sub_df$study_id),]
  
  # Create Euclidean distance matrix
  dist_mat <- vegdist(sub_spca, 
                    method = "euclidean")
  
  # Test 
  adonis_result <- adonis2(
  as.formula(paste0("dist_mat ~ sex + bmi + age + vegan_duration + ", variable)),
  data = metadata_sub_df , 
  permutations = 999,
  by="terms")

  adonis_result_all <- rbind(adonis_result_all,as.data.frame(adonis_result) %>% 
                               rownames_to_column("Var"))
}


adonis_result_all <- adonis_result_all %>% dplyr::filter(grepl("PDI_",Var))
adonis_result_all$adjusted_p <- p.adjust(adonis_result_all$`Pr(>F)`,method = "BH")

knitr::kable(adonis_result_all)

```

```{r}
labels <- c(
  "PDI_legume_vegetarian_prot_a_avg" = "Legumes & vegetarian protein alternatives",
  "PDI_sweets_and_desserts_avg"      = "Sweets & desserts",
  "PDI_sugar_sweetened_beverage_avg" = "Sugar-sweetened beverages",
  "PDI_fruits_avg" = "Fruits",
  "PDI_vegetables_avg" = "Vegetables",
  "PDI_nuts_avg" = "Nuts",
  "PDI_whole_grains_avg" = "Whole grains",
  "PDI_refined_grains_avg" = "Refined grains",
  "PDI_vegetable_oils_avg" = "Vegetable oils",
  "PDI_tea_and_coffee_avg" = "Tea & coffee",
  "PDI_fruit_juices_avg" = "Fruit juices",
  "PDI_potatoes_avg" = "Potatoes",
  "PDI_nuts_avg" = "Nuts")

adonis_result_all$Var <- labels[adonis_result_all$Var]
adonis_result_all$Category <- c("unhealthy","healthy","healthy","healthy",
                                "unhealthy","unhealthy","unhealthy","unhealthy",
                                "healthy","healthy","healthy","healthy")

adonis_result_all$Var <- factor(
  adonis_result_all$Var,
  levels = adonis_result_all$Var[order(adonis_result_all$R2,decreasing = TRUE)]
)

p_permanova <- ggplot(adonis_result_all, aes(x = Var, y = R2*100, fill = Category)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.3) +
  # mark non-significant bars with "ns"
  geom_text(data = filter(adonis_result_all, adjusted_p > 0.05),
            aes(label = "ns", y = R2*100 + 0.05),
            size = 3, colour = "grey40", fontface = "italic") +
  scale_fill_manual(values = c("healthy"= "#66c2a5","unhealthy"= "#f46d43"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "R² (%)") +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x      = element_text(face = "bold",angle=80,hjust = 1),
    plot.title       = element_blank()
  ) 
p_permanova
```

## Network analysis 

We want to model each metabolite as a function of sex, BMI, age, country, and each food group. Keep the estimated effect (and significance) of each food group on each metabolite.Then, we will use those significant effects to construct a bipartite network, where:

- one node type = metabolites

- other node type = food groups

- edges = significant relationships from the regression models.

Then we’ll analyze and visualize that network.

```{r}
nutri <- pdi_df[,-c(14:21)] %>% column_to_rownames("study_id")

# delete the zero variance columns
cols_to_del <- which(apply(nutri, 2, var, na.rm = TRUE) == 0)
if (length(cols_to_del)>0){
  nutri <- nutri[,-cols_to_del]
}

# delete the zero variance columns
nutri <- nutri %>%
  rownames_to_column("study_id")

metadata_for_lm <- merge(nutri,metadata_scaled_df,by="study_id",all=TRUE)

concat_for_uni <- merge(
  concat_integrated_df %>% rownames_to_column("study_id"),
  metadata_for_lm,
  by="study_id",all = TRUE)
```

### LM

For each metabolite and PDI group, calculate: 

Metabolite ~ sex + bmi + age + country + PDI_group

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "pdi_groups"))

results_df <- do.call(rbind, results_list)

results_df <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

pdi_effects <- results_df %>% dplyr::filter(term %in% colnames(nutri))

concat_significant_cat <-  pdi_effects %>%
  filter(p_adj < 0.05)

concat_significant_cat
```

```{r}
results_all <- pdi_effects

# save 
write.table(results_all,"../results/vegan/Q5_PDIgroups/lm_continous_all.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")
```

### Graph construction

**Edge list**

```{r}
edges <- pdi_effects %>%
  dplyr::filter(p_adj < 0.05) %>%
  dplyr::select(from = term, to = Metabolite, 
         estimate = estimate) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(estimate))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
pdis <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(pdis, metabs),
  type = c(rep(TRUE, length(pdis)), rep(FALSE, length(metabs))) # TRUE = tertile, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r}
categories <- list(
  "healthy" = c("PDI_fruits_avg",
                "PDI_legume_vegetarian_prot_a_avg",
                "PDI_nuts_avg",
                "PDI_tea_and_coffee_avg",
                "PDI_vegetable_oils_avg",
                "PDI_vegetables_avg",
                "PDI_whole_grains_avg"),
  
  "unhealthy" = c("PDI_fruit_juices_avg",
                "PDI_refined_grains_avg",
                "PDI_potatoes_avg",
                "PDI_sugar_sweetened_beverage_avg",
                "PDI_sweets_and_desserts_avg",
                "PDI_sweets_and_desserts_alt_avg"))	
```

```{r,fig.height=6,fig.width=6}
set.seed(123)
pdi_colors <- colorRampPalette(
  c("#2596be","#9e0142","#d53e4f","#f46d43",
    "#fdae61","#e6f598","#66c2a5","#5e4fa2"))(12)

pdi_colors <- c(rep("#66c2a5",length(categories$healthy)),
                rep("#f46d43",length(categories$unhealthy)))

names(pdi_colors) <- c(categories$healthy,categories$unhealthy)

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     pdi_colors[V(g)$name],
                     "gray70")

node_labels <- c(
  "PDI_legume_vegetarian_prot_a_avg" = "Legumes & vegetarian protein alternatives",
  "PDI_sweets_and_desserts_avg"      = "Sweets & desserts",
  "PDI_sugar_sweetened_beverage_avg" = "Sugar-sweetened beverages",
  "PDI_fruits_avg" = "Fruits",
  "PDI_vegetables_avg" = "Vegetables",
  "PDI_nuts_avg" = "Nuts",
  "PDI_whole_grains_avg" = "Whole grains",
  "PDI_refined_grains_avg" = "Refined grains",
  "PDI_vegetable_oils_avg" = "Vegetable oils",
  "PDI_tea_and_coffee_avg" = "Tea & coffee",
  "PDI_fruit_juices_avg" = "Fruit juices",
  "PDI_nuts_avg" = "Nuts",
  "Indoleacrylic.acid"            = "Indoleacrylic acid",
    "Acetylornithin"                = "Acetylornithine",        # NB: source typo (missing 'e'), not a dot issue
    "Ergothionenine"                = "Ergothioneine",          # NB: source typo, not a dot issue
    "Betaine"                       = "Betaine",
    "Asparagine"                    = "Asparagine",
    "Formate"                       = "Formate",
    "Glutamylglutamine"             = "Glutamylglutamine",
    "Hydroxydecanoylcarnitine"      = "Hydroxydecanoylcarnitine",
    "Hydroxyoctanoylcarnitine"      = "Hydroxyoctanoylcarnitine",
    "Glycine"                       = "Glycine",
    "3-Methyl-2-oxovalerate"        = "3-Methyl-2-oxovalerate",
    "Acetate"                       = "Acetate",
    "Citrate"                       = "Citrate",
    "Octenoylcarnitine"             = "Octenoylcarnitine",
    "Dodecenoylcarnitine"           = "Dodecenoylcarnitine",
    "Dimethylamine"                 = "Dimethylamine",
    "Uric.acid"                     = "Uric acid",
    "Alanine"                       = "Alanine",
    "3-Hydroxybutyrate"             = "3-Hydroxybutyrate",
    "Glucose"                       = "Glucose",
    "Hydroxybutyrylcarnitine"       = "Hydroxybutyrylcarnitine",
    "Lactate"                       = "Lactate",
    "Tyrosine"                      = "Tyrosine",
    "Inosine"                       = "Inosine",
    "Histidine"                     = "Histidine",
    "Phenylalanine"                 = "Phenylalanine",
    "Threonine"                     = "Threonine",
    "Phenylalanylphenylalanine"     = "Phenylalanylphenylalanine",
    "Methylguanidine"               = "Methylguanidine",
    "Glycerophosphocholine"         = "Glycerophosphocholine",
    "Acetone"                       = "Acetone",
    "Caffeine"                      = "Caffeine",
    "Hexanoylcarnitine"             = "Hexanoylcarnitine",
    "Phenylbutyrylglutamine"        = "Phenylbutyrylglutamine",
    "Trimethyllysine"               = "Trimethyllysine",
    "Isoleucine"                    = "Isoleucine",
    "Carnitine"                     = "Carnitine",
    "2-Oxoisocaproate"              = "2-Oxoisocaproate",
    "Propionylcarnitine"            = "Propionylcarnitine",
    "Glutamine"                     = "Glutamine",
    "Creatinine"                    = "Creatinine",
    "Valine"                        = "Valine",
    "Butyrylcarnitine"              = "Butyrylcarnitine",
    "Glutarylcarnitine"             = "Glutarylcarnitine",
    "2-Hydroxybutyrate"             = "2-Hydroxybutyrate",
    "Pyruvate"                      = "Pyruvate",
    "Lysine"                        = "Lysine",
    "Tryptophan"                    = "Tryptophan",
    "Diaminonaphthalene"            = "Diaminonaphthalene",
    "Ascorbate"                     = "Ascorbate",
    
    # ---- lipids: dots restored to ( ) and : ----
    "PC.36.2."                      = "PC(36:2)",
    "LysoPC.18.2."                  = "LysoPC(18:2)",
    "SM.d42.2."                     = "SM(d42:2)",
    "PC.P.38.4."                    = "PC(P-38:4)",
    "PC.38.4."                      = "PC(38:4)",
    "SM.d38.2."                     = "SM(d38:2)",
    "PC.36.4."                      = "PC(36:4)",
    "SM.d34.1."                     = "SM(d34:1)",
    "LysoPC.14.0."                  = "LysoPC(14:0)",
    "PC.34.1."                      = "PC(34:1)",
    "PC.40.6."                      = "PC(40:6)",
    "LysoPC.22.6."                  = "LysoPC(22:6)",
    "PA.44.5."                      = "PA(44:5)",
    "PC.32.1."                      = "PC(32:1)",
    "LysoPC.P.16.0."                = "LysoPC(P-16:0)",
    "LysoPC.17.0."                  = "LysoPC(17:0)",
    "SM.d32.1."                     = "SM(d32:1)",
    "SM.d33.1."                     = "SM(d33:1)",
    
    # ---- acylcarnitine chain-length annotations ----
    "Acylcarnitine..C8."            = "Acylcarnitine (C8)",
    "Acylcarnitine..C5."            = "Acylcarnitine (C5)",
  
  # ---- VERIFY THESE: adducts / ambiguous combined names ----
  # The reconstruction below is my best guess; confirm against your raw list.
  "Citrulline..M.Na."             = "Citrulline (M+Na)",
  "CE.18.2..M.NH4.."              = "CE(18:2)[M+NH4]",
  "PC.20.3.OH.P.18.1."            = "PC(20:3-OH/P-18:1)",
  "PC.P.38.4..PC.O.38.5."         = "PC(P-38:4)/PC(O-38:5)")


# create a label attribute; fall back to the original name if not in the dict
V(g)$label <- ifelse(V(g)$name %in% names(node_labels),
                     node_labels[V(g)$name],
                     V(g)$name)

set.seed(123)
p <- ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = estimate > 0), alpha = 0.7) +
  scale_edge_width_continuous(name = "EMM magnitude") +
  scale_edge_color_manual(
    values = c("lightblue", "lightpink"), 
    name="EMM direction",
    labels = c("<0", ">0")) +
  geom_node_point(aes(shape=type,color = name), size = 6,,name="Node",labels=c("Metabolite","PDI group")) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +  # node colors
  geom_node_text(aes(label = label), repel = TRUE, size = 3,
                 max.overlaps = Inf, box.padding = 0.6, point.padding = 0.4,
                 min.segment.length = 1, segment.size = 0.2,
                 segment.color = "grey60", seed = 123) + 
  scale_shape_manual(name = "Variable",
                     values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("TRUE" = "PDI group", "FALSE" = "Metabolite"))  +
  guides(color = "none") + theme_void()

p

```

```{r}
f <- p
save(f,file="../results/vegan/figures/F5e_Network.RData")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q5_PDIgroups/network_pdi.pdf",height = 6,width = 6)
p
dev.off()
```

### Degree calculation

Interpretation

*High-degree PDI nodes (type == TRUE)*

→ These dietary variables are associated with many metabolites
→ Could suggest broad metabolic influence

*High-degree metabolite nodes (type == FALSE)*

→ These metabolites are associated with multiple food groups
→ Potential nutritional biomarkers


```{r}
# Calculate degree
deg <- degree(g, mode = "all")

# Add to node dataframe
node_degree_df <- data.frame(
  name = names(deg),
  degree = deg,
  type = V(g)$type
)

# Separate interpretation
food_nodes <- node_degree_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_degree_df %>% dplyr::filter(type == FALSE)

food_nodes %>% arrange(desc(degree))
metabolite_nodes %>% arrange(desc(degree))
```

**Weighted degree**
- Should reflect cumulative association magnitude, not just number of links.

```{r}
strength_vals <- strength(g, weights = E(g)$weight)

node_strength_df <- data.frame(
  name = names(strength_vals),
  strength = strength_vals,
  type = V(g)$type
)

# Separate interpretation
food_nodes <- node_strength_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_strength_df %>% dplyr::filter(type == FALSE)

food_nodes %>% arrange(desc(strength))
metabolite_nodes %>% arrange(desc(strength))
```
