---
title: "VEGANScreener + metabolomics"
subtitle: "Q1: Differences in the metabolome of vegans across different countries"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code is used to explore patterns in the metabolome profiles of vegans, particularly to distinguish between different countries of origin.**

The script is divided into two parts, according to the integration method of the metabolome datasets (see preprocessing scripts for details). At first, the **sPCA dataset** is analyzed. This dataset is used for multivariate analysis, visualization, PERMANOVA testing and ML approach. The **concatenated dataset** of both LCMS and NMR is used for the univariate analysis - linear modelling with a visualization thorugh the bipartite graphs. 

The main parts of the analysis is composed of:

1.  Global visualization - 3 components color-coding country.

2.  PERMANOVA - how much of the variation can be explained by various external variables:
  dist(Metabolome) \~ sex + bmi + age + VS_total + veganDuration + country

3.  Univariate analysis (feature-wise model) - linear modelling of metabolite using various external variables:
  Metabolite \~  sex + bmi + age + VS_total + veganDuration + country
  
Individual means and p-values are obtained by the emmeans() function. The bipartite network is then constructed and color-coded based on the strongest association to the analyzed factor.

4.  Random forest (RF) classification - the model is trained and validated through bootstrapping with 500 iterations. The RF model is used to distinguish country of origin. The two types of models are trained:
a) ML model on original sPCA dataset
b) ML model on sPCA dataset after regressing out the covariates.

```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  file = "../../data/preprocessed/metadata_preprocessed_v3.tsv", 
  header = TRUE) %>%
  dplyr::filter(group=="VG")

lcms_df <- read.table(
  file = "../../data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",
  check.names = FALSE)

nmr_df <- read.table(
  file = "../../data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",
  check.names = FALSE)

spca_integrated_df <-  read.table(
  file = "../../data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",
  check.names = FALSE)
```


**Scaling** - when using columns from metadata as covariates, they need to be scaled, so mean is 0 and SD is 0.

```{r}
metadata_scaled_df <- metadata_df
metadata_scaled_df[,-c(1,3,5,6,8)] <- apply(metadata_df[,-c(1,3,5,6,8)],2,scale)
metadata_scaled_df <- metadata_scaled_df %>%
  mutate(sex=as.factor(sex),
         country=as.factor(country),
         group=as.factor(group),
         diet_supplements=as.factor(diet_supplements))
```

# Descriptive statistics

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=10,fig.height=5}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total")
i <- 1
for (col in c("age","bmi","vegan_duration","VS_total")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country")
for (col in c("sex","country")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20)) +
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
    xlab(name[i]) +
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count() %>%
              dplyr::mutate(n=n+10), aes(x=!!sym(col),y=n,label=n))
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```

# SPCA

The SPCA dataset was created by concatenating LCMS and NMR dataset and applying Mixomics::spca() function. See preprocessing script for details. 

mixOmics::spca() performs **Sparse Principal Component Analysis**, which is similar to regular PCA but adds a sparsity constraint so that only a subset of variables contributes to each component.
This sparsity is controlled by penalization, making interpretation easier because irrelevant or noisy variables get loadings of zero.

## Visualization

### PCA

```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "country",
              perform_pca = FALSE)

p <- ggtheme(p)

p

p_pca <- list()
p_pca[[1]] <- p
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/PCA.pdf",height = 4,width = 6)
p
dev.off()
```

```{r,fig.width=15,fig.height=4}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p1 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "country",
              perform_pca = FALSE,axis = c("PC1","PC2"))

p1 <- ggtheme(p1)

p2 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "country",
              perform_pca = FALSE,axis = c("PC2","PC3"))

p2 <- ggtheme(p2)

p3 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "country",
              perform_pca = FALSE,axis = c("PC1","PC3"))

p3 <- ggtheme(p3)

p1 + p2 + p3 + 
plot_layout(guides = "collect")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/PCA_all.pdf",height = 4,width = 15)
p1 + p2 + p3 + 
plot_layout(guides = "collect")
dev.off()
```

### Boxplots

```{r,fig.height=15,fig.width=20}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_scaled_df,by="study_id",all = TRUE) 

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = "country",y = col,fill="country")) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values=c(
      BE = "#FDDA24",   # gold
      CZ = "#11457E",   # blue
      DE = "#000000",   # black
      ES = "#630356",   # purple
      CH = "#D52B1E"    # red
    )) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$country)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p)
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q1_country/pca_boxplots.pdf",height = 15,width = 20)
a
dev.off()
```

### Heatmap

```{r}
library(pheatmap)

country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E"   # red
)

p_df <- spca_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,country) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "country"=country_colors
)
names(ann_colors[["country"]]) <- names(country_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q1_country/pca_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```

## PERMANOVA

The test shows how much of the variation in metabolomics is explained by: sex - age - bmi - VS total - vegan duration - Country

Variables were added sequentially (Type I SS), so each p-value and R² reflects a partial effect, after adjusting for those above. This model accounts for variation due to sex, age, vegan diet duration and BMI. Then accounts for the main effects of country. 

Country explains **9.02%** of variation after adjusting for sex, age, bmi, VS_total and duration of the diet. This result is statistically significant (p<0.001).

```{r}
metadata_to_test <- metadata_scaled_df %>% drop_na(VS_total,vegan_duration)

# Create Euclidean distance matrix
dist_mat <- vegdist(spca_integrated_df, 
                    method = "euclidean")

# Test whether countries differ in metabolomics
adonis_result <- adonis2(
  dist_mat ~ sex + bmi + age + VS_total + vegan_duration + country,
  data = metadata_scaled_df , 
  permutations = 999,
  by="terms")

print(adonis_result)
```

```{r,fig.width=6,fig.height=2}
adonis_result <- adonis_result[-8,] %>% rownames_to_column("Var")
adonis_result$Var <- factor(adonis_result$Var,levels = rev(c("sex","age","bmi","VS_total", "vegan_duration","country", "Residual")))
p <- ggplot(adonis_result, aes(fill=Var, x=R2, y="Metabolome")) + 
    geom_bar(position="stack", stat="identity") + 
  ylab("") + xlab("Partial variance explained (R²)") + 
  scale_fill_manual(values=rev(c("#ffffcc","#b3cde3","#ccebc5",
                                 "#decbe4","#fed9a6","darkred",
                                 "lightgrey")),name = "Covariate")  

p <- ggtheme(p)

p

p_pca[[2]] <- p
p_pca[[3]] <- adonis_result
```

**Pairwise PERMANOVA**

```{r}
pp_main <- pairwise.adonis(x = t(spca_integrated_df),
                metadata = metadata_scaled_df %>%
                  dplyr::mutate(SampleID=study_id),
                formula = "dist_mat ~ sex + bmi + age + VS_total + vegan_duration + country",main_variable = "country",
                sim.method = "euclidean"
                  )
```

```{r}
pp_main
```

## RF model

Model performance significance was assessed using a bootstrap-based z-test. The AUC was estimated over 500 bootstrap resamples, and the bootstrap standard error was used to compute a Wald-type statistic z = (A − 0.5)/SE, with p-values obtained from the standard normal distribution.

### Original model

```{r}
model_name <- "country_model_vegan" 

prepared_df <- model_preparation(spca_integrated_df,
                                 metadata_df,
                                 predicting_col = "country")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 500,
                          file = model_name,
                          reuse=TRUE)

rf_model$model_summary
roc_c <- roc_curve(model_object = rf_model,CI = TRUE)

roc_c <- ggtheme(roc_c)

roc_c
```

**Accuracy**

```{r}
mean((rf_model$predictions %>%
  dplyr::group_by(iteration) %>%
  dplyr::summarize(accuracy = mean(predicted == outcome, na.rm = TRUE)))$accuracy)
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/orig_rf_country.pdf",width = 6,height = 5)
roc_c
dev.off()
```

```{r}
# saving the results
save_rf_details("../results/vegan/RF_models_performance.xlsx",model_name,rf_model)
```

### Regressed out model

```{r}
# Covariates to adjust for
covars <- c("sex","bmi", "age", "VS_total", "vegan_duration")
integrated_df <- spca_integrated_df
all(rownames(integrated_df) == metadata_df$study_id)
adjusted_data <- adjust_for_covars(integrated_df, covars, metadata_scaled_df)

model_name <- "adjusted_country_model_vegan" 

prepared_df <- model_preparation(adjusted_data,
                                 metadata_df,
                                 predicting_col = "country")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 500,
                          file = model_name,
                          reuse=TRUE)
rf_model$model_summary
roc_c <- roc_curve(model_object = rf_model,CI = TRUE)

roc_c <- ggtheme(roc_c) 

roc_c 

```

```{r}
f <- roc_c 
save(f,file="../results/vegan/figures/F3a_ROCcountry.RData")
```

**Confusion matrix**

```{r}
conf_matrix <- rf_model$predictions %>%
  group_by(iteration, outcome, predicted) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(outcome, predicted) %>%
  summarise(mean_count = mean(n), .groups = "drop") %>%
  pivot_wider(names_from = predicted, values_from = mean_count, values_fill = 0) %>%
  as.data.frame()

conf_matrix

conf_matrix[,-1] <- conf_matrix[,-1]/rowSums(conf_matrix[,-1])

conf_matrix
```

```{r}
conf_matrix <- conf_matrix[,-1]
countries <- as.character(rf_model$mapping)
rownames(conf_matrix) <- countries
colnames(conf_matrix) <- countries

p <- confusion_matrix_plot(conf_matrix,countries)
p

```


```{r}
f <- p
save(f,file="../results/vegan/figures/F3b_CMcountry.RData")
```


```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/adjusted_rf_country.pdf",width = 6,height = 5)
roc_c
dev.off()
```

```{r}
# saving the results
save_rf_details("../results/vegan/RF_models_performance.xlsx",model_name,rf_model)
```

## RF importance

### Original model

**PC4** and **PC10** are the most important for RF model. However, no significant drop occurred in other PCs.

```{r,fig.width=10}
load("../intermediate_files/models/country_model_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Adjusted model

Similarly as above, the most important is **PC4**. The PC10 dropped comparing to 1st value in the original model.

```{r}
load("../intermediate_files/models/adjusted_country_model_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Visualization

To explore these two important PCs more, their visualization is shown below. 

*PCA* 

```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_scaled_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "country",
              perform_pca = FALSE,
              axis = c("PC4","PC10"))

p <- ggtheme(p)

p

p_pca[[4]] <- p
```
```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/important_PCA.pdf",height = 4,width = 6)
p
dev.off()
```

*Boxplots*

```{r}
plot_list <- list()
for (col in c("PC4","PC10")){
 p <- ggplot(spca_for_uni, aes_string(x = "country",y = col,fill="country")) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values=c(
      BE = "#FDDA24",   # gold
      CZ = "#11457E",   # blue
      DE = "#000000",   # black
      ES = "#630356",   # purple
      CH = "#D52B1E"    # red
    )) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$country)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p) 
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a

```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/important_pca_boxplots.pdf",height = 4,width = 8)
a
dev.off()
```

## SPCA loadings

Here, the loadings of the two important PCs (**PC4** and **PC10**) are explored and saved.

```{r,collapse=TRUE}
load("../results/preprocessing/spca_result_all.RData")

df <- data.frame (
  Metabolite=names(spca_result$loadings$X[,4]),
  Loading=spca_result$loadings$X[,4])

df

write.table(df,sep="\t","../results/vegan/Q1_country/loadings_PC4.tsv")
```

This plot shows the loading of individual metabolites to **PC4**, ordered by its absolute value.

```{r,fig.height=10,fig.width=5,collapse=TRUE}
ggplot(df %>% dplyr::mutate(
           Metabolite=factor(
             Metabolite,
             levels = Metabolite[order(abs(Loading),decreasing = FALSE)])
           )) + 
  geom_col(aes(x=Metabolite,y=Loading)) + 
  theme(axis.text.x = element_text(angle=90,hjust = 1)) + 
  coord_flip()

spca_loadings <- df$Metabolite[abs(df$Loading)>=0.1]
```

```{r}
df <- data.frame (
  Metabolite=names(spca_result$loadings$X[,10]),
  Loading=spca_result$loadings$X[,10])

write.table(df,sep="\t","../results/vegan/Q1_country/loadings_PC10.tsv")
```

This plot shows the loading of individual metabolites to **PC10**, ordered by its absolute value.

```{r,fig.height=10,fig.width=5,collapse=TRUE}
ggplot(df %>% 
         dplyr::mutate(
           Metabolite=factor(Metabolite,
                             levels = Metabolite[order(abs(Loading),decreasing = FALSE)])
           )) + 
  geom_col(aes(x=Metabolite,y=Loading)) + 
  theme(axis.text.x = element_text(angle=90,hjust = 1)) + 
  coord_flip()

spca_loadings <- c(spca_loadings,df$Metabolite[abs(df$Loading)>=0.1])
```

# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```

## Visualization

### Boxplots

```{r,fig.height=40,fig.width=20}
concat_for_uni <- merge(concat_integrated_df %>% rownames_to_column("study_id"),
                      metadata_scaled_df,by="study_id",all = TRUE) 

plot_list <- list()
for (col in colnames(concat_integrated_df)){
  p <- ggplot(concat_for_uni, 
              aes(x = country,y = !!sym(col),fill=country)) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values=c(
      BE = "#FDDA24",   # gold
      CZ = "#11457E",   # blue
      DE = "#000000",   # black
      ES = "#630356",   # purple
      CH = "#D52B1E"    # red
    )) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(concat_for_uni$country)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p)
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q1_country/concat_boxplots.pdf",height = 35,width = 20)
a
dev.off()
```

### Heatmap

```{r}
library(pheatmap)
p_df <- concat_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,country) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "country"=country_colors
)
names(ann_colors[["country"]]) <- names(country_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              show_rownames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q1_country/concat_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```


## Univariate analysis

The lm model is used to test the differences in countries, while adjusting for covariates:

Metabolite ~ sex + bmi + age + VS_total + vegan_duration + Country
    

### LM

**LM**

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "country"))
results_df <- do.call(rbind, results_list)

country_effects <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

concat_significant <-  country_effects %>%
  filter(p_adj < 0.05)

# save and show
write.table(concat_significant,"../results/vegan/Q1_country/lm_emmeans.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")

knitr::kable(concat_significant)
```

```{r}
results_all <- country_effects

# save 
write.table(results_all,"../results/vegan/Q1_country/lm_emmeans_all.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")
```

**Dispersion**

```{r}
dispersion <- country_effects %>%
  group_by(metabolite) %>%
  summarise(
    range      = max(emmean) - min(emmean),      # simplest: spread between highest and lowest country
    sd         = sd(emmean),                      # standard deviation of the 5 country means
    cv         = sd(emmean) / abs(mean(emmean)),  # coefficient of variation (relative dispersion)
    min_mean   = min(emmean),
    max_mean   = max(emmean),
    min_country = country[which.min(emmean)],
    max_country = country[which.max(emmean)]
  ) %>%
  arrange(desc(range))

dispersion

write.table(dispersion,"../results/vegan/Q1_country/dispersion.tsv",sep="\t",row.names = FALSE,quote=FALSE)
```

```{r}
results_list_pairwise <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm_pairwise(x,concat_for_uni,
                                           tested_variable = "country"))
results_df_pairwise <- do.call(rbind, results_list_pairwise)

country_effects_pairwise <- results_df_pairwise %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

concat_significant_pairwise <-  country_effects_pairwise %>%
  filter(p_adj < 0.05)

knitr::kable(concat_significant_pairwise)
```


### Interesting metabolites

```{r,fig.height=7, fig.width=5}
p_emmeans_heatmap <- direction_heatmap(country_effects)
p_emmeans_heatmap <- fix_labels(p_emmeans_heatmap)
p_emmeans_heatmap
```

```{r}
f <- p_emmeans_heatmap
save(f,file="../results/vegan/figures/F3c_CountryHeatmap.RData")
```

### Forest plot

```{r, fig.height=15,fig.width=5}

country_effects <- merge(country_effects,dispersion[,c("metabolite","range"),all.x=TRUE,by="metabolite"])

# Define custom country colours
country_colors <- c(
  "BE" = "#FDDA24",  # gold
  "CZ" = "#11457E",  # blue
  "DE" = "#000000",  # black
  "ES" = "#630356",  # purple
  "CH" = "#D52B1E"   # red
)

# Optional: order metabolites by mean effect size
forest_df <- country_effects %>%
  group_by(metabolite) %>%
  mutate(range_eff = mean(range)) %>%
  ungroup()

forest_df$metabolite <- factor(
  forest_df$metabolite,
  levels = forest_df %>% 
    arrange(range_eff) %>% 
    pull(metabolite) %>% 
    unique()
)

# Forest plot WITHOUT ANY TEXT LABELS
forest_df <- forest_df %>% mutate(sig = p_adj < 0.05)
p_forest <- ggplot(forest_df, 
                   aes(x = emmean, 
                       y = metabolite, 
                       color = country)) +

  geom_errorbarh(aes(xmin = lower.CL, xmax = upper.CL), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2.5,aes(shape = sig)) +
  scale_shape_manual(values = c(`FALSE` = 20, `TRUE` = 19),   # 1 = open, 19 = solid
                     labels = c("ns", "FDR < 0.05"),
                     name   = "Significance") + 
  scale_color_manual(values = country_colors) +

  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6),
        panel.grid.minor = element_blank()) +

  labs(
    x = "Estimated mean (with 95% CI)",
    color = "Country"
  )

p_forest <- fix_labels(p_forest)

p_forest

```



```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/final_significance_forest.pdf",height = 15,width = 6)
p_forest
dev.off()
```

```{r}
f <- p_forest
save(f,file="../results/vegan/figures/FS_CountryForest.RData")
```

### Heatmap

```{r}
df2 <- country_effects_pairwise %>%
  mutate(
    contrast = gsub(" ", "", contrast),   # remove spaces, e.g. "BE-DE"
    signif = p_adj < 0.05
  )

df2$Significance <- ifelse(df2$p_adj<0.001,"***",
                                       ifelse(df2$p_adj<0.01,"**",
                                              ifelse(df2$p_adj<0.05,"*","")))

```

```{r,fig.height=8,fig.width=5}
p_pairwise_heatmap <- ggplot(df2, aes(x = contrast, y = Metabolite, fill = estimate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Significance,vjust=0.6), color = "black", size = 2) +
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0) + 
  theme_minimal() +
  theme(axis.text.y = element_text(size=4),
        axis.text.x = element_text(angle = 45,face="bold",hjust = 1),
        panel.grid = element_blank()) + 
  xlab("") + ylab("")

p_pairwise_heatmap
```

### Network

**Edge list**

```{r}
edges <- country_effects %>%
  #filter(abs(emmean) > 0.3) %>%
  filter(p_adj<0.05) %>%
  dplyr::select(from = country, to = metabolite, 
         emmean = emmean) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(emmean))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
library(igraph)
library(ggraph)

countries <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(countries, metabs),
  type = c(rep(TRUE, length(countries)), rep(FALSE, length(metabs))) # TRUE = country, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r,fig.height=7,fig.width=7}
country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E"   # red
)

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     country_colors[V(g)$name],
                     "gray70")

set.seed(123)
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue", "lightpink"), labels = c("<0", ">0")) +
  geom_node_point(aes(shape=type,color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +  # node colors
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  guides(color = "none")

```

#### Degree calculation

A country with high degree means it has many metabolites significantly associated with it.
We can interpret it in different ways. It has:
- Strong metabolic distinctiveness
- Broad metabolic shift
- Possibly strong dietary/cultural metabolic signature
- Or stronger statistical power !!

High-degree metabolite can mean:
- High cross-country variability
- Reflects common axis of variation

Low-degree metabolite (1–2 countries)
- More country-specific signal
- Better candidate geographic biomarker

```{r}
# Calculate degree
deg <- degree(g, mode = "all")

# Add to node dataframe
node_degree_df <- data.frame(
  name = names(deg),
  degree = deg,
  type = V(g)$type
)

# Separate interpretation
country_nodes <- node_degree_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_degree_df %>% dplyr::filter(type == FALSE)

country_nodes %>% arrange(desc(degree))
metabolite_nodes %>% arrange(desc(degree))
```

**Weighted degree**
- Should reflect cumulative association magnitude, not just number of links.

```{r}
strength_vals <- strength(g, weights = E(g)$weight)

node_strength_df <- data.frame(
  name = names(strength_vals),
  strength = strength_vals,
  type = V(g)$type
)

# Separate interpretation
country_nodes <- node_strength_df %>% dplyr::filter(type == TRUE)
metabolite_nodes <- node_strength_df %>% dplyr::filter(type == FALSE)

country_nodes %>% arrange(desc(strength))
metabolite_nodes %>% arrange(desc(strength))
```

#### Top-group assignment

**Based on positive associations**

```{r}
edges_pos <- edges %>%
  filter(emmean > 0)  # keep only positive associations

g_pos <- graph_from_data_frame(
  d = edges_pos, 
  vertices = nodes, 
  directed = FALSE)

country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E",   # red
  "other" = "gray70"
)

# Assign colors to nodes: countries get their color, metabolites gray
V(g_pos)$color <- ifelse(V(g_pos)$type,
                     country_colors[V(g_pos)$name],
                     "gray70")

top_group <- edges_pos %>%
  dplyr::group_by(to) %>%
  slice_max(order_by = weight, n = 1, with_ties = FALSE) %>%
  dplyr::select(from,to, weight) %>%
  ungroup()

top_group

top_group_viz <- rbind(top_group, 
                       data.frame(
                         from=unique(edges$from),
                         to=unique(edges$from),
                         weight=NA))

top_group_viz <- rbind(top_group_viz,
                   data.frame(from="other",
                              to=names(V(g_pos))[!names(V(g_pos)) %in% top_group_viz$to],
                              weight=NA))

membership_df <- data.frame(
  vertex = top_group_viz$to,
  cluster = top_group_viz$from) 

membership_df <- membership_df %>% arrange(match(vertex,names(V(g_pos))))


```

```{r}
country_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:5,]

(membership_df %>%
  group_by(cluster) %>%
  summarise(Number_of_metabolites = length(unique(vertex))-1))[1:5,]

country_clusters
```

```{r}
V(g_pos)$fill <- country_colors[membership_df$cluster]
V(g_pos)$color <- country_colors[membership_df$cluster]
```


```{r,fig.height=6,fig.width=6}
set.seed(123)
p_pos <- ggraph(g_pos, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightpink"), labels = c(">0")) +
  geom_node_point(aes(size=type,color = name,fill=name),shape=21,stroke=1.2) +
  scale_color_manual(values = setNames(V(g_pos)$color, V(g_pos)$name))   + # node colors
scale_fill_manual(values = setNames(V(g_pos)$fill, V(g_pos)$name)) +  # node fill
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() + 
  theme(legend.position = "none")

p_pos
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/network_concat_pos.pdf",height = 6,width = 6)
p_pos
dev.off()
```

**Based on negative associations**

```{r}
edges_neg <- edges %>%
  filter(emmean < 0)  # keep only negative associations

g_neg <- graph_from_data_frame(
  d = edges_neg, 
  vertices = nodes, 
  directed = FALSE)

country_colors <- c(
  BE = "#FDDA24",  # gold
  CZ = "#11457E",  # blue
  DE = "#000000",  # black
  ES = "#630356",  # purple
  CH = "#D52B1E",   # red
  "other" = "gray70"
)

# Assign colors to nodes: countries get their color, metabolites gray
V(g_neg)$color <- ifelse(V(g_neg)$type,
                     country_colors[V(g_neg)$name],
                     "gray70")

top_group <- edges_neg %>%
  group_by(to) %>%
  slice_max(order_by = weight, n = 1, with_ties = FALSE) %>%
  dplyr::select(from,to, weight) %>%
  ungroup()

top_group

top_group_viz <- rbind(top_group, 
                       data.frame(
                         from=unique(edges$from),
                         to=unique(edges$from),
                         weight=NA))

top_group_viz <- rbind(top_group_viz,
                   data.frame(from="other",
                              to=names(V(g_neg))[!names(V(g_neg)) %in% top_group_viz$to],
                              weight=NA))

membership_df <- data.frame(
  vertex = top_group_viz$to,
  cluster = top_group_viz$from) 

membership_df <- membership_df %>% arrange(match(vertex,names(V(g_neg))))
```
```{r}
country_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:5,]

(membership_df %>%
  group_by(cluster) %>%
  summarise(Number_of_metabolites = length(unique(vertex))-1))[1:5,]
```

```{r}
vs_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:5,]

vs_clusters 
```

```{r}
V(g_neg)$fill <- country_colors[membership_df$cluster]
V(g_neg)$color <- country_colors[membership_df$cluster]
```

```{r,fig.height=6,fig.width=6}
set.seed(123)
p_neg <- ggraph(g_neg, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue"), labels = c("<0")) +
  geom_node_point(aes(size=type,color = name,fill=name),shape=21,stroke=1.2) +
  scale_color_manual(values = setNames(V(g_neg)$color, V(g_neg)$name))   + # node colors
scale_fill_manual(values = setNames(V(g_neg)$fill, V(g_neg)$name)) +  # node fill
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() + 
  theme(legend.position = "none")

p_neg
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/network_concat_neg.pdf",height = 6,width = 6)
p_neg
dev.off()
```

# Final significant metabolites

::: {.output-background style="background-color: lightgreen; padding: 10px;"}

```{r,fig.height=10,fig.width=15}

list_core <- list()

list_core[["concat_univariate"]] <- unique(concat_significant$metabolite)

```

```{r}
final_significance <- list_core[["concat_univariate"]]
final_significance
```
:::

# Plot for publication

## PCA

```{r,fig.height=4,fig.width=8}
p1 <- p_pca[[1]] 
country_leg <- ggpubr::get_legend(p1)
p1 <- p1 + theme(legend.position = "none")

p2 <- p_pca[[4]] 
p2 <- p2 + theme(legend.position = "none")

p3 <- p_pca[[2]] + 
  geom_text(aes(x=0.15,y=1),label=paste0("R²=",round(p_pca[[3]]$R2[6],2),"\n p<0.001"),size=2,color="white")

factor_leg <- ggpubr::get_legend(p3)
p3 <- p3 + theme(legend.position = "none")

legends <- ggarrange(country_leg, factor_leg, nrow = 2, align = "hv")

layout <- "
AABBDD
AABBDD
AABBDD
CCCCDD
"

# Now plot!
p1 + p2 + p3 + legends + 
plot_layout(design = layout, heights = c(2,2,2,1),widths = c(1,1,1,1,1,1))
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/pca_final.pdf",height = 4,width = 8)
p1 + p2 + p3 + legends + 
plot_layout(design = layout, heights = c(2,2,2,1),widths = c(1,1,1,1,1,1))
dev.off()
```

## Heatmap

### Emmeans

```{r,fig.height=10,fig.width=15}
concat_significant_df <- concat_significant
concat_significant_df <- concat_significant_df %>%
  dplyr::filter(metabolite %in% final_significance) %>%
  dplyr::arrange(emmean)

heatmap_concat_integrated_df <- concat_integrated_df

# Create a grouping variable for each block of 10 rows

# Calculate the mean for each group
df_grouped <- aggregate(heatmap_concat_integrated_df, 
                        by = list(Group = metadata_df$country), FUN = median)

```

LM score

```{r,fig.height=8,fig.width=5}
country_effects$Significance <- ifelse(country_effects$p_adj<0.001,"***",
                                       ifelse(country_effects$p_adj<0.01,"**",
                                              ifelse(country_effects$p_adj<0.05,"*","")))

p_heat <- ggplot(country_effects, aes(x = country, y = metabolite, fill = emmean)) +
  geom_tile(color = "white",width=1) +
   geom_text(aes(label = Significance,vjust=0.6), color = "black", size = 2) +  # Add asterisks for significant points
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold",size = 9),
        axis.text.y = element_text(size=4),
        panel.grid = element_blank()) + 
  labs(fill = "Estimated mean") +
  xlab("") + 
  ylab("") + 
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0)

p_heat
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/final_significance.pdf",height = 7,width = 6)
p_heat
dev.off()
```

### Interesting metabolites

```{r,fig.height=7,fig.width=5}
p_emmeans_heatmap
```

```{r,echo=FALSE}
cairo_pdf("../results/vegan/Q1_country/interesting_metab_heatmap.pdf", 
          height = 7, width = 5)
p_emmeans_heatmap
dev.off()
```

```{r}
fsup1b <- p_emmeans_heatmap
save(fsup1b,file="../results/vegan/figures/fsup1b.RData")
```

### Pairwise

```{r,fig.height=8,fig.width=5}
p_pairwise_heatmap
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/final_significance_pairwise.pdf",height = 8,width = 5)
p_pairwise_heatmap
dev.off()
```

## Forest plot

```{r,fig.height=20,fig.width=8}
p_forest
```

```{r}
save(p_forest,file = "../results/vegan/figures/FS1.RData")
```

## Network

```{r,fig.width=11,fig.height=6}
p_net <- ggarrange(p_pos,p_neg,labels = c("A","B"))
p_net
```

```{r,echo=FALSE}
pdf("../results/vegan/Q1_country/final_significance_networks.pdf",height = 6,width = 11)
p_net
dev.off()
```