---
title: "VEGANScreener + metabolomics"
subtitle: "Q5: Associations between PDI groups and the metabolome."
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code explores patterns in the metabolome profiles of vegans, particularly by exploring associations with their dietary records (PDI groups).**


The script is divided into two parts, according to the integration method of metabolome datasets (see preprocessing scripts for details). At first, the **sPCA dataset** is analyzed. This dataset is used for multivariate analysis, PCA visualization, PERMANOVA and ML approach. The **concatenated dataset** of both LCMS and NMR is used for univariate analysis. 

Prediction Goal: Predict metabolome from nutrition	
Question: Given someone’s diet, what metabolites do I expect to see?	
Type of Model: Forward modeling / Regression

Prediction Goal: Predict nutrition from metabolome	
Question: Can I infer what someone eats from their metabolome?
Type of Model: Reverse modeling / Classification or Regression


The main parts of the analysis is composed of:

1. Canonical correlation analysis (CCA) - find linear combinations of **PDI groups** and metabolites that are maximally correlated. - SPCA

A) Predict Metabolome from Nutrition

B) Predict Nutrition from Metabolome

2. Mantel test - provides a global similarity measure - SPCA

3. Linear modeling - Metabolite ~ sex + bmi + age + country + **PDI_group**

4. Network analysis - bipartite graph, calculate degrees for nodes as **PDI groups**, nodes as metabolites. Identify communities in graphs thorugh Louvain method. Edges of the graph correspond to the significant associations (Estimates from LM). 


```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  "../../data/preprocessed/dietary/veganscreener_preprocessed.tsv",
   header = TRUE)  %>%
  dplyr::filter(eq_included_group=="vegan")

lcms_df <- read.table("../../data/preprocessed/metabolomics/lc_ms_preprocessed.tsv",check.names = FALSE)
nmr_df <- read.table("../../data/preprocessed/metabolomics/nmr_preprocessed.tsv",check.names = FALSE)
spca_integrated_df <-  read.table("../../data/preprocessed/metabolomics/spca_integrated_all.tsv",
                                  check.names = FALSE)

pdi_df <- as.data.frame(fread("../../data/preprocessed/dietary/pdi_preprocessed_11.tsv")) %>% column_to_rownames("V1")
```

Remove outliers (other sex, bmi < 15)

```{r}
metadata_df <- metadata_df %>% dplyr::filter(bmi>15,sex != "OTHER")
```

**Get only important samples**

```{r}
# metabolomics
common_samples <- Reduce(intersect, list(rownames(lcms_df),
                                     rownames(nmr_df), 
                                     rownames(spca_integrated_df),
                                     metadata_df$study_id,
                                     rownames(pdi_df)))

lcms_df <- lcms_df[common_samples,]
nmr_df <- nmr_df[common_samples,]
spca_integrated_df <- spca_integrated_df[as.character(common_samples),]
metadata_df <- metadata_df %>% dplyr::filter(study_id %in% common_samples)
pdi_df <- pdi_df[common_samples,]
```


# Descriptive statistics

## Metadata

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(Country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=15}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total")
i <- 1
for (col in c("age","bmi","eq_vegan_dur_total","VS_total")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country")
for (col in c("sex","Country")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20)) +
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count(), aes(x=!!sym(col),y=n+10,label=n)) + 
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
    xlab(name[i]) 
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```

## PDI

Basic plots for visualizing the distribution:

```{r,fig.width=25,warning=FALSE}
plot_list <- list()
name <- colnames(pdi_df)
i <- 1
for (col in name){
  p <- ggplot(pdi_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(pdi_df[,col],na.rm = TRUE),2),"+-", round(sd(pdi_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}

wrap_plots(plot_list)

```

```{r}
pdi_df <- pdi_df %>% rownames_to_column("study_id")
```

# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```

## Visualization

### UMAP

```{r,eval=FALSE}
p1 <- umap_plot(concat_integrated_df,pdi_df,color_col = "hpdi_an",
          neighbors = 40,min_dist = 0.2)

p2 <- umap_plot(concat_integrated_df,pdi_df,color_col = "updi_an",
          neighbors = 40,min_dist = 0.2)

p <- p1 + p2

p
```



```{r,eval=FALSE,echo=FALSE}
pdf("../results/vegan/Q5_pdi/concat_umap.pdf",height = 4,width = 13)
p
dev.off()
```

### PCA

```{r,fig.width=9,fig.height=3,eval=FALSE}
concat_plot_df <- merge(
  concat_integrated_df %>% rownames_to_column("study_id"),
  pdi_df,by="study_id")

p1 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "hpdi_an",
              perform_pca = TRUE)

p1 <- ggtheme(p1)

p2 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "updi_an",
              perform_pca = TRUE)

p2 <- ggtheme(p2)

p <- p1 + p2

p
```

```{r,echo=FALSE,eval=FALSE}
pdf("../results/vegan/Q5_pdi//concat_PCA.pdf",height = 4,width = 12)
p
dev.off()
```

```{r,fig.width=15,fig.height=4,eval=FALSE}
concat_plot_df <- merge(
  concat_integrated_df %>% rownames_to_column("study_id"),
  pdi_df,by="study_id")

p1 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "hpdi_an",
              perform_pca = TRUE,axis = c("PC1","PC2"))

p1 <- ggtheme(p1)

p2 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "hpdi_an",
              perform_pca = TRUE,axis = c("PC2","PC3"))

p2 <- ggtheme(p2)

p3 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "hpdi_an",
              perform_pca = TRUE,axis = c("PC1","PC3"))

p3 <- ggtheme(p3)

p1 + p2 + p3 + 
plot_layout(guides = "collect")
```

```{r,echo=FALSE,eval=FALSE}
pdf("../results/vegan/Q5_pdi//concat_PCA_h_all.pdf",height = 4,width = 15)
p1 + p2 + p3 + 
plot_layout(guides = "collect")
dev.off()
```

```{r,fig.width=15,fig.height=4,eval=FALSE}
concat_plot_df <- merge(
  concat_integrated_df %>% rownames_to_column("study_id"),
  pdi_df,by="study_id")

p1 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "updi_an",
              perform_pca = TRUE,axis = c("PC1","PC2"))

p1 <- ggtheme(p1)

p2 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "updi_an",
              perform_pca = TRUE,axis = c("PC2","PC3"))

p2 <- ggtheme(p2)

p3 <- pca_plot(concat_integrated_df,
              pdi_df,
              color_col = "updi_an",
              perform_pca = TRUE,axis = c("PC1","PC3"))

p3 <- ggtheme(p3)

p1 + p2 + p3 + 
plot_layout(guides = "collect")
```

```{r,echo=FALSE,eval=FALSE}
pdf("../results/vegan/Q5_pdi//concat_PCA_u_all.pdf",height = 4,width = 15)
p1 + p2 + p3 + 
plot_layout(guides = "collect")
dev.off()
```


## Mantel test

The Mantel test measures the correlation between two distance (or dissimilarity) matrices.

In our case:

Dataset 1: PDI profiles (e.g., pdi groups intake per client)
Dataset 2: Metabolomic profiles (e.g., metabolite concentrations per client)

The Mantel test asks: Are clients who eat similarly also similar in their metabolome?

----------------------

```{r}
#nutri <- pdi_df[,c(-19,-20)] %>% column_to_rownames("study_id")
nutri <- pdi_df[,c(-2)] %>% column_to_rownames("study_id")
# delete the zero variance columns
nutri <- nutri[,-which(apply(nutri, 2, var, na.rm = TRUE) == 0)]
```

### Euclidean distance

The Mantel statistic r: 0.02931 is **not** significant (p=0.221).

**Simple Mantel test**

```{r}
# Compute distance matrices (e.g., Euclidean)
dist_nutri <- vegdist(nutri,method = "euclidean")
dist_metabo <-  vegdist(concat_integrated_df,method="euclidean")

# Mantel test
mantel_result <- mantel(dist_nutri, dist_metabo, method = "spearman", permutations = 999)
print(mantel_result)
```

**Partial Mantel test**

The partial mantel is also a correlation test between two matrices, but controlling for the effect of a third distance matrix. This control is done by calculating the correlation between the residuals of each of the two main distance matrices after a linear regression on the third distance matrix.


Even after adjusting for the country effect, the mantel statistic r: 0.02286 remains **non-significant** (p=0.275) 
      
```{r}
country_vec <- metadata_df$Country  # same order as in your data

dist_country <- as.dist(
  outer(country_vec, country_vec, FUN = Vectorize(function(x, y) as.numeric(x != y)))
)

mantel.partial(dist_nutri, dist_metabo, dist_country, permutations = 999,method = "pearson")

```

### Manhattan distance

**Simple Mantel test **

Here, different distance metric is used, to check if the results are consistent when not using euclidean distance. The Mantel statistic r: 0.04442  is still **not** significant (p=0.076). The p-value is, however, closer to the our defined significance level (0.05).
      
```{r}
# Compute distance matrices (e.g., Euclidean)
dist_nutri <- vegdist(nutri,
                      method="manhattan")
dist_metabo <-  vegdist(concat_integrated_df,method="manhattan")

# Mantel test
mantel_result <- mantel(dist_nutri, dist_metabo, method = "spearman", permutations = 999)
print(mantel_result)
```

**Partial Mantel test**

Even after adjusting for the country effect, the mantel statistic r: 0.03869 remains **non-significant** (p=0.16). 

```{r}
country_vec <- metadata_df$Country  # same order as in your data

dist_country <- as.dist(
  outer(country_vec, country_vec, FUN = Vectorize(function(x, y) as.numeric(x != y)))
)

mantel.partial(dist_nutri, dist_metabo, dist_country, permutations = 999,method = "pearson")

```

## Canonical correlation analysis (CCA) 

https://www.karlin.mff.cuni.cz/~maciak/NMST539/cvicenie11.html

Canonical correlation analysis (CCA) is a multidimensional exploratory statistical method which operates on the same principle as the principal component analysis. The main purpose of the canonical correlation approach is the exploration of sample correlations between two sets of quantitative variables observed on the same experimental units. On the other hand PCA method deals with one data set only and it tries to to reduce the overall dimensionality of the dataset using some linear combination of the initial variables.

```{r}
library(CCA)

X <- nutri
Y <- concat_integrated_df

# Run Canonical Correlation Analysis
cca_model <- cc(X, Y)

# Canonical correlations 
cca_model$cor
```

```{r}
plot(cca_model$scores$xscores[,1], cca_model$scores$yscores[,1],
     xlab="U1 (X set)", ylab="V1 (Y set)", main="Canonical Variate 1")
```

**Canonical Correlations barplot**

```{r}
barplot(cca_model$cor, main="Canonical Correlations", ylab="Correlation", col="skyblue")
abline(h=0.5, lty=2, col="red")
```

**Significance**

```{r}
# Significance testing
library(CCP)
sig <- p.asym(rho = cca_model$cor, N = nrow(X), p = ncol(X), q = ncol(Y), tstat = "Wilks")

data.frame(Canonical = 1:length(cca_model$cor),
           Correlation = cca_model$cor,
           p.value = sig$p.value,
           sign=sig$p.value<0.05)
```
The first two canonical variates are significant. To explore these relationship, we will explore circle plot and then individual loadings. 

**Circle plot**

The circle is the unit circle in correlation space (radius = 1). Variables close to the circle edge are strongly correlated with the canonical axes (they contribute most). Variables near the center have weak contributions.

In other words, the direction and distance of a variable vector show:
- Direction → which canonical axis it’s correlated with (positive/negative side).
- Length → strength of the correlation.

Axes:
X-axis: First canonical variate of set X (U₁ = X·a₁)
Y-axis: First canonical variate of set Y (V₁ = Y·b₁)

Each axis represents a linear combination of original variables — one for the nutrient group (X), one for the metabolite group (Y).

Points:
Each point in the plot represents a variable. Variables from X (nutrients) and Y (metabolites) are plotted in a shared space. Their coordinates correspond to correlations (loadings) between each original variable and the canonical variates.

If a nutrient variable (e.g. “fruit_juices”) lies near the metabolite X variable, it means they co-vary strongly in the canonical dimension.
If two variables are close together, they are correlated. If they are on opposite sides of the origin, they are negatively correlated. Variables near the origin contribute little to that canonical function.

```{r,fig.width=20,fig.height=20}
plt.cc(cca_model, var.label = TRUE,type = "v")
```

**Loadings**

*First variate*

```{r}
# PDI groups
cca_model$xcoef[,c(1)][order(abs(cca_model$xcoef[,c(1)]),decreasing = TRUE)]
```

```{r}
# metabolites groups
cca_model$ycoef[,c(1)][order(abs(cca_model$ycoef[,c(1)]),decreasing = TRUE)][1:20]
```
*Second variate*

```{r}
# PDI groups
cca_model$xcoef[,c(2)][order(abs(cca_model$xcoef[,c(2)]),decreasing = TRUE)]
```

```{r}
# metabolites groups
cca_model$ycoef[,c(2)][order(abs(cca_model$ycoef[,c(2)]),decreasing = TRUE)][1:20]
```

## Network analysis 

We want to model each metabolite as a function of sex, BMI, age, country, and each food group. Keep the estimated effect (and significance) of each food group on each metabolite.Then, we will use those significant effects to construct a bipartite network, where:

- one node type = metabolites

- other node type = food groups

- edges = significant relationships from the regression models.

Then we’ll analyze and visualize that network (degrees, communities, etc.).

```{r}
#nutri <- pdi_df[,c(-19,-20)] %>% column_to_rownames("study_id")
nutri <- pdi_df[,c(-2)] %>% column_to_rownames("study_id")
# delete the zero variance columns
nutri <- nutri[,-which(apply(nutri, 2, var, na.rm = TRUE) == 0)] %>%
  rownames_to_column("study_id")

metadata_for_lm <- merge(nutri,metadata_df,by="study_id",all=TRUE)
concat_for_uni <- merge(
  concat_integrated_df %>% rownames_to_column("study_id"),
  metadata_for_lm,
  by="study_id",all = TRUE)
```

### LM

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "pdi_groups"))

results_df <- do.call(rbind, results_list)

results_df <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

pdi_effects <- results_df %>% dplyr::filter(term %in% colnames(nutri))

concat_significant_cat <-  pdi_effects %>%
  filter(p_adj < 0.05)

concat_significant_cat
```

### Graph construction

**Edge list**

```{r}
edges <- pdi_effects %>%
  dplyr::filter(p_adj < 0.05) %>%
  dplyr::select(from = term, to = Metabolite, 
         estimate = estimate) %>%
  dplyr::mutate(from=as.character(from),
         weight=abs(estimate))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
pdis <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(pdis, metabs),
  type = c(rep(TRUE, length(pdis)), rep(FALSE, length(metabs))) # TRUE = tertile, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**


```{r,fig.height=6,fig.width=6}
set.seed(123)
pdi_colors <- sample(colors(), sum(nodes$type))
names(pdi_colors) <- nodes$name[nodes$type]

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     pdi_colors[V(g)$name],
                     "gray70")

set.seed(123)
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = estimate > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue", "lightpink"), labels = c("<0", ">0")) +
  geom_node_point(aes(shape=type,color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +  # node colors
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  guides(color = "none")

```

## glmnet

### PDI groups -> metabolites

```{r}
X <- as.matrix(nutri %>% column_to_rownames("study_id"))
Y <- as.matrix(concat_integrated_df)

model_glmnet <- make_glmnet_model(X,Y)
```

To consider model to be well fitted, we would expect following values:
R²	> 0.5 (ideally closer to 0.7–0.8)
Correlation (r)	> 0.6
RMSE	As low as possible (relative to Y scale)

```{r}
results_df_model <- data.frame(
  r2=colMeans(do.call(rbind, model_glmnet$valid_performance$r2)),
  corr=colMeans(do.call(rbind, model_glmnet$valid_performance$corr)),
  rmse=colMeans(do.call(rbind, model_glmnet$valid_performance$rmse))
)

results_df_model

```
### metabolites -> PDI groups

```{r}
Y <- as.matrix(nutri %>% column_to_rownames("study_id"))
X <- as.matrix(concat_integrated_df)

model_glmnet <- make_glmnet_model(X,Y)
```

To consider model to be well fitted, we would expect following values:
R²	> 0.5 (ideally closer to 0.7–0.8)
Correlation (r)	> 0.6
RMSE	As low as possible (relative to Y scale)

```{r}
results_df_model <- data.frame(
  r2=colMeans(do.call(rbind, model_glmnet$valid_performance$r2)),
  corr=colMeans(do.call(rbind, model_glmnet$valid_performance$corr)),
  rmse=colMeans(do.call(rbind, model_glmnet$valid_performance$rmse))
)

results_df_model

```

