---
title: "VEGANScreener + metabolomics"
subtitle: "Q0: Associations of metabolome profile with various external variables."
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code is used to explore patterns in the metabolome profiles of vegans.**

In the very first part of analysis (Q0), the general plots generated from the studied cohort are shown:

- descriptive statistics

- PCA, color-coded according to different external variables

- correlations of individual PCs with different external variables


```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  file = "../../data/preprocessed/metadata_preprocessed_v3.tsv", 
  header = TRUE)

spca_integrated_df <-  read.table(
  file = "../../data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",
  check.names = FALSE)

pdi_df <- read.table(
  file = "../../data/preprocessed/dietary/pdi/pdi_avg_residuals_scaled_v3.tsv",
  check.names = FALSE,
  header = TRUE) %>% 
  dplyr::select(study_id,pdi_an_avg,hpdi_an_avg,updi_an_avg)

metadata_df <- merge(metadata_df,pdi_df,by = "study_id",all.x = TRUE)
```

**Scaling** - when using columns from metadata as covariates, they need to be scaled, so mean is 0 and SD is 1.

```{r}
metadata_scaled_df <- metadata_df
metadata_scaled_df[,-c(1,3,5,6,8)] <- apply(metadata_df[,-c(1,3,5,6,8)],2,scale)
metadata_scaled_df <- metadata_scaled_df %>%
  mutate(sex=as.factor(sex),
         country=as.factor(country),
         group=as.factor(group),
         diet_supplements=as.factor(diet_supplements))
```

# Descriptive statistics

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=15,fig.height=10,warning=FALSE}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total",
          "VS_total_plus","VS_total_minus",
          "UPF [%]",
          "PDI","hPDI","uPDI")
i <- 1
for (col in c("age","bmi","vegan_duration","VS_total","UPF_perc_avg",
              "pdi_an_avg","hpdi_an_avg","updi_an_avg")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country","Supplementation")
for (col in c("sex","country","diet_supplements")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          axis.text.x=element_text(angle = 45,hjust = 1)) +
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
    xlab(name[i]) +
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count() %>%
              dplyr::mutate(n=n+10), aes(x=!!sym(col),y=n,label=n))
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```

# Table 1

Generating Table 1 (Baseline characteristics of the study participants.)

```{r}
# Define variables
continuous_vars <- c("age", "bmi", "vegan_duration", 
                     "VS_total","VS_total_plus","VS_total_minus",
                     "pdi_an_avg","hpdi_an_avg","updi_an_avg", "UPF_perc_avg")

categorical_vars <- c("sex", "diet_supplements")  # add more if needed

# Optional: stratify by country
strata_var <- "country"  # set to NULL if not stratifying

# Overall (non-stratified)
table_overall <- CreateTableOne(vars = c(continuous_vars, categorical_vars),
                                data = metadata_df,
                                factorVars = categorical_vars,
                                test = FALSE)

# Stratified by country
table_strat <- CreateTableOne(vars = c(continuous_vars, categorical_vars),
                              strata = strata_var,
                              data = metadata_df,
                              factorVars = categorical_vars,
                              test = FALSE)

# Combine into one table for display
# Extract as data frames
df_overall <- print(table_overall, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
df_strat <- print(table_strat, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)

# Combine overall + stratified
table_final <- cbind(Overall = df_overall, df_strat)

# Display 
knitr::kable(table_final, caption = "Table 1. Baseline characteristics of study participants")
```

```{r}
write.xlsx(table_final,"../results/vegan/Table1.xlsx",rowNames=TRUE)
```

```{r}
skim(metadata_df)
```

# Metabolites

```{r}
lcms_df <- read.table(
  file = "../../data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",
  check.names = FALSE)

nmr_df <- read.table(
  file = "../../data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",
  check.names = FALSE)

common_samples <- Reduce(intersect, 
                         list(rownames(lcms_df),
                              rownames(nmr_df),
                              metadata_df$study_id))

lcms_df <- lcms_df[common_samples,]
nmr_df <- nmr_df[common_samples,]

dim(lcms_df)
dim(nmr_df)
```

# PERMANOVA

Testing each individual variable to see how much variance they explain in serum metabolome profile. 

```{r}
vars_to_test <- c("sex","bmi", "age", "VS_total","VS_total_plus","VS_total_minus", "vegan_duration","country", "diet_supplements","UPF_perc_avg","pdi_an_avg","hpdi_an_avg",
                  "updi_an_avg")

adonis_result_all <- data.frame(Var=NA,
                                Df=NA,
                                SumOfSqs=NA,
                                R2=NA,
                                `F`=NA,
                                `Pr(>F)`=NA,
                                check.names = FALSE)
for (var in vars_to_test){
  metadata_sub_df <- metadata_scaled_df %>% dplyr::select(all_of(c("study_id",var))) %>% drop_na()
  sub_spca <- spca_integrated_df[as.character(metadata_sub_df$study_id),]
  
  # Create Euclidean distance matrix
  dist_mat <- vegdist(sub_spca, 
                    method = "euclidean")
  
  # Test 
  set.seed(123)
  adonis_result <- adonis2(
  as.formula(paste0("dist_mat ~ ", var)),
  data = metadata_sub_df , 
  permutations = 999,
  by="margin")

  adonis_result_all <- rbind(adonis_result_all,as.data.frame(adonis_result)[1,] %>% 
                               rownames_to_column("Var"))
}


adonis_result_all$adjusted_p <- p.adjust(adonis_result_all$`Pr(>F)`,method = "BH")

adonis_result_all <- adonis_result_all[-1,]
print(adonis_result_all)
```

```{r,fig.height=4,fig.width=9}
# Significance grouping
adonis_result_all$significance <- ifelse(
  adonis_result_all$adjusted_p < 0.05, "Significant", "Not significant"
)

# Order by R2 ascending (so largest appears at top in coord_flip)
adonis_result_all$Var <- factor(
  adonis_result_all$Var,
  levels = adonis_result_all$Var[order(adonis_result_all$R2)]
)

# Plot
p <- ggplot(adonis_result_all, aes(x = Var, y = R2 * 100, fill = significance)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_fill_manual(
    values = c("Significant" = "#1D9E75", "Not significant" = "#B4B2A9"),
    name = "PERMANOVA"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = NULL,
    y = "Variance explained (R², %)"
  ) 

p <- ggtheme(p) + 
  theme(axis.text.x = element_text(face="bold",size=15),
        axis.text.y = element_text(face="bold",size=15),
        panel.grid.major.x = element_line(color = "grey90", linewidth = 0.4),
        title = element_text(face="bold",size=15),
        legend.position = "bottom") + scale_x_discrete(labels = c("country" = "Country",
                                  "VS_total" = "VEGANScreener score",
                                  "sex" = "Sex",
                                  "UPF_perc_avg" = "UPF",
                                  "age" = "Age",
                                  "bmi" = "BMI",
                                  "diet_supplements" = "Supplementation",
                                  "vegan_duration" = "Vegan duration",
                                  "pdi_an_avg" = "PDI"
                                  ))

p
```

```{r}
f <- p
save(f,file="../results/vegan/figures/F2a_permanovaVar.RData")
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q0/R2_permanova.pdf",height = 4,width = 9)
p
dev.off()
```

# Correlations with PCs

The SPCA dataset was created by concatenating LCMS and NMR dataset and applying Mixomics::spca() function. See preprocessing script for details. 

mixOmics::spca() performs **Sparse Principal Component Analysis**, which is similar to regular PCA but adds a sparsity constraint so that only a subset of variables contributes to each component.
This sparsity is controlled by penalization, making interpretation easier because irrelevant or noisy variables get loadings of zero.

```{r}
metadata_analysis_df <- metadata_df %>% 
  dplyr::select(study_id, VS_total,VS_total_plus,VS_total_minus, age, bmi, sex, country, 
                vegan_duration, diet_supplements,UPF_perc_avg,pdi_an_avg,hpdi_an_avg,updi_an_avg) %>%
  column_to_rownames("study_id")

cor_results <- corr_with_pcs(spca_integrated_df,metadata_analysis_df)

### Filter significant correlations (p < 0.05)
df_sig <- cor_results %>% 
  filter(is.finite(rho), pvalue < 0.05)
df_sig$var1 <- factor(df_sig$var1, levels = names(spca_integrated_df))
```

Generated plot below shows Spearman correlations of individual PCs from sPCA metabolome dataset with various external variables. The most strongest associations can be seen with **sex**, individual **countries**, **VS_total** and **bmi**. These variables, will be explored in more details in other scripts.  

The strongest signal can be seen in individual **countries**, **VS_total** and **sex**. 

```{r,fig.height=7,fig.width=12}
meta_order <- df_sig %>%
  group_by(var2) %>%
  dplyr::summarize(max_abs_rho = max(sum(abs(rho)), na.rm = TRUE)) %>%
  arrange((max_abs_rho)) %>%
  pull(var2)

meta_order <- c(levels(adonis_result_all$Var)[-length(adonis_result_all$Var)],
                "country_ES","country_DE","country_CH","country_BE",
                "country_CZ")

df_sig$var2 <- factor(df_sig$var2, levels = meta_order)

p <- ggplot(df_sig, aes(x = var1, y = var2, fill = rho)) +
  geom_point(shape = 21,size=8,color="white") +
  scale_fill_gradient2(
    
    low = "darkblue",mid = "white",high = "darkred", midpoint = 0, name = "Spearman correlation",limits = c(-1, 1)) +
  labs(
    x = "",
    y = "",
    title = "Key Variable Associations with Metabolome PCs") 

p <- ggtheme(p) + 
  theme(axis.text.x = element_text(face="bold",size=15,angle=45,hjust=1),
        axis.text.y = element_text(face="bold",size=15),
        title = element_text(face="bold",size=15),
        legend.position = "bottom")

p
```

```{r}
f <- p
save(f,file="../results/vegan/figures/F2b_PcaCorr.RData")
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q0/pca_corr_main_covariates.pdf",height = 6,width = 9)
p
dev.off()
```


# PCA and colored metadata

The strongest pattern is seen with **sex** variable, as was evident in the correlation plot above. 

**PC1 vs PC2**

```{r,fig.width=20,fig.height=15}
plot_list <- list()
for (col in c("sex","country","age","bmi","VS_total","vegan_duration","UPF_perc_avg","pdi_an_avg")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE) 

  p <- ggtheme(p)
  
  plot_list[[length(plot_list)+1]] <- p 
}


a12 <- wrap_plots(plot_list)
```

**PC1 vs PC3**

```{r,fig.width=20,fig.height=15}
plot_list <- list()
for (col in c("sex","country","age","bmi","VS_total","vegan_duration","UPF_perc_avg","pdi_an_avg")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE,
              axis=c("PC1","PC3")) 

  p <- ggtheme(p)
  
  plot_list[[length(plot_list)+1]] <- p 
}


a13 <- wrap_plots(plot_list)
```

**PC2 vs PC3**

```{r,fig.width=20,fig.height=15}
plot_list <- list()
for (col in c("sex","country","age","bmi","VS_total","vegan_duration","UPF_perc_avg","pdi_an_avg")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE,
              axis=c("PC2","PC3")) 

  p <- ggtheme(p)
  plot_list[[length(plot_list)+1]] <- p 
}


a23 <- wrap_plots(plot_list)
```

```{r,fig.width=10,fig.height=5}
a12
a13
a23
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q0/pca_main_covariates.pdf",height = 5,width = 10)
a12
a13
a23
dev.off()
```