---
title: "VEGANScreener + metabolomics"
subtitle: "Q2: Associations between VS_total and the metabolome in vegans."
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

**This code explores patterns in the metabolome profiles of vegans, particularly by linking them to the VEGANScreener score (VS_total).**


The script is divided into two parts, according to the integration method of metabolome datasets (see preprocessing scripts for details). At first, the **sPCA dataset** is analyzed. This dataset is used for multivariate analysis, PCA visualization, PERMANOVA and ML approach. The **concatenated dataset** of both LCMS and NMR is used for univariate analysis. 

The main parts of the analysis is composed of:

1.  Global visualization - 3 components color-coding **VS_total**

2.  PERMANOVA - how much of the variation can be explained by various external variables:
  dist(Metabolome) \~ sex + bmi + age + veganDuration + country***VS_total**

3.  Univariate analysis (feature-wise model) - linear modelling of metabolite using various external variables:
  Metabolite \~  sex + bmi + age + veganDuration + Country***VS_total**
  

4.  RF classification - the model is trained and validated through bootstrapping. The RF model is used to distinguish between participants with **1st and 3rd tercil of VS_total**.
a) ML model on original sPCA dataset
b) ML model on sPCA dataset after regressing out the covariates.

```{r,echo=FALSE}
source("custom_functions_vegans.R")
```

# Import

```{r}
metadata_df <- read.table(
  "../../data/preprocessed/dietary/veganscreener_preprocessed.tsv",
   header = TRUE)  %>%
  dplyr::filter(eq_included_group=="vegan")

lcms_df <- read.table("../../data/preprocessed/metabolomics/lc_ms_preprocessed.tsv",check.names = FALSE)
nmr_df <- read.table("../../data/preprocessed/metabolomics/nmr_preprocessed.tsv",check.names = FALSE)
spca_integrated_df <-  read.table("../../data/preprocessed/metabolomics/spca_integrated_all.tsv",
                                  check.names = FALSE)
```

Remove outliers (other sex, bmi < 15)

```{r}
metadata_df <- metadata_df %>% dplyr::filter(bmi>15,sex != "OTHER")
```

**Get only important samples**

```{r}
# metabolomics
common_samples <- Reduce(intersect, list(rownames(lcms_df),
                                     rownames(nmr_df), 
                                     rownames(spca_integrated_df),
                                     metadata_df$study_id))

lcms_df <- lcms_df[common_samples,]
nmr_df <- nmr_df[common_samples,]
spca_integrated_df <- spca_integrated_df[as.character(common_samples),]
metadata_df <- metadata_df %>% dplyr::filter(study_id %in% common_samples)

```

# Descriptive statistics

Summary of counts per diet group, country and sex:

```{r}
metadata_df %>% dplyr::group_by(Country,sex) %>% count()
```

Basic plots for visualizing the distribution of selected variables:

```{r,fig.width=15}
plot_list <- list()
name <- c("Age (years)","BMI","Vegan diet duration (years)","VS_total")
i <- 1
for (col in c("age","bmi","eq_vegan_dur_total","VS_total")){
  p <- ggplot(metadata_df) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(metadata_df[,col],na.rm = TRUE),2),"+-", round(sd(metadata_df[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}
i <- 1
name <- c("Sex","Country")
for (col in c("sex","Country")){
  p <- ggplot(metadata_df) +
        theme_minimal() + 
     theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20)) +
  geom_text(data=metadata_df %>% 
              dplyr::group_by(!!sym(col)) %>% count(), aes(x=!!sym(col),y=n+10,label=n)) + 
    stat_count(aes(x=!!sym(col)),fill="darkblue") + 
    xlab(name[i]) 
  
  plot_list[[length(plot_list)+1]] <- p
                 
  i <- i+1                           
}

wrap_plots(plot_list)

```
# SPCA

The SPCA dataset was created by concatenating LCMS and NMR dataset and applying Mixomics::spca() function. See preprocessing script for details. 

mixOmics::spca() performs **Sparse Principal Component Analysis**, which is like regular PCA but adds a sparsity constraint so that only a subset of variables contributes to each component.
This sparsity is controlled by penalization, making interpretation easier because irrelevant or noisy variables get loadings of zero.

## Visualization

### UMAP

```{r}
p <- umap_plot(spca_integrated_df,metadata_df,color_col = "VS_total",
          neighbors = 40,min_dist = 0.2)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/umap.pdf",height = 4,width = 6)
p
dev.off()
```

### PCA

```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE)

p <- ggtheme(p)

p
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/PCA.pdf",height = 4,width = 6)
p
dev.off()
```

```{r,fig.width=15,fig.height=4}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p1 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC1","PC2"))

p1 <- ggtheme(p1)

p2 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC2","PC3"))

p2 <- ggtheme(p2)

p3 <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,axis = c("PC1","PC3"))

p3 <- ggtheme(p3)

p1 + p2 + p3 + 
plot_layout(guides = "collect")
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/PCA_all.pdf",height = 4,width = 15)
p1 + p2 + p3 + 
plot_layout(guides = "collect")
dev.off()
```

**PC1 vs PC2**

```{r,fig.width=10}
plot_list <- list()
for (col in c("Country","sex","age","bmi","VS_total","eq_vegan_dur_total")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE) 

  p <- ggtheme(p)
  
  plot_list[[length(plot_list)+1]] <- p 
}


a12 <- wrap_plots(plot_list)
```

**PC1 vs PC3**

```{r,fig.width=10}
plot_list <- list()
for (col in c("Country","sex","age","bmi","VS_total","eq_vegan_dur_total")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE,
              axis=c("PC1","PC3")) 

  p <- ggtheme(p)
  
  plot_list[[length(plot_list)+1]] <- p 
}


a13 <- wrap_plots(plot_list)
```

**PC2 vs PC3**

```{r,fig.width=10}
plot_list <- list()
for (col in c("Country","sex","age","bmi","VS_total","eq_vegan_dur_total")){
  p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = col,
              perform_pca = FALSE,
              axis=c("PC2","PC3")) 

  p <- ggtheme(p)
  plot_list[[length(plot_list)+1]] <- p 
}


a23 <- wrap_plots(plot_list)
```

```{r,fig.width=10,fig.height=5}
a12
a13
a23
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_main_covariates.pdf",height = 5,width = 10)
a12
a13
a23
dev.off()
```

### Boxplots

```{r,fig.height=15,fig.width=20}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE) 

spca_for_uni <- spca_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  )

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = "tertile_label",y = col,fill="tertile_label")) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) +
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p)
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_boxplots.pdf",height = 15,width = 20)
a
dev.off()
```


### Heatmap

```{r}
metadata_df <- metadata_df %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  ) %>%
  dplyr::select(-tertile_cut)
```

```{r}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

p_df <- spca_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,tertile_label) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "tertile_label"=VS_colors
)
names(ann_colors[["tertile_label"]]) <- names(VS_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```

**Tertile averages**

```{r}
metadata_df %>%
  select(VS_total, tertile_label) %>%
  group_by(tertile_label) %>%
  summarise(mean_VS = mean(VS_total, na.rm = TRUE))
```

## PERMANOVA

The test show how much of the variation in metabolomics is explained by: - sex - age - bmi -vegan duration - Country - **VS_total**

Variables were added sequentially (Type I SS), so each p-value and R² reflects a partial effect, after adjusting for those above. This model accounts for variation due to sex, age, vegan diet duration, BMI and country of origin. Then accounts for the main effects of VS_total. Then tests whether the effect of VS_total varies by country of origin (interaction).

----------------------

Country explains **2.47%** of variation after adjusting for sex, age, bmi,country and duration of the diet. This result is statistically significant (p<0.001).

However, the country of origin is still explaining the most of the variation ~ **10%**. The interaction effect is not significant, which suggest the differences in VS_total is not inconsistent throughout the countries.

```{r}
# Create Euclidean distance matrix
dist_mat <- vegdist(spca_integrated_df, 
                    method = "euclidean")

# Test whether vegan vs omnivore differ in metabolomics
adonis_result <- adonis2(
  dist_mat ~ sex + bmi + age + eq_vegan_dur_total + Country*VS_total,
  data = metadata_df , 
  permutations = 999,
  by="terms")

print(adonis_result)
```
## Univariate analysis

Here, we will plot density plot to gather a basic understanding of individual metabolites (in this case PCs) when comparing the VS_total. Next, we will test the differences in VS total usinglinear model while adjusting for covariates.

### Density plot

```{r,fig.height=10,fig.width=20}
spca_for_uni <- merge(spca_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE) 

spca_for_uni <- spca_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  ) 

plot_list <- list()
for (col in colnames(spca_integrated_df)){
  p <- ggplot(spca_for_uni, 
              aes_string(x = col,fill="tertile_label")) +
    geom_density(alpha = 0.5) +
    ggtitle(col) + 
    theme_minimal()
  
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list)

a
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/pca_density.pdf",height = 10,width = 20)
a
dev.off()
```

### Metabolite-wise model

Calling glm to explore which metabolites (PCs) are significantly related to **VS_total** while adjusting for age, sex, vegan duration and country.

```{r}
results_list <- lapply(colnames(spca_integrated_df), 
                       function(x) fit_glm(x,spca_for_uni,
                                           tested_variable = "VS_total"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results - **PC6**, **PC12**, **PC19**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total")

```

There is minor negative association between **PC6** and **VS total**, which corresponds to do correlation coefficient of -0.39.

```{r}
cor_val <- cor(spca_for_uni$PC6,
               spca_for_uni$VS_total,
               method = "spearman")

ggplot(spca_for_uni, 
       aes_string(x = "PC6", y = "VS_total",color="tertile_label")) +
       geom_point() + 
       theme_minimal() + 
  ggtitle(paste("Rho =",round(cor_val,2)))
```


There is minor negative association between **PC12** and **VS total**, which corresponds to do correlation coefficient of -0.26.

```{r}
cor_val <- cor(spca_for_uni$PC12,
               spca_for_uni$VS_total,
               method = "spearman")

ggplot(spca_for_uni, 
       aes_string(x = "PC12", y = "VS_total",color="tertile_label")) +
       geom_point() + 
       theme_minimal() + 
  ggtitle(paste("Rho =",round(cor_val,2)))
```

There is no association between **PC19** and **VS total**.

```{r}
cor_val <- cor(spca_for_uni$PC19,
               spca_for_uni$VS_total,
               method = "spearman")

ggplot(spca_for_uni, 
       aes_string(x = "PC19", y = "VS_total",color="tertile_label")) +
       geom_point() + 
       theme_minimal() + 
  ggtitle(paste("Rho =",round(cor_val,2)))
```

Check the interaction effect - **none** - no evidence that the effect of VS_total differs by country.

```{r}
df <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(str_ends(variable, ":VS_total"))

df

pc_to_be_removed <- df$metabolite
```

**Volcano plot**

```{r}
vp <- volcano_plot(results_df,comparison = "VS_total")
vp
```

**All correlations**

```{r}
apply(spca_integrated_df,2, function(x) cor(x,spca_for_uni$VS_total,method="spearman"))
```

## RF model

### Original model

```{r}
sub_metadata_df <- metadata_df %>%
dplyr::filter(
  tertile_label %in% c("1st_VStertile","3rd_VStertile"))

sub_spca_integrated_df <- spca_integrated_df[as.character(sub_metadata_df$study_id),]
```

```{r,fig.height=5,fig.width=5}
model_name <- "VS_tertile_vegan" 

prepared_df <- model_preparation(sub_spca_integrated_df,
                                 sub_metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                            outcome="Group",
                            N = 500,
                           file = model_name,
                            reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("low VS_total","high VS_total"),CI = TRUE) 

roc_c <- ggtheme(roc_c) + 
  theme(legend.position="none",title = element_text(size=8))

roc_c
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/orig_rf_vstotal.pdf",width = 6,height = 5)
roc_c
dev.off()
```

### Regressed out model

```{r,fig.height=5,fig.width=5}
# Covariates to adjust for
covars <- c("sex","bmi", "age", "Country", "eq_vegan_dur_total")
integrated_df <- sub_spca_integrated_df
all(rownames(integrated_df) == sub_metadata_df$study_id)
adjusted_data <- adjust_for_covars(integrated_df, covars, sub_metadata_df)

model_name <- "adjusted_vstotal_model_vegan" 

prepared_df <- model_preparation(adjusted_data,
                                 sub_metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 500,
                          file = model_name,
                          reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model, 
                   group = c("low VS_total","high VS_total"),CI = TRUE) 

roc_c <- ggtheme(roc_c) + 
  theme(legend.position="none",title = element_text(size=8))

roc_c

```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/adjusted_rf_vstotal.pdf",width = 6,height = 5)
roc_c
dev.off()
```

## RF importance

### Original model

**PC6** is the most important for RF model. Afterwards, there is a significant drop in the importance, which highligts the importance of the **PC6**.

```{r,fig.width=10}
load("../intermediate_files/models/VS_tertile_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

### Adjusted model

Similarly as above, the most important is **PC6**, with significant drop of importance after this PC. 

```{r,fig.width=10}
load("../intermediate_files/models/adjusted_vstotal_model_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("PC") %>%
  `colnames<-`(c("PC","Importance")) 

var_imp <- var_imp %>% dplyr::mutate(PC=factor(PC,levels = PC[order(var_imp$Importance,decreasing = TRUE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=PC,y=Importance))
```

**Visualization**

To explore this important PC more, its visualization is shown below. 


```{r,fig.width=4,fig.height=3}
spca_plot_df <- merge(
  spca_integrated_df %>% rownames_to_column("study_id"),
  metadata_df,by="study_id")

p <- pca_plot(spca_integrated_df,
              metadata_df,
              color_col = "VS_total",
              perform_pca = FALSE,
              axis = c("PC6","PC9"))

p <- ggtheme(p)

p
```


```{r}
p <- ggplot(spca_for_uni, 
             aes_string(x = "tertile_label",y = "PC6",fill="tertile_label")) +
    geom_boxplot() +
    ggtitle("PC6") + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) + 
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
p <- ggtheme(p) 

p

```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/important_pca_boxplots.pdf",height = 4,width = 8)
p
dev.off()
```


## SPCA loadings {#SPCA_loadings}

Here, we will explore the loadings of the most important PC (**PC6**)

```{r,collapse=TRUE}
load("../results/preprocessing/spca_result_all.RData")
df <- data.frame (
  Metabolite=names(spca_result$loadings$X[,6]),
  Loading=spca_result$loadings$X[,6])

write.table(df,sep="\t","../results/vegan/Q2_VStotal/loadings_PC6.tsv")
```

This plot shows the loading of individual metabolites to **PC6**, ordered by its absolute value.

```{r,fig.height=10,fig.width=5,collapse=TRUE}
ggplot(df %>% 
         dplyr::mutate(
           Metabolite=factor(Metabolite,
                             levels = Metabolite[order(abs(Loading),decreasing = FALSE)])
           )) + 
  geom_col(aes(x=Metabolite,y=Loading)) + 
  theme(axis.text.x = element_text(angle=90,hjust = 1)) + 
  coord_flip()

spca_loadings <- df$Metabolite[abs(df$Loading)>=0.1]
```
# Concatenated dataset

```{r}
concat_integrated_df <- cbind(lcms_df,nmr_df)
```

## Visualization

### UMAP

```{r}
p <- umap_plot(concat_integrated_df,metadata_df,color_col = "VS_total",
          neighbors = 10,min_dist = 0.1)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_umap.pdf",height = 4,width = 6)
p
dev.off()
```

### Boxplots

```{r,fig.height=40,fig.width=20}
concat_for_uni <- merge(concat_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE) 

concat_for_uni <- concat_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  )

plot_list <- list()
for (col in colnames(concat_integrated_df)){
  p <- ggplot(concat_for_uni, 
              aes(x = tertile_label,
                  y = !!sym(col),
                  fill=tertile_label)) +
    geom_boxplot() +
    ggtitle(col) + 
    theme_minimal() + 
    scale_fill_manual(values = c("#D73027","orange","#1A9850")) +
    geom_signif(comparisons = 
                  combn(as.character(unique(spca_for_uni$tertile_label)),
                        m=2, simplify = FALSE),
                map_signif_level = TRUE, step_increase = 0.1,textsize = 2)
  
  p <- ggtheme(p) + 
    theme(axis.text.x = element_text(angle=45,hjust = 1))
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list,guides = "collect")

a
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_boxplots.pdf",height = 35,width = 20)
a
dev.off()
```

### Heatmap

```{r}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

p_df <- concat_integrated_df %>% t()

ann_df <- metadata_df %>% dplyr::select(study_id,tertile_label) %>%
  column_to_rownames("study_id")

ann_colors <- list(
  "tertile_label"=VS_colors
)
names(ann_colors[["tertile_label"]]) <- names(VS_colors)

p <- pheatmap(p_df,cluster_rows = FALSE,show_colnames = FALSE,
              show_rownames = FALSE,
              annotation_col = ann_df,
              annotation_colors = ann_colors,silent = TRUE)

p
```

```{r,eval=TRUE,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_heatmap.pdf",height = 10,width = 10)
p
dev.off()
```


## Univariate analysis

Here, we will create density plot to gather a basic understanding of individual metabolites when comparing diet group. Next, we will test the differences between diets and VS total using generalize linear model while adjusting for covariates.

### Density plot

```{r,fig.height=40,fig.width=40}
concat_for_uni <- merge(concat_integrated_df %>% rownames_to_column("study_id"),
                      metadata_df,by="study_id",all = TRUE)

concat_for_uni <- concat_for_uni %>%
  mutate(
    tertile_cut = ntile(VS_total, 3), # assign 1, 2, or 3 (tertile)
    tertile_label = case_when(                          # label tertiles
      tertile_cut == 1 ~ "1st_VStertile",
      tertile_cut == 2 ~ "2nd_VStertile",
      tertile_cut == 3 ~ "3rd_VStertile"
    )
  ) 

plot_list <- list()
for (col in colnames(concat_integrated_df)){
  p <- ggplot(concat_for_uni, 
              aes(x = !!sym(col), fill = tertile_label)) +
    geom_density(alpha = 0.5) +
    ggtitle(col) + 
    theme_minimal()
  
  plot_list[[length(plot_list) + 1]] <- p
}


a <- wrap_plots(plot_list)
a
```

```{r,echo=FALSE}
pdf("../Results/vegan/Q2_VStotal/density_diet.pdf",height = 40,width = 40)
a
dev.off()
```

### Metabolite-wise model

#### Continuous analysis

Call glm to explore which metabolites are significantly related to VS score while adjusting for age, sex, Country and dietary group

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_glm(x,concat_for_uni,
                                           tested_variable = "VS_total"))
results_df <- do.call(rbind, results_list)

results_df$p_adjusted <- p.adjust(results_df$Pr...t..,method = "BH")
```

Look at significant results - **9 metabolites**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total")

```

Check the interaction effect - **none**

```{r}
df <- results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(str_ends(variable, ":VS_total"))

df

pc_to_be_removed <- df$metabolite
```

```{r,collapse=TRUE}
concat_significant_df <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))

concat_univariate_results <- results_df

concat_significant <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))$metabolite
```

**Volcano plot**

```{r}
vp <- volcano_plot(results_df,comparison = "VS_total")
vp
```

**Correlation**

```{r}
sign <- (results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="VS_total"))$metabolite

```

```{r,fig.height=5,fig.width=5}
df_for_plot <- concat_integrated_df %>% dplyr::select(sign)
corr_object <- cor(concat_for_uni$VS_total,df_for_plot)
corrplot::corrplot(corr_object,method = "number")
```


**Vegan duration**

```{r}
results_df %>%
  dplyr::filter(p_adjusted<0.05) %>%
  dplyr::filter(variable=="eq_vegan_dur_total")

```

#### Network analysis

**LM**

```{r}
results_list <- lapply(colnames(concat_integrated_df), 
                       function(x) fit_netglm(x,concat_for_uni,
                                           tested_variable = "tertile_label"))
results_df <- do.call(rbind, results_list)

vs_effects <- results_df %>%
  dplyr::mutate(p_adj = p.adjust(`p.value`, method = "BH"))

concat_significant_cat <-  vs_effects %>%
  filter(p_adj < 0.05)
```

**Edge list**

```{r}
edges <- vs_effects %>%
  filter(p_adj < 0.05) %>%
  select(from = tertile_label, to = metabolite, 
         emmean = emmean) %>%
  mutate(from=as.character(from),
         weight=abs(emmean))

edges$from <- as.character(edges$from)
edges$to   <- as.character(edges$to)
```

**Nodes table**

```{r}
tertiles <- as.character(unique(edges$from))
metabs <- as.character(unique(edges$to))

nodes <- data.frame(
  name = c(tertiles, metabs),
  type = c(rep(TRUE, length(tertiles)), rep(FALSE, length(metabs))) # TRUE = tertile, FALSE = metabolite
)
```

**Build igraph object**

```{r}
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
```

**Visualize**

```{r,fig.height=6,fig.width=6}
VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

# Assign colors to nodes: countries get their color, metabolites gray
V(g)$color <- ifelse(V(g)$type,
                     VS_colors[V(g)$name],
                     "gray70")

set.seed(123)
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue", "lightpink"), labels = c("<0", ">0")) +
  geom_node_point(aes(shape=type,color = name), size = 6) +
  scale_color_manual(values = setNames(V(g)$color, V(g)$name)) +  # node colors
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  guides(color = "none")

```

##### Louvain clustering

**Based on positive associations**

```{r}
edges_pos <- edges %>%
  filter(emmean > 0)  # keep only positive associations

g_pos <- graph_from_data_frame(
  d = edges_pos, 
  vertices = nodes, 
  directed = FALSE)

VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

# Assign colors to nodes: countries get their color, metabolites gray
V(g_pos)$color <- ifelse(V(g_pos)$type,
                     VS_colors[V(g_pos)$name],
                     "gray70")

cl_pos <- cluster_louvain(g_pos)
groups <- communities(cl_pos)  
membership_df <- data.frame(
  vertex = names(membership(cl_pos)),
  cluster = membership(cl_pos)
)
membership_df
```
```{r}
vs_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:3,]

vs_clusters
```

```{r}
g_VS_colors <- VS_colors[c(3,1,2)]
g_VS_colors <- c(g_VS_colors,rep("gray70",10))
g2_VS_colors <- g_VS_colors
V(g_pos)$fill <- g_VS_colors[membership_df$cluster]
V(g_pos)$color <- g2_VS_colors[membership_df$cluster]
```


```{r,fig.height=6,fig.width=6}
set.seed(123)
p <- ggraph(g_pos, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightpink"), labels = c(">0")) +
  geom_node_point(aes(size=type,color = name,fill=name),shape=21,stroke=1.2) +
  scale_color_manual(values = setNames(V(g_pos)$color, V(g_pos)$name))   + # node colors
scale_fill_manual(values = setNames(V(g_pos)$fill, V(g_pos)$name)) +  # node fill
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() + 
  theme(legend.position = "none")

p
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/network_concat_pos.pdf",height = 6,width = 6)
p
dev.off()
```

**Based on negative associations**

```{r}
edges_neg <- edges %>%
  filter(emmean < 0)  # keep only negative associations

g_neg <- graph_from_data_frame(
  d = edges_neg, 
  vertices = nodes, 
  directed = FALSE)

VS_colors <- c("1st_VStertile"="#D73027",
               "2nd_VStertile"="orange",
               "3rd_VStertile"="#1A9850")

# Assign colors to nodes: countries get their color, metabolites gray
V(g_neg)$color <- ifelse(V(g_neg)$type,
                     VS_colors[V(g_neg)$name],
                     "gray70")

cl_neg <- cluster_louvain(g_neg)
groups <- communities(cl_neg)  
membership_df <- data.frame(
  vertex = names(membership(cl_neg)),
  cluster = membership(cl_neg)
)
membership_df
```
```{r}
vs_clusters <- (membership_df %>%
  group_by(cluster) %>%
  summarise(metabolites = paste(unique(vertex), collapse = ", ")))[1:3,]

vs_clusters
```

```{r}
g_VS_colors <- VS_colors[c(3,1,2)]
g_VS_colors <- c(g_VS_colors,rep("gray70",10))
g2_VS_colors <- g_VS_colors
V(g_neg)$fill <- g_VS_colors[membership_df$cluster]
V(g_neg)$color <- g2_VS_colors[membership_df$cluster]
```


```{r,fig.height=6,fig.width=6}
set.seed(123)
p <- ggraph(g_neg, layout = "fr") +
  geom_edge_link(aes(width = abs(weight), color = emmean > 0), alpha = 0.7) +
  scale_edge_color_manual(values = c("lightblue"), labels = c("<0")) +
  geom_node_point(aes(size=type,color = name,fill=name),shape=21,stroke=1.2) +
  scale_color_manual(values = setNames(V(g_neg)$color, V(g_neg)$name))   + # node colors
scale_fill_manual(values = setNames(V(g_neg)$fill, V(g_neg)$name)) +  # node fill
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() + 
  theme(legend.position = "none")

p 
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/network_concat_neg.pdf",height = 6,width = 6)
p
dev.off()
```


## RF model

```{r}
sub_metadata_df <- metadata_df %>%
dplyr::filter(
  tertile_label %in% c("1st_VStertile","3rd_VStertile"))

sub_concat_integrated_df <- concat_integrated_df[as.character(sub_metadata_df$study_id),]
```

### Original model

```{r}
model_name <- "concat_VS_tertile_vegan" 

prepared_df <- model_preparation(sub_concat_integrated_df,
                                 metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 10,
                          file = model_name,
                          reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model,
                   group = c("low VS_total","high VS_total"),
                   CI = TRUE)

roc_c <- ggtheme(roc_c) + 
  theme(legend.position = "none")

roc_c
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_orig_rf_vstotal.pdf",width = 6,height = 5)
roc_c
dev.off()
```

### Regressed out model

```{r}
# Covariates to adjust for
covars <- c("sex","bmi", "age", "Country", "eq_vegan_dur_total")
integrated_df <- sub_concat_integrated_df
all(rownames(integrated_df) == metadata_df$study_id)
adjusted_data <- adjust_for_covars(integrated_df, covars, sub_metadata_df)

model_name <- "concat_adjusted_vstotal_model_vegan" 

prepared_df <- model_preparation(adjusted_data,
                                 sub_metadata_df,
                                 predicting_col = "tertile_label")

rf_model <- make_rf_model(data=prepared_df,
                          outcome="Group",
                          N = 10,
                          file = model_name,
                          reuse=TRUE)

roc_c <- roc_curve(model_object = rf_model,
                    group = c("low VS_total","high VS_total"),
                   CI = TRUE)

roc_c <- ggtheme(roc_c) + 
  theme(legend.position = "none")

roc_c

```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/concat_adjusted_rf_vstotal.pdf",width = 6,height = 5)
roc_c
dev.off()
```

## RF importance

### Original model

```{r,fig.height=10,fig.width=5}
load("../intermediate_files/models/concat_VS_tertile_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("Metabolite") %>%
  `colnames<-`(c("Metabolite","Importance")) 
                                                 
var_imp <- var_imp %>%
  dplyr::mutate(Metabolite=factor(Metabolite,levels = Metabolite[order(var_imp$Importance,decreasing = FALSE)])) 

ggplot(var_imp) + 
  geom_col(aes(x=Metabolite,y=Importance)) + 
  theme(axis.text.x = element_text(angle=90)) + 
  coord_flip()
```

```{r}
concat_importance <- var_imp$Metabolite[var_imp$Importance>1.5]
```

### Adjusted model

```{r,fig.height=10,fig.width=5}
load("../intermediate_files/models/concat_adjusted_vstotal_model_vegan/rf_model.RData")

var_imp <- as.data.frame(importance(rf_model$trained_model)) %>%
  rownames_to_column("Metabolite") %>%
  `colnames<-`(c("Metabolite","Importance")) 

var_imp <- var_imp %>%
  dplyr::mutate(Metabolite=factor(Metabolite,levels = Metabolite[order(var_imp$Importance,decreasing = FALSE)])) 


ggplot(var_imp) + 
  geom_col(aes(x=Metabolite,y=Importance)) + 
  theme(axis.text.x = element_text(angle=90)) + 
  coord_flip()
```

```{r}
concat_importance <- var_imp$Metabolite[var_imp$Importance>1.5]
```


# Importance across different approaches

```{r,fig.height=10,fig.width=15}

list_core <- list()

list_core[["scpa_loadings"]] <- spca_loadings
list_core[["concat_univariate"]] <- unique(concat_significant)

venn <- ggvenn(list_core) 
venn
```

The recurring variables in each analysis were:

::: {.output-background style="background-color: lightgreen; padding: 10px;"}

```{r,echo=FALSE,eval=FALSE}
final_significance <- Reduce(intersect, list_core)
final_significance
```
```{r,echo=TRUE}
final_significance <- list_core[["concat_univariate"]]
```

:::

# Vegan index

To create a vegan index based on concatenated dataset, we will use linear regression in order to get coefficients on significant variables that were revealed across the analysis:

```{r}
sign
```


```{r}
plot_list <- list()
rhos <- rep(NA,10)

concat_for_index <- concat_for_uni[sample(rownames(concat_for_uni)),]

cv_splits <- cv_folds(concat_for_index)

vars <- sign

fml <- as.formula(paste("VS_total ~", paste(vars, collapse = " + ")))

for (i in 1:length(cv_splits)){
  model <- lm(fml, data = concat_for_uni[cv_splits[[i]]$train,])
  VG_index <- predict(model,concat_for_uni[cv_splits[[i]]$val,])
  
  sub_df <- concat_for_uni[cv_splits[[i]]$val,]
  sub_df$VGindex <- VG_index

  cor_val <- cor(sub_df$VGindex,
               sub_df$VS_total,
               method = "spearman")

  p <- ggplot(sub_df, aes_string(x = "VGindex", y = "VS_total"),color="blue") +
       geom_point() + 
  geom_smooth(method = "lm") + 
       theme_minimal() + 
  ggtitle(paste("Rho =",round(cor_val,2)))

  plot_list[[i]] <- p
  rhos[i] <- round(cor_val,2)
  
}

```

```{r}
wrap_plots(plot_list)
```

```{r}
mean(rhos)
```

# Plot for publication

```{r,fig.height=10,fig.width=15}
ordered_participants <- metadata_df %>%
  dplyr::arrange(VS_total)

concat_significant_df <- concat_significant_df %>%
  dplyr::filter(metabolite %in% final_significance) %>%
  dplyr::arrange(Estimate)

heatmap_concat_integrated_df <- concat_integrated_df

heatmap_concat_integrated_df <- heatmap_concat_integrated_df[as.character(ordered_participants$study_id),
                             as.character(concat_significant_df$metabolite)]

# Create a grouping variable for each block of 10 rows
group_id <- rep(1:ceiling(nrow(heatmap_concat_integrated_df) / 10), each = 10, 
                length.out = nrow(heatmap_concat_integrated_df))

# Calculate the mean for each group
df_grouped <- aggregate(heatmap_concat_integrated_df, 
                        by = list(Group = group_id), FUN = median)

# Remove the Group column name prefix if you want just the values
rownames(df_grouped) <- paste0("Group_", df_grouped$Group)
df_grouped$Group <- NULL

```


```{r}
df_long <- df_grouped %>% t() %>% as.data.frame() %>%
  rownames_to_column("Metabolite") %>%
  pivot_longer(cols = -Metabolite, names_to = "Participant", values_to = "value")

df_long$Metabolite <- factor(df_long$Metabolite, levels=concat_significant_df$metabolite)
df_long$Participant <- factor(as.character(df_long$Participant), levels=paste0("Group_",1:37))


p_heat <- ggplot(df_long, aes(x = Participant, y = Metabolite, fill = value)) +
  geom_tile(color = "white",width=1) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_text(face = "bold",size = 9),
        axis.text.y = element_blank(),
        panel.grid = element_blank()) + 
  labs(fill = "Median metabolite z-score") +
  xlab("Vegans (ordered and grouped by VS_total)") + 
  ylab("") + 
  scale_fill_gradient2(
    low = "darkcyan", mid = "white", high = "brown", midpoint = 0)

# get legend
leg_heat_p <- ggpubr::get_legend(p_heat)

# remove legend
p_heat <- p_heat + theme(legend.position = "none")
```

```{r}
annotation_df <- concat_significant_df
#annotation_df$metabolite <- c("Phosphatidylcholine (34:1)",
#                              "Sphingomyelin (d42:2)",
                             #  "Octenoylcarnitine",
                             # "Lysophosphatidylcholine (17:0)",
                             # "Indoleacrylic acid",
                             # "Ergothioneine",
                             # "Phosphatidic acid (44:5)",
                             # "Phosphatidylcholine (40:6)",
                             # "Acetylornithin",
                             # "Lysophosphatidylcholine (22:6)")

annotation_df$metabolite <- c("Phosphatidylcholine (34:1)",
                              "Sphingomyelin (d42:2)",
                              "Glutarylcarnitine",
                             "Lysophosphatidylcholine (17:0)",
                             "Indoleacrylic acid",
                             "Phosphatidic acid (44:5)",
                             "Acetylornithin",
                             "Phosphatidylcholine (40:6)",
                             "Lysophosphatidylcholine (22:6)")
annotation_df$metabolite <- factor(annotation_df$metabolite,
                                   levels=annotation_df$metabolite)


ann_p <- ggplot(
  data=annotation_df,
  aes(x = "Estimate", y = metabolite, fill = Estimate)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "#313695", mid = "white", high = "#A50026", midpoint = 0) +
  theme_minimal() + 
  theme(axis.text.y = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.title.y= element_text(face = "bold"),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) + 
  ylab("Significant metabolites") + 
  xlab("") + 
  labs(fill="LM estimate")

# get legend
leg_ann_p <- ggpubr::get_legend(ann_p)

# Remove legend from the original plot
ann_p <-ann_p + theme(legend.position = "none")

```

```{r}
# correlations
df_for_plot <- concat_integrated_df %>% dplyr::select(final_significance)
df_for_plot <- df_for_plot[,concat_significant_df$metabolite]
corr_object <- cor(concat_for_uni$VS_total,df_for_plot,method="spearman")
corr_df <- data.frame(Metabolite=colnames(corr_object),
                      Rho=corr_object[1,])
corr_df$Metabolite <- factor(corr_df$Metabolite,
                             levels=corr_df$Metabolite)

p_corr <- ggplot(corr_df) + 
  geom_point(aes(x="Rho",y= Metabolite,size = abs(Rho), fill = Rho, color=Rho), 
             shape = 21) +
  scale_fill_gradient2(name = "Rho", 
                       low = "blue", mid = "white", high = "red", midpoint = 0) +
    scale_color_gradient2(name = "Rho", 
                          low = "blue", mid = "white", 
                          high = "red", midpoint = 0) + 
  guides(size = "none") + 
  theme_minimal() + 
  theme(axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y= element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) + 
  xlab("") 


leg_corr_p <- ggpubr::get_legend(p_corr)

# Remove legend from the original plot
p_corr <-
  p_corr +
  theme(
    legend.position = "none")
```

```{r}
legends <- ggarrange(leg_heat_p, leg_ann_p, leg_corr_p,nrow = 3, align = "hv")

p <- wrap_plots(ann_p,p_corr,p_heat,legends,ncol=4,widths = c(0.1,0.15, 1, 1))
```

```{r,fig.height=5,fig.width=20}
p
```

```{r,echo=FALSE}
pdf("../results/vegan/Q2_VStotal/final_significance.pdf",height = 5,width = 7)
p
dev.off()
```