---
title: "VEGANScreener + metabolomics"
subtitle: "Preprocessing part 2: Metabolomics"
author: "Petra Polakovicova"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    df-print: paged
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
execute: 
  warning: false
  message: false
  echo: true
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

This script serves for preprocessing the metabolomics data. Here, we have two datasets derived from the same samples. One dataset is a result of LC-MS approach, the other is obtained by NMR.

How will the analysis be addressed:

**Preprocessing per Platform**

Treating the NMR and LC-MS results as two separate datasets at first. Normalize and QC them separately:

-   NMR → log-transform if needed, auto-scale.
-   LC-MS → batch correction (ComBat), impute missing values, scaling.

**Data Integration**

Keep both NMR and LC-MS features, concatenate after scaling. Use mixOmics sPCA that is designed for multi-omics from heterogeneous platforms. This is the modern state-of-the-art for such cases. Pros: Keeps full richness of data. Cons: Requires careful handling of batch/variance scaling.

```{r}
library(sva)
library(mixOmics)
library(mice)
library(openxlsx)
library(ggplot2)
library(reshape2)
library(DataExplorer)
library(dplyr)
library(mice)
library(MOFA2)
library(patchwork)
library(ggvenn)
library(tidyverse)
library(ggpubr)
library(pheatmap)
library(data.table)
```

# Preprocessing

## Listing identified metabolites

**LCMS**

```{r}
lc_ms_df <- read.xlsx(
  "../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx")

metabolites_names_lcms <- read.xlsx(
  "../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx",
  sheet="metab")

metabolites_lcms <- metabolites_names_lcms$ident
```

**NMR**

```{r}
nmr_df <- read.xlsx(
  "../../../Data/raw/metabolomics/edited/NMR_metabolome_VGScreener.xlsx")

metabolites_nmr <- nmr_df[1,4:30]
nmr_df <- nmr_df[-c(1,2),]

samplenames <- nmr_df[,1]
rownames(nmr_df) <- samplenames

nmr_df <- nmr_df[,-c(1,2,3)]
colnames(nmr_df) <- metabolites_nmr

nmr_df <- as.data.frame(apply(nmr_df,2,function(x) as.numeric(x))) %>%
  `row.names<-`(samplenames)

metabolites_nmr <- as.character(metabolites_nmr)
```

**Comparison**

```{r}
list_core <- list()
list_core[["NMR"]] <- metabolites_nmr
list_core[["LCMS"]] <- metabolites_lcms

venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("NMR vs LCMS") + 
    guides(fill = "none")

venn
```

## Import

**Load**

```{r}
# Metadata
metadata_df <- read.table(
  file = "../../../data/raw/veganscreener/metadata_extracted_Fulldata.tsv", 
  header = TRUE) %>%
  dplyr::filter(group=="VG")

# LCMS
lc_ms_df <- read.xlsx(
  "../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx")

lc_ms_df <- lc_ms_df %>% 
  column_to_rownames("study_id") %>%
  t() %>%
  as.data.frame(check.names = FALSE) %>%
  rownames_to_column("Metabolite")

# NMR
nmr_df <- read.xlsx(
  "../../../Data/raw/metabolomics/edited/NMR_metabolome_VGScreener.xlsx")

nmr_df <- nmr_df %>% 
  column_to_rownames("study_id") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Metabolite")
```

**Remove outliers** (other sex, bmi \< 15)

```{r}
metadata_df <- metadata_df %>% dplyr::filter(bmi>15,sex != "OTHER") %>% drop_na(VS_total)
```

**Get only important samples**

```{r}
# get common samples
common_samples <- Reduce(intersect, 
                         list(colnames(lc_ms_df)[-1],
                              colnames(nmr_df)[-1], 
                              metadata_df$study_id))

# extract only common samples in these datasets
lc_ms_df <- lc_ms_df[,c("Metabolite",common_samples)]
nmr_df <- nmr_df[,c("Metabolite",common_samples)]
metadata_df <- metadata_df %>% dplyr::filter(study_id %in% common_samples) 

```

```{r}
write.table(metadata_df,"../../../data/preprocessed/metadata_preprocessed_v3.tsv",
            row.names = FALSE,quote=FALSE,sep="\t")
```

## LC-MS

### Missing value test

We miss 0.02% of data.

```{r}
df_for_test <- lc_ms_df
df_for_test[df_for_test==0] <- NA
df_for_test[df_for_test==""] <- NA

100*sum(is.na(df_for_test))/(dim(df_for_test)[1]*dim(df_for_test)[2])
```

### Removing NaNs

Removing metabolites when we have 20% missing. All metabolites preserved.

```{r}
df_for_test[df_for_test==0] <- NA
data_clean <- df_for_test %>%
  rowwise() %>%
  filter(mean(is.na(c_across(-Metabolite))) < 0.2) %>%
  ungroup()
```

### Imputation

```{r}
data_clean[data_clean==0] <- NA

imputed_data <- mice(as.data.frame(t(data_clean[,-1])), 
                     m = 5, method = 'pmm', seed = 123)

completed_data <- complete(imputed_data, 1)

colnames(completed_data) <- data_clean$Metabolite

which(is.na(completed_data), arr.ind = TRUE)
100*sum(is.na(completed_data))/(50*382)
```

We successfully imputed the 0.2% of data.

**Comparison**

To check the quality of imputation, we will plot the imputed values.

```{r}
data_clean <- data_clean %>% 
  column_to_rownames("Metabolite") %>%  
  t() %>% as.data.frame()

nans <- which(is.na(data_clean), arr.ind = TRUE)

plot_df <- as.data.frame(completed_data[,unique(nans[,2])])
plot_df$imputed <- "NO"
plot_df$imputed[unique(nans[,1])] <- "YES"
colnames(plot_df) <- c("value","imputed")

p <- ggplot(plot_df) + 
  geom_jitter(aes(x="",y=value,color=imputed))

p
```

The imputation looks fine, now lets see if the imputation did not disturb the overall composition using PCA.

```{r}
data_before <- data_clean
data_before[is.na(data_before)] <- 0

pca_before <- as.data.frame(
  prcomp(data_before,center = TRUE,scale=TRUE)$x 
)
pca_before$imputed_sample <- "NO"
pca_before$imputed_sample[unique(nans[,1])] <- "YES"

p_before <- ggplot(pca_before) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", 
                                       fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 

```

```{r}
pca_after <- as.data.frame(
  prcomp(completed_data,center = TRUE,scale=TRUE)$x
)
pca_after$imputed_sample <- "NO"
pca_after$imputed_sample[unique(nans[,1])] <- "YES"

p_after <- ggplot(pca_after) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 
```

Looks fine.

```{r}
wrap_plots(p_before,p_after,guides = "collect")
```

### INT

```{r}
x <- apply(completed_data,2,function(x) qnorm((rank(x,na.last="keep")-0.5)/sum(!is.na(x))))
```

Basic plots for visualizing the distribution:

```{r,fig.width=25}
plot_list <- list()
name <- colnames(x)
i <- 1
for (col in name){
  p <- ggplot(x) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(x[,col],na.rm = TRUE),2),"+-", round(sd(x[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}

wrap_plots(plot_list)

```

```{r}
write.table(x,"../../../Data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv",row.names = TRUE,sep="\t")
```

## NMR

### Missing value test

We miss 0.0% of data.

```{r}
df_for_test <- nmr_df
df_for_test[df_for_test==0] <- NA
df_for_test[df_for_test==""] <- NA

100*sum(is.na(df_for_test))/(dim(df_for_test)[1]*dim(df_for_test)[2])
```

### Removing NaNs

All metabolites preserved, as we didn't have any NA.

```{r}
df_for_test[df_for_test==0] <- NA
data_clean <- df_for_test %>%
  rowwise() %>%
  filter(mean(is.na(c_across(-Metabolite))) < 0.2) %>%
  ungroup()
```

### Imputation

```{r}
completed_data <- data_clean %>% column_to_rownames("Metabolite") %>% t()
```

### INT

```{r}
x <- apply(completed_data,2,function(x) qnorm((rank(x,na.last="keep")-0.5)/sum(!is.na(x))))

```

Basic plots for visualizing the distribution:

```{r,fig.width=25}
plot_list <- list()
name <- colnames(x)
i <- 1
for (col in name){
  p <- ggplot(x) + 
    geom_histogram(aes(x=!!sym(col)),fill="darkblue") + 
    theme_minimal() + 
    theme(axis.title = element_text(face="bold",size=20),
          axis.text=element_text(face="bold",size=20),
          title = element_text(face = "plain",size=20)) + 
    xlab(name[i]) +
    ggtitle(paste(round(mean(x[,col],na.rm = TRUE),2),"+-", round(sd(x[,col],na.rm = TRUE),2)))
  
  plot_list[[length(plot_list)+1]] <- p
  i <- i + 1 
                                            
}

wrap_plots(plot_list)

```

```{r}
write.table(x,"../../../Data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv",row.names = TRUE,sep="\t")
```

# Integration

Firstly, we need to import our data.

```{r}
lcms_df <- read.table(
  "../../../Data/preprocessed/metabolomics/lc_ms_preprocessed_INT_v3.tsv")

nmr_df <- read.table(
  "../../../Data/preprocessed/metabolomics/nmr_preprocessed_INT_v3.tsv")

metadata_df <- read.table(
  "../../../Data/preprocessed/metadata_preprocessed_v3.tsv",header=TRUE)

common_samples <- Reduce(intersect, list(rownames(lcms_df),
                                     rownames(nmr_df), 
                                     metadata_df$study_id))

lcms_vegans <- lcms_df[common_samples,]
nmr_vegans <- nmr_df[common_samples,]

```

### MixOmics SPCA

```{r}
integrated_df <- cbind(nmr_vegans,lcms_vegans)

spca_result <- mixOmics::spca(integrated_df,ncomp = 20)

spca_data <- spca_result$variates$X

write.table(spca_data,"../../../Data/preprocessed/metabolomics/spca_integrated_vegans_INT_v3.tsv",row.names = TRUE, sep="\t")
```

```{r}
save(spca_result,file="../../Results/preprocessing/spca_result_vegans_INT_v3.RData")
```
