---
title: "Dietary metadata - veganscreener -preprocessing"
format: html
editor: visual
---

```{r}
library(dplyr)
library(magrittr)
library(mice)
library(VIM)
library(naniar)
library(tidyverse)
library(openxlsx)
```

# VEGANSCREENER

```{r}
dietary_df <- read.table("../../../Data/raw/veganscreener/metadata_extracted.tsv",header=TRUE)

dietary_df <- dietary_df %>%
  dplyr::select(!ends_with("_2"))
```

In DE, there are vegans only.

```{r}
# number of vegans and omnivores
dietary_df %>% dplyr::group_by(eating_habits) %>%
  count() 

dietary_df %>% dplyr::group_by(eq_included_group) %>%
  count() 

dietary_df %>% dplyr::group_by(eq_included_group,Country) %>%
  count() 
```

## Missing values imputation 

### Missing data analysis

#### Summary

```{r}
dietary_df[dietary_df==""] <- NA

dietary_df <- 
  dietary_df %>% 
  mutate_if(is.character, as.factor)

summary(dietary_df)

miss_var_summary(dietary_df)
```

```{r}
df <- as.data.frame(miss_var_summary(dietary_df)) %>%
  dplyr::mutate(pct_miss=as.numeric(pct_miss))

write.xlsx(df,"../../Results/preprocessing/missing_values_summary.xlsx")
```

- 196 people are missing bdi_score, happy, physical_health, health_status at once

- 176 people have all values filled out

- 62 people miss the information the information about vegan_excpt_something

- 49 people miss the previous + bdi_score, happy, physical_health, health_status -> these are all omnivores (62+49+1=112)

```{r,fig.width=10,fig.height=4}
aggr(dietary_df[,1:23], numbers = TRUE, prop = FALSE)
```

The same thing with custom plot:

```{r}
p <- md.pattern(dietary_df[,1:23], plot = FALSE)

library(pheatmap)

pheatmap(p[-10,-24],
         cluster_rows = FALSE,
         cluster_cols = FALSE, col = c("red","blue"))
```

#### Correlations

Correlations between missing values: There is strong correlation between some missing values: (education,family_income, residence area), (bdi_score, happy,physical_health, health_status).

```{r}
dummyNA <- 
  as.data.frame(abs(is.na(dietary_df))) %>% 
  select(bmi,eq_vegan_dur_total,education,family_income,residence_area,bdi_score,happy,physical_health,health_status) # pick variable with missing values only

p <- cor(dummyNA) %>% round(digits = 2)
pheatmap(p,cluster_rows = FALSE,cluster_cols = FALSE)
```

Correlation between variables with missing values and variable with non-missing values. 

```{r,fig.height=25,fig.width=5}
p <- cor(dietary_df %>% mutate_if(is.factor, as.numeric), 
    dummyNA, use = "pairwise.complete.obs") %>% 
  round(digits = 2)

p[p<=-0.4] <- -1
p[p>=0.4] <- 1
p[abs(p)!=1] <- 0

pheatmap(p,cluster_rows = FALSE,cluster_cols = FALSE)
```

To make it more visible, we extract only rows with evident correlation.
study_id, country, eating_habits, eq_included_group, diet_supplements, malt_num, milk_num, chee_num, b12_num, vitd_num, VS_total


The missing information about physical and emotional health is country specific - countries with lower IDs like BE, CZ tends to not have this information. 

The correlation of eq_vegan_dur_total is not relevant, as this points out the patterns between vegans and omnivores (as NA=omnivore).

```{r}
p <- cor(dietary_df %>% 
           dplyr::select(study_id, Country, eating_habits, 
                         eq_included_group, diet_supplements, 
                         malt_num, milk_num, chee_num, b12_num, vitd_num,
                         VS_total) %>%
           mutate_if(is.factor, as.numeric), 
    dummyNA, use = "pairwise.complete.obs") %>% 
  round(digits = 2)

pheatmap(p,cluster_rows = FALSE,cluster_cols = FALSE)
```

```{r}
dietary_df %>% dplyr::group_by(Country) %>%
   summarise(across(everything(), ~ sum(is.na(.)), .names = "na_{.col}"))
```

#### MCAR test

The null hypothesis is the missing data is MCAR. Thus, a higher p value indicates probability of missing data is MCAR. The test shows that the missingness in our data is either MAR or MNAR.

**Our data are MAR or MNAR.**


```{r}
mcar_test(dietary_df %>% 
           dplyr::select(Country, eating_habits, 
                         eq_included_group, diet_supplements, 
                         malt_num, milk_num, chee_num, b12_num, vitd_num,
                         VS_total, bmi, eq_vegan_dur_total, education,
                      family_income, residence_area, bdi_score, happy,
                      physical_health, health_status))
```

### Imputation with mice()

Initialization of the model

```{r}
#dietary_df_forimp <- dietary_df %>%
#  select(-study_id) %>%
#  mutate(across(where(is.character), as.factor)) %>%
#  mutate(across(where(is.factor), as.integer)-1)

dietary_df_forimp <- dietary_df %>%
  select(-study_id) %>%
  mutate(across(
    where(~ is.character(.) || is.factor(.)),
    ~ as.integer(as.factor(.)) - 1,
    .names = "{.col}_mutated"
  ))

ini <- mice(dietary_df_forimp, m = 1, maxit = 0) 
pm <- ini$predictorMatrix
meth <- ini$method
```

We extracted these pre-set terms and edited it manually according to the recommendations from https://stefvanbuuren.name/fimd/sec-modelform.html#sec:predictors. 

```{r,eval=FALSE}
write.xlsx(pm,"../../Results/preprocessing/prediction_matrix.xlsx",rowNames = TRUE)
```

```{r}
write.xlsx(quickpred(dietary_df_forimp),"../../Results/preprocessing/quick_pred.xlsx",
           rowNames = TRUE)
```

We will impute values in 9 columns:

- bmi
- sex
- education
- family_income
- residence_area
- diet_supplements
- sleep_duration
- wbre_num

```{r}
pm <- read.xlsx("../../Results/preprocessing/quick_pred_mutated.xlsx",
                rowNames = TRUE)

meth <- ini$method
imputed_columns <- rownames(pm %>% dplyr::filter(rowSums(.)>0))
imputed_columns <- c("bmi","sex","education","family_income","residence_area",
                     "diet_supplements","sleep_duration","wbre_num")
meth[!(names(meth) %in% imputed_columns)] <- ""
#meth[meth!=""] <- "pmm"
```

https://stackoverflow.com/questions/73682067/mice-not-imputing-all-variables-with-missing-values


```{r}

regImp <- mice(dietary_df_forimp,
               m = 5, 
               method = meth,
               seed = 123,
               predictorMatrix = as.matrix(pm))

dietary_regImp <- complete(regImp, 1)
summary(dietary_regImp)[,c(2,3,15,16,17,14,22,30)]
```

```{r,fig.width=10,fig.height=10}
library(lattice)
imputed_plot <- stripplot(regImp, col=c("grey",mdc(2)),pch=c(1,20))
```

```{r}
pdf("../../Results/preprocessing/strip_plot.pdf",width = 20,height = 20)
imputed_plot
dev.off()
```

The number of chains is equal to the number of imputed datasets. A chain refers to the chain of regression models that is used to generate the imputed values. The length of each chain is equal to the number of iterations.

The convergence can be visualised by plotting the means in a convergence plot.

```{r}
plot(regImp)
```

```{r}
dietary_df_clean <- dietary_regImp %>%
  select(-ends_with("_mutated")) %>%
  `rownames<-`(dietary_df$study_id) %>%
  rownames_to_column("study_id")
```

```{r}
write.table(dietary_df_clean,"../../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",sep="\t",row.names = FALSE)
```

