---
title: "Metabolomics preprocessing"
author: "Petra Polakovicova"
format: html
editor: visual
---

This script serves for preprocessing the metabolomics data. Here, we have two datasets derived from the same samples. One dataset is a result of LC-MS approach, the other is obtained by NMR.

How will the analysis be addressed:

**Preprocessing per Platform**

Treating the NMR and LC-MS results as two separate datasets at first. Normalize and QC them separately:

-   NMR → log-transform if needed, auto-scale.
-   LC-MS → batch correction (ComBat), impute missing values, scaling.

**Data Integration Options**

*Option A: Integrate After Feature Extraction* Summarize both NMR and LC-MS to biological pathways (using e.g. Mummichog, MetaboAnalyst, MS Peaks to Pathways). Pros: Easier, avoids platform bias. C ons: Loses fine-grained info.

*Option B: Integrate Full Features* Keep both NMR and LC-MS features, concatenate after scaling. Use MOFA2 or DIABLO (from mixOmics) — these tools are designed for multi-omics from heterogeneous platforms. This is the modern state-of-the-art for such cases. Pros: Keeps full richness of data. Cons: Requires careful handling of batch/variance scaling.

MOFA2: https://biofam.github.io/MOFA2/ DIABLO: https://mixomics.org/mixdiablo/

```{r}
library(sva)
library(mixOmics)
library(mice)
library(openxlsx)
library(ggplot2)
library(reshape2)
library(DataExplorer)
library(dplyr)
library(mice)
library(MOFA2)
library(patchwork)
library(ggvenn)
library(tidyverse)
library(ggpubr)
```

# Preprocessing

## Listing identified metabolites

**LCMS**

```{r}
lc_ms_df <- read.xlsx("../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx")

metabolites_names_lcms <- read.xlsx("../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx",sheet="metab")

metabolites_lcms <- metabolites_names_lcms$ident
```

**NMR**

```{r}
nmr_df <- read.xlsx("../../../Data/raw/metabolomics/NMR_metabolome_VGScreener.xlsx")

metabolites_nmr <- nmr_df[1,4:30]
nmr_df <- nmr_df[-c(1,2),]

samplenames <- nmr_df[,1]
rownames(nmr_df) <- samplenames

nmr_df <- nmr_df[,-c(1,2,3)]
colnames(nmr_df) <- metabolites_nmr

nmr_df <- as.data.frame(apply(nmr_df,2,function(x) as.numeric(x))) %>%
  `row.names<-`(samplenames)

metabolites_nmr <- as.character(metabolites_nmr)
```

**Comparison**

```{r}
list_core <- list()
list_core[["NMR"]] <- metabolites_nmr
list_core[["LCMS"]] <- metabolites_lcms

venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("NMR vs LCMS") + 
    guides(fill = "none")

venn
```

## LC-MS

```{r}
lc_ms_df <- read.xlsx("../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx")

metabolites_names <- read.xlsx("../../../Data/raw/metabolomics/edited/VeganScreener_lcms_edited.xlsx",sheet="metab")

all(colnames(lc_ms_df)[-1] == metabolites_names$`m/z`)

metabolites_names$Metabolite <- metabolites_names$ident

colnames(lc_ms_df) <- c("study_id",metabolites_names$Metabolite)

lc_ms_df <- lc_ms_df %>% 
  column_to_rownames("study_id") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("Metabolite")

rownames(lc_ms_df) <- NULL

write.table(metabolites_names,"../../../Data/preprocessed/metabolomics/lcms_metabolites_identification.tsv",sep="\t")

```

### Missing value test

We miss 0.02% of data. 

```{r}
df_for_test <- lc_ms_df
df_for_test[df_for_test==0] <- NA
df_for_test[df_for_test==""] <- NA

100*sum(is.na(df_for_test))/(50*499)
```

### Removing NaNs

Removing metabolites when we have 20% missing. All metaboolites preserved. 

```{r}
data_clean <- df_for_test %>%
  rowwise() %>%
  filter(mean(is.na(c_across(-Metabolite))) < 0.2) %>%
  ungroup()
```

### Imputation

```{r}
data_clean[data_clean==0] <- NA

imputed_data <- mice(as.data.frame(t(data_clean[,-1])), 
                     m = 5, method = 'pmm', seed = 123)

completed_data <- complete(imputed_data, 1)

colnames(completed_data) <- data_clean$Metabolite

which(is.na(completed_data), arr.ind = TRUE)
100*sum(is.na(completed_data))/(50*499)
```

We successfully imputed the 0.2% of data.

**Comparison**

To check the quality of imputation, we will plot the imputed values.

```{r}
data_clean <- data_clean %>% 
  column_to_rownames("Metabolite") %>%  
  t() %>% as.data.frame()

nans <- which(is.na(data_clean), arr.ind = TRUE)

plot_df <- as.data.frame(completed_data[,unique(nans[,2])])
plot_df$imputed <- "NO"
plot_df$imputed[unique(nans[,1])] <- "YES"
colnames(plot_df) <- c("value","imputed")

p <- ggplot(plot_df) + 
  geom_jitter(aes(x="",y=value,color=imputed))

p
```

The imputation looks fine, now lets see if the imputation did not disturb the overall composition using PCA.

```{r}
data_before <- data_clean
data_before[is.na(data_before)] <- 0

pca_before <- as.data.frame(
  prcomp(data_before,center = TRUE,scale=TRUE)$x 
)
pca_before$imputed_sample <- "NO"
pca_before$imputed_sample[unique(nans[,1])] <- "YES"

p_before <- ggplot(pca_before) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", 
                                       fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 

```

```{r}
pca_after <- as.data.frame(
  prcomp(completed_data,center = TRUE,scale=TRUE)$x
)
pca_after$imputed_sample <- "NO"
pca_after$imputed_sample[unique(nans[,1])] <- "YES"

p_after <- ggplot(pca_after) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 
```

Looks fine. 

```{r}
wrap_plots(p_before,p_after,guides = "collect")
```

### Non-tranformed plots 

**Boxplots and density**

```{r}
metadata_df <- read.table("../../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",header=TRUE)
```

```{r}
plot_list_boxplots <- list()
plot_list_density <- list()

plot_df <- merge(completed_data %>% rownames_to_column("study_id"),
                 metadata_df %>% dplyr::select("study_id","eq_included_group"),all=TRUE) %>%
  drop_na()

for (column in colnames(completed_data)){
  # boxplot
    p <- ggplot(plot_df) + 
      geom_boxplot(aes(x="",y=!!sym(column)),outliers = FALSE) +
    geom_jitter(height=0, width=0.3,
                aes(x="",y=!!sym(column),color=eq_included_group)) + 
    scale_color_manual(values=c("brown","darkgreen")) + 
      theme(legend.position = "right")
    
    plot_list_boxplots[[length(plot_list_boxplots)+1]] <- p
    
    pd <- ggplot(plot_df) + 
      geom_density(aes(x=!!sym(column),fill=eq_included_group),alpha=0.5) +
      theme(legend.position = "right") + 
    scale_fill_manual(values=c("brown","darkgreen")) 
    
    plot_list_density[[length(plot_list_density)+1]] <- pd
    
}

pb <- wrap_plots(plot_list_boxplots,guides = "collect")
pd <- wrap_plots(plot_list_density,guides = "collect")
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/lcms_boxplots.pdf",height = 50,width = 50)
pb
dev.off()
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/lcms_density.pdf",height = 50,width = 60)
pd
dev.off()
```

### Log-transformation

Log-transformation.

```{r}
data_clean_log <- as.data.frame(
  apply(completed_data,2,function(x) as.numeric(x)))

data_clean_log <- log(data_clean_log)
rownames(data_clean_log) <- rownames(completed_data)
```

### Tranformed plots 

**Boxplots and density**

```{r}
plot_list_boxplots <- list()
plot_list_density <- list()

plot_df <- merge(data_clean_log %>% rownames_to_column("study_id"),
                 metadata_df %>% dplyr::select("study_id","eq_included_group"),all=TRUE) %>%
  drop_na()

for (column in colnames(data_clean_log)){
  # boxplot
    p <- ggplot(plot_df) + 
      geom_boxplot(aes(x="",y=!!sym(column)),outliers = FALSE) +
    geom_jitter(height=0, width=0.3,
                aes(x="",y=!!sym(column),color=eq_included_group)) + 
    scale_color_manual(values=c("brown","darkgreen")) + 
      theme(legend.position = "right")
    
    plot_list_boxplots[[length(plot_list_boxplots)+1]] <- p
    
    pd <- ggplot(plot_df) + 
      geom_density(aes(x=!!sym(column),fill=eq_included_group),alpha=0.5) +
      theme(legend.position = "right") + 
    scale_fill_manual(values=c("brown","darkgreen")) 
    
    plot_list_density[[length(plot_list_density)+1]] <- pd
    
}

pb <- wrap_plots(plot_list_boxplots,guides = "collect")
pd <- wrap_plots(plot_list_density,guides = "collect")
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/log_lcms_boxplots.pdf",height = 50,width = 50)
pb
dev.off()
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/log_lcms_density.pdf",height = 50,width = 60)
pd
dev.off()
```

### Autoscaling

```{r}
data_scaled <- data_clean_log
data_scaled <- scale(data_scaled, center = TRUE, scale = TRUE)
```

```{r}
write.table(data_scaled,"../../../Data/preprocessed/metabolomics/lc_ms_preprocessed.tsv",row.names = TRUE,sep="\t")
```

## NMR

```{r}
nmr_df <- read.xlsx(
  "../../../Data/raw/metabolomics/NMR_metabolome_VGScreener.xlsx")

metabolites <- nmr_df[1,4:30]
nmr_df <- nmr_df[-c(1,2),]

samplenames <- nmr_df[,1]
rownames(nmr_df) <- samplenames

nmr_df <- nmr_df[,-c(1,2,3)]
colnames(nmr_df) <- metabolites

nmr_df <- as.data.frame(apply(nmr_df,2,function(x) as.numeric(x))) %>%
  `row.names<-`(samplenames)

```

### Missing value test

0.44% of values are missing.

```{r}
df_for_test <- nmr_df
df_for_test[df_for_test<=0] <- NA
df_for_test[df_for_test==""] <- NA

100*sum(is.na(df_for_test))/(488*27)
```

### Removing NaNs

Removing metabolites when we have 20% missing.

```{r}
data_clean <- df_for_test %>%
  t() %>% as.data.frame() %>% 
  rownames_to_column("Metabolite") %>%
  rowwise() %>%
  filter(mean(is.na(c_across(-Metabolite))) < 0.2) %>%
  ungroup()
```


### Imputation

```{r}
imputed_data <- mice(as.data.frame(t(data_clean[,-1])), 
                     m = 5, method = 'pmm', seed = 123)

completed_data <- complete(imputed_data, 1)

colnames(completed_data) <- data_clean$Metabolite

which(is.na(completed_data), arr.ind = TRUE)
100*sum(is.na(completed_data))/(27*488)
```

We have succesfully imputed the missing data.

**Comparison**

To check the quality of imputation, we will plot the imputed values.

```{r,fig.width=10}
data_clean <- data_clean %>% 
  column_to_rownames("Metabolite") %>%  
  t() %>% as.data.frame()

nans <- which(is.na(data_clean), arr.ind = TRUE)

# first metabolite
nans1 <- as.data.frame(nans) %>% dplyr::filter(col==unique(nans[,2])[1])

plot_df <- as.data.frame(completed_data[,unique(nans1[,2])])
plot_df$imputed <- "NO"
plot_df$imputed[unique(nans1[,1])] <- "YES"
colnames(plot_df) <- c("M1","imputed")

p1 <- ggplot(plot_df) + 
  geom_jitter(aes(x="",y=M1,color=imputed))

# second metabolite
# first metabolite
nans2 <- as.data.frame(nans) %>% dplyr::filter(col==unique(nans[,2])[2])

plot_df <- as.data.frame(completed_data[,unique(nans2[,2])])
plot_df$imputed <- "NO"
plot_df$imputed[unique(nans2[,1])] <- "YES"
colnames(plot_df) <- c("M2","imputed")

p2 <- ggplot(plot_df) + 
  geom_jitter(aes(x="",y=M2,color=imputed))

p1 + p2
```

The imputation looks fine, now lets see if the imputation did not disturb the overall composition using PCA.

```{r}
data_before <- data_clean
data_before[is.na(data_before)] <- 0

pca_before <- as.data.frame(
  prcomp(data_before,center = TRUE,scale=TRUE)$x 
)
pca_before$imputed_sample <- "NO"
pca_before$imputed_sample[unique(nans[,1])] <- "YES"

p_before <- ggplot(pca_before) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", 
                                       fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 

```

```{r}
pca_after <- as.data.frame(
  prcomp(completed_data,center = TRUE,scale=TRUE)$x
)
pca_after$imputed_sample <- "NO"
pca_after$imputed_sample[unique(nans[,1])] <- "YES"

p_after <- ggplot(pca_after) + 
     geom_point(aes(x=PC1,y=PC2,color=imputed_sample)) + 
     theme_minimal() + 
     theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
           axis.ticks.x = element_line(size=0.3,color = "black"),
           axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
           axis.text = element_text(face = "bold",colour = "black"),
           axis.title = element_text(face = "bold",colour = "black"),
           panel.grid = element_blank()) + 
     theme(legend.position = "right") 
```

Looks fine. 

```{r}
wrap_plots(p_before,p_after,guides = "collect")
```

### Non-tranformed plots 

**Boxplots and density**

```{r}
metadata_df <- read.table("../../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",header=TRUE)
```

```{r}
plot_list_boxplots <- list()
plot_list_density <- list()

plot_df <- merge(completed_data %>% rownames_to_column("study_id"),
                 metadata_df %>% dplyr::select("study_id","eq_included_group"),all=TRUE) %>%
  drop_na()

for (column in colnames(completed_data)){
  # boxplot
    p <- ggplot(plot_df) + 
      geom_boxplot(aes(x="",y=!!sym(column)),outliers = FALSE) +
    geom_jitter(height=0, width=0.3,
                aes(x="",y=!!sym(column),color=eq_included_group)) + 
    scale_color_manual(values=c("brown","darkgreen")) + 
      theme(legend.position = "right")
    
    plot_list_boxplots[[length(plot_list_boxplots)+1]] <- p
    
    pd <- ggplot(plot_df) + 
      geom_density(aes(x=!!sym(column),fill=eq_included_group),alpha=0.5) +
      theme(legend.position = "right") + 
    scale_fill_manual(values=c("brown","darkgreen")) 
    
    plot_list_density[[length(plot_list_density)+1]] <- pd
    
}

pb <- wrap_plots(plot_list_boxplots,guides = "collect")
pd <- wrap_plots(plot_list_density,guides = "collect")
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/nmr_boxplots.pdf",height = 50,width = 50)
pb
dev.off()
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/nmr_density.pdf",height = 50,width = 60)
pd
dev.off()
```

### Log-transformation

Log-transformation and visualizing density plot.

```{r}
data_clean_log <- as.data.frame(
  apply(completed_data,2,function(x) as.numeric(x))) %>%
  `row.names<-`(samplenames)
data_clean_log <- log(data_clean_log)
```

### Tranformed plots 

**Boxplots and density**

```{r}
plot_list_boxplots <- list()
plot_list_density <- list()

plot_df <- merge(data_clean_log %>% rownames_to_column("study_id"),
                 metadata_df %>% dplyr::select("study_id","eq_included_group"),all=TRUE) %>%
  drop_na()

for (column in colnames(data_clean_log)){
  # boxplot
    p <- ggplot(plot_df) + 
      geom_boxplot(aes(x="",y=!!sym(column)),outliers = FALSE) +
    geom_jitter(height=0, width=0.3,
                aes(x="",y=!!sym(column),color=eq_included_group)) + 
    scale_color_manual(values=c("brown","darkgreen")) + 
      theme(legend.position = "right")
    
    plot_list_boxplots[[length(plot_list_boxplots)+1]] <- p
    
    pd <- ggplot(plot_df) + 
      geom_density(aes(x=!!sym(column),fill=eq_included_group),alpha=0.5) +
      theme(legend.position = "right") + 
    scale_fill_manual(values=c("brown","darkgreen")) 
    
    plot_list_density[[length(plot_list_density)+1]] <- pd
    
}

pb <- wrap_plots(plot_list_boxplots,guides = "collect")
pd <- wrap_plots(plot_list_density,guides = "collect")
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/log_nmr_boxplots.pdf",height = 50,width = 50)
pb
dev.off()
```

```{r,eval=FALSE}
pdf("../../Results/preprocessing/log_nmr_density.pdf",height = 50,width = 60)
pd
dev.off()
```

### Autoscaling

```{r}
data_scaled <- data_clean_log
data_scaled <- scale(data_scaled, center = TRUE, scale = TRUE)

```

```{r}
write.table(data_scaled,"../../../Data/preprocessed/metabolomics/nmr_preprocessed.tsv",row.names = TRUE,sep="\t")
```

# Integration

## vegans only

Firstly, we need to import our data and get only vegans.

```{r}
lc_ms_df <- read.table("../../../Data/preprocessed/metabolomics/lc_ms_preprocessed.tsv")
nmr_df <- read.table("../../../Data/preprocessed/metabolomics/nmr_preprocessed.tsv")
metadata_df <- read.table("../../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",header=TRUE)

metadata_vegans <- metadata_df %>%
  dplyr::filter(eq_included_group=="vegan")

lc_ms_vegans <- lc_ms_df %>% 
  dplyr::mutate(study_id=rownames(lc_ms_df)) %>% 
  dplyr::filter(study_id %in% metadata_vegans$study_id) %>%
  dplyr::select(-study_id)

nmr_vegans <- nmr_df %>% 
  dplyr::mutate(study_id=rownames(nmr_df)) %>% 
  dplyr::filter(study_id %in% metadata_vegans$study_id) %>%
  dplyr::select(-study_id)

common_samples <- intersect(rownames(lc_ms_vegans), rownames(nmr_vegans))

lc_ms_vegans <- lc_ms_vegans[common_samples,]
nmr_vegans <- nmr_vegans[common_samples,]

metadata_vegans <- metadata_vegans[metadata_vegans$study_id %in% common_samples,]
```

### MixOmics

**SPCA**

```{r}
integrated_df <- cbind(nmr_vegans,lc_ms_vegans)

spca_result <- mixOmics::spca(integrated_df,ncomp = 20)

spca_data <- spca_result$variates$X

write.table(spca_data,"../../../Data/preprocessed/metabolomics/spca_integrated_vegans.tsv",row.names = TRUE, sep="\t")
```

```{r}
save(spca_result,file="../../Results/preprocessing/spca_result_vegans.RData")

```

## all samples

Firstly, we need to import our data

```{r}
lc_ms_df <- read.table("../../../Data/preprocessed/metabolomics/lc_ms_preprocessed.tsv")

nmr_df <- read.table("../../../Data/preprocessed/metabolomics/nmr_preprocessed.tsv")

metadata_df <- read.table("../../../Data/preprocessed/dietary/veganscreener_preprocessed.tsv",header=TRUE)

common_samples <- intersect(rownames(lc_ms_df), rownames(nmr_df))

lc_ms_df <- lc_ms_df[common_samples,]
nmr_df <- nmr_df[common_samples,]

```

### MixOmics

**SPCA**

```{r}
integrated_df <- cbind(nmr_df,lc_ms_df)

spca_result <- mixOmics::spca(integrated_df,ncomp = 20)

spca_data <- spca_result$variates$X

write.table(spca_data,"../../../Data/preprocessed/metabolomics/spca_integrated_all.tsv",row.names = TRUE, sep="\t")
```

```{r}
save(spca_result,file="../../Results/preprocessing/spca_result_all.RData")

```


# Exploratory analysis

-   Histograms / density plots for each feature
-   Correlation heatmaps (Spearman is robust)
-   PCA or UMAP for metabolomics or tSNE
-   Clustering: hierarchical or k-means, DBSCAN
-   Predictive models
-   univariate analysis
-   multivariate regression analysis
-   network anlaysis - DCPC?
-   pathway analysis - mapping and enrichment analysis

## LC-MS alone

## NMR alone

## Integrated dataset

```{r}
sessionInfo()
```
